### 2018-04-24 doc/genesetdp.tex
get going

\documentclass[11pt]{article}
\begin{document}

This algorithm deals with a genome of size $m$, where each gene $i$'s number of links to a given pathway is given by a index $l_i$. We also define $l_0=0$ and $l_{m+1}=0$. The number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $r=\max_{\forall i}{l_i}$.
We are interested in calculating the score distribution i.e. the number of times a summed number of links would appear for a query containing $q$ genes.
We can do so by calculating the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway.
We first notice that there are $n\choose a$ ways to select $a$ from $n$ elements.

This can be done by solving the recursion

\[
N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b),
\]
where $N_a(s',c')=0$ for all $s'\le 0$ or $c' \le -1$, with the exception for $N_{-1}(0,0)=1$.

The final score distribution is given by $N(s)=N(s,m+1,q+1)$.

\end{document}

### 2018-04-24 src/genesetdp.py

@@ -1,0 +1,44 @@
+#/usr/bin python
+import numpy as np
+import operator as op
+import functools
+
+def ncr(n, r):
+    r = min(r, n-r)
+    numer = functools.reduce(op.mul, range(n, n-r, -1), 1)
+    denom = functools.reduce(op.mul, range(1, r+1), 1)
+    return numer//denom
+
+def find_maxscore(k,q):
+    max_s, ix, rem = 0, len(k)-1, q
+    while rem > k[ix]:
+        max_s += k[ix]*ix
+        rem -= k[ix]
+        ix -= 1
+    max_s += rem*ix
+    return max_s
+
+def genesetdp(k,q):
+    max_s = find_maxscore(k,q)
+    #print(max_s,q)
+    N = np.zeros((max_s+1,q+1),dtype='u4')
+    N[0,0] = 1
+
+    for a in range(len(k)):
+        for s in range(max_s,-1,-1):
+            for c in range(q,-1,-1):
+                for b in range(k[a],0,-1): # Skipping zero (leftover)
+                    if c-b>=0 and s-a*b>=0:
+                        #print("pre\n",N,"a=",a,"s=",s,"c=",c,"b=",b,"k[a]=",k[a],"N[s-a*b,c-b]=",N[s-a*b,c-b],s-a*b,c-b)
+                        N[s,c] += ncr(k[a], b) * N[s-a*b,c-b]
+                        #print("post\n",N,"a=",a,"s=",s,"c=",c,"b=",b,"k[a]=",k[a])
+                    #else:
+                        #print("skip", N,"a=",a,"s=",s,"c=",c,"b=",b,"k[a]=",k[a],s-a*b,c-b)
+    return N[:,q]
+
+if __name__ == "__main__":
+    # Query size q, number of links k[a]
+    q = 8
+    k = np.array([8,5,3,3,4],dtype='u4')
+    #k = np.array([3,3,3],dtype='u4')
+    print(genesetdp(k,q))

### 2018-04-24 doc/genesetdp.tex
Go for it.

@@
 
-This algorithm deals with a genome of size $m$, where each gene $i$'s number of links to a given pathway is given by a index $l_i$.
-We also define $l_0=0$ and $l_{m+1}=0$.
-The number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $r=\max_{\forall i}{l_i}$.
+In network-based gene set analysis one score a query set of genes $ \{g_1 \ldots g_Q\} $ from a genome with genes $\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$.
+In network-based gene set analysis the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.
+One frequently employed metric to score such pathways are to sum up all conections, i.e.
+the pathway score is $f=\sum_{i=1}^Q\sum_{j=1}^P x_{ij}$.
+We can reformulate such scores as a set of pathway specific gene scores by defining the number-of-links, $l_i\sum_{j=1}^P x_{ij}$, which gives us a score $f=\sum_{i=1}^Q l_i$.
+Such number-of-links, $l_i$, can be precomputed ahead of time for all genes $i, \ldots, G$.
+Furtheremore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{\forall i}{l_i}$.
 We are interested in calculating the score distribution i.e. the number of times a summed number of links would appear for a query containing $q$ genes.

### 2018-04-24 doc/genesetdp.tex
Changing computers

@@
 Such number-of-links, $l_i$, can be precomputed ahead of time for all genes $i, \ldots, G$.
-Furtheremore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{\forall i}{l_i}$.
-We are interested in calculating the score distribution i.e. the number of times a summed number of links would appear for a query containing $q$ genes.
+Furthermore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{\forall i}{l_i}$.
+
+To assess confidence metrics for $f$, we are interested in calculating the score distribution i.e. the number of times a summed number of links would appear for a query containing $Q$ genes.
 We can do so by calculating the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway.
+
 We first notice that there are $n\choose a$ ways to select $a$ from $n$ elements.
 
-This can be done by solving the recursion
+This helps us formulate the recursion,
 
@@
 
-The final score distribution is given by $N(s)=N(s,m+1,q+1)$.
+The final score distribution is given by $N(s)=N_R(s,Q)$.
+
+There is a memory efficient implementation. We first note that
+
+\[
+N_a(s,c)=\sum_{b=1}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b) + N_{a-1}(s,c).
+\]
+
+This enables us to calculate $N$, by nested iterations over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$ over one single $(RQ+1) \times (Q+1)$ matrix
 

### 2018-04-24 doc/genesetdp.tex
done?

@@
 
-This enables us to calculate $N$, by nested iterations over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$ over one single $(RQ+1) \times (Q+1)$ matrix
+This enables us to calculate $N$, by nested iterations over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$ over one single $(RQ+1) \times (Q+1)$ matrix.
+
+\subsection*{$p$ value calculation}
+
+Once we have our full score distribution, we can evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are random''.
+This can be expressed as an unbiased empirical $p$~value for obtaining a score $s$,
+
+\[
+p(s)=\frac{N(s')/2 +\sum_{s'=s+1}^{RQ} N(s')}{\sum_{s'=0}^{RQ} N(s')}
+\]
 

### 2018-04-26 doc/genesetdp.tex
More f.
Used new style, added an introduction and reformated algorithm

@@
-\documentclass[11pt]{article}
+\documentclass[a4paper,USenglish]{lipics-v2016}
+\usepackage{microtype}
+\bibliographystyle{plainurl}
+\title{A dynamic programing algorithm to assess the reliability of inferences from network enrichment analysis}
+\titlerunning{Dynamic Programing for Network Enrichment inferences} %optional, in case that the title is too long; the running title should fit into the top page column
+
+%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
+\author[1]{Gustavo S. Jeuken}
+\author[2]{Lukas K\"{a}ll}
+\affil[1]{Science for Life Laboratory, School of
+Engineering Sciences in Chemistry, Biotechnology and Health,
+Royal Institute of Technology -- KTH, Box 1031, 17121 Solna, Sweden\\ \texttt{gustavo.jeuken@scilifelab.se}}
+\affil[2]{Science for Life Laboratory, School of
+Engineering Sciences in Chemistry, Biotechnology and Health,
+Royal Institute of Technology -- KTH, Box 1031, 17121 Solna, Sweden\\ \texttt{lukas.kall@scilifelab.se}}
+\authorrunning{G. S. Jeuken and L. K\"{a}ll}
 \begin{document}
 
+\maketitle
+
+\section*{Introduction}
+
+Gene enrichement analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}.
+The technique is used to esimate which functional modules such as complexes or pathways that are alterated between different biological conditions, such as disease state or treatment group.
+Particularly in expression analysis, GEA is used to assess alternation in activity of functional models by examining significantly different concentrations of analytes between experimental conditions.
+
+Most GEA methods are investigating the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test.
+However, variants such as Gene Set Enrichment Analysis (GSEA)\cite{subramanian2005gene} also includes information on expression levels of the analytes of the query set.
+
+A limiting factor of GEA are the pathway definition databases. While \url{http://pathguide.org} list 708 databases of
+pathway definitions\cite{bader2006pathguide},  some chritique has been voiced concerning the completenes and rigour of current pathway databases.
+Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string}.
+Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{shojaie2010network, alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}.
+This is known as Network Enrichment Analysis (NEA).
+
+The quality of the inferences from NEA has previously been assessed by Monte carlo-based methods that investigates the scores distribution when evaluating the query set using random networks.
+
+Here, we present a dynamic programing algorithm to calculate the full score distribution of any query of a given number of analytes.
+The algorithm enables us to calculate well defined and exact statistics of inferences from NEA.
+Using simulations we demonstarated that the algorithm produced unbiased statistics.
+
+\section*{Algorithm}
+
 In network-based gene set analysis one score a query set of genes $ \{g_1 \ldots g_Q\} $ from a genome with genes $\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$.
-In network-based gene set analysis the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.
-One frequently employed metric to score such pathways are to sum up all conections, i.e.
-the pathway score is $f=\sum_{i=1}^Q\sum_{j=1}^P x_{ij}$.
-We can reformulate such scores as a set of pathway specific gene scores by defining the number-of-links, $l_i\sum_{j=1}^P x_{ij}$, which gives us a score $f=\sum_{i=1}^Q l_i$.
+In NEA the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.
+The pathways are scored by summing up all conections, i.e.
+the pathway score is
+
+\begin{equation}
+s=\sum_{i=1}^Q\sum_{j=1}^P x_{ij}.
+\label{eq:sum_ij}
+\end{equation}
+
+Here, we wanted to determine a full score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$.
+This allows us to evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are randomly selected''.
+This can be expressed as an unbiased empirical $p$~value for obtaining a score $s$,
+
+\begin{equation}
+p(s)=\frac{N(s)/2 +\sum_{s'=s+1}^{S} N(s')}{\sum_{s'=0}^{S} N(s')},
+\end{equation}
+where $S$ is the maximal score.
+
+\subsection*{Computation of the full score distribution}
+
+We can reformulate Equation \ref{eq:sum_ij} by defining the number-of-links, $l_i\sum_{j=1}^P x_{ij}$, which gives us a score $s=\sum_{i=1}^Q l_i$.
 Such number-of-links, $l_i$, can be precomputed ahead of time for all genes $i, \ldots, G$.
-Furthermore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{\forall i}{l_i}$.
+Furthermore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{i}{l_i}$.
 
-To assess confidence metrics for $f$, we are interested in calculating the score distribution i.e. the number of times a summed number of links would appear for a query containing $Q$ genes.
-We can do so by calculating the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway.
-
-We first notice that there are $n\choose a$ ways to select $a$ from $n$ elements.
-
-This helps us formulate the recursion,
-
-\[
+In our strive to calculate the score distribution, $N(s)$, we first want to investigate the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway.
+This helps us formulate the recursion, using the fact that there are $k_a \choose b$ ways to select $b$ from $k_a$ elements, as
+\begin{equation}
 N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b),
-\]
+\end{equation}
 where $N_a(s',c')=0$ for all $s'\le 0$ or $c' \le -1$, with the exception for $N_{-1}(0,0)=1$.
@@
 The final score distribution is given by $N(s)=N_R(s,Q)$.
+
+\subsection*{Implementation}
 
@@
 
-\subsection*{$p$ value calculation}
+\section*{Methods}
 
-Once we have our full score distribution, we can evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are random''.
-This can be expressed as an unbiased empirical $p$~value for obtaining a score $s$,
+\section*{Results}
 
-\[
-p(s)=\frac{N(s')/2 +\sum_{s'=s+1}^{RQ} N(s')}{\sum_{s'=0}^{RQ} N(s')}
-\]
+\section*{Discussion}
+
+\bibliography{refs}
 

### 2018-04-28 doc/genesetdp.tex
Added result and disussion sections

@@
-\documentclass[a4paper,USenglish]{lipics-v2016}
-\usepackage{microtype}
+\documentclass[a4paper,american]{lipics-v2016}
+%\usepackage[USenglish]{babel}
+%\usepackage{microtype}
 \bibliographystyle{plainurl}
@@
 
-Gene enrichement analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}.
-The technique is used to esimate which functional modules such as complexes or pathways that are alterated between different biological conditions, such as disease state or treatment group.
+Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}.
+The technique is used to estimate which functional modules such as complexes or pathways that diverge between different biological conditions, such as disease state or treatment group.
 Particularly in expression analysis, GEA is used to assess alternation in activity of functional models by examining significantly different concentrations of analytes between experimental conditions.
@@
 A limiting factor of GEA are the pathway definition databases. While \url{http://pathguide.org} list 708 databases of
-pathway definitions\cite{bader2006pathguide},  some chritique has been voiced concerning the completenes and rigour of current pathway databases.
+pathway definitions\cite{bader2006pathguide},  some chritique has been voiced concerning the completeness and rigor of current pathway databases.
 Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string}.
@@
 
-The quality of the inferences from NEA has previously been assessed by Monte carlo-based methods that investigates the scores distribution when evaluating the query set using random networks.
+The quality of the inferences from NEA has previously been assessed by Monte Carlo-simulations   that investigates the scores distribution when evaluating the query set using random networks.
 
-Here, we present a dynamic programing algorithm to calculate the full score distribution of any query of a given number of analytes.
+Here, we present a dynamic programing algorithm, that we dubbed GeneSetDP, to calculate the full score distribution of any query of a given number of analytes.
 The algorithm enables us to calculate well defined and exact statistics of inferences from NEA.
-Using simulations we demonstarated that the algorithm produced unbiased statistics.
+Using simulations we demonstrated that the algorithm produced unbiased statistics.
 
@@
 Here, we wanted to determine a full score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$.
-This allows us to evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are randomly selected''.
+This allows us to evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
 This can be expressed as an unbiased empirical $p$~value for obtaining a score $s$,
@@
 p(s)=\frac{N(s)/2 +\sum_{s'=s+1}^{S} N(s')}{\sum_{s'=0}^{S} N(s')},
+\label{eq:pval}
 \end{equation}
@@
 
-We can reformulate Equation \ref{eq:sum_ij} by defining the number-of-links, $l_i\sum_{j=1}^P x_{ij}$, which gives us a score $s=\sum_{i=1}^Q l_i$.
+We can reformulate Equation \ref{eq:sum_ij} by defining the number-of-links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
+\begin{equation}
+s=\sum_{i=1}^Q l_i.
+\label{eq:sum_i}
+\end{equation}
+
 Such number-of-links, $l_i$, can be precomputed ahead of time for all genes $i, \ldots, G$.
@@
 
+\begin{lstlisting}[caption={
+The central part of the dynamic programing algorithm for finding $N(s)$.
+The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input.
+The function depends on two additional functions {\tt find\_maxscore(k,Q) }, which calculates the maximal score a query of size $Q$ can obtain, and {\tt ncr(a,b)}, which calculates $ a \choose b $.
+}
+, label=lst:gensetdp,captionpos=t,float,abovecaptionskip=-\medskipamount]
+def genesetdp(k,Q):
+    max_s = find_maxscore(k,Q)
+    N = np.zeros((max_s+1,Q+1),dtype='u4')
+    N[0,0] = 1
+
+    for a in range(len(k)):
+        for s in range(max_s,-1,-1):
+            for c in range(q,-1,-1):
+                for b in range(k[a],0,-1):
+                    if c-b>=0 and s-a*b>=0:
+                        N[s,c] += ncr(k[a], b) * N[s-a*b,c-b]
+    return N[:,q]
+\end{lstlisting}
 There is a memory efficient implementation. We first note that
@@
 \[
-N_a(s,c)=\sum_{b=1}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b) + N_{a-1}(s,c).
+N_a(s,c)=\sum_{b=1}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b) + N_{a-1}(s,c)=D_a+N_{a-1}(s,c).
 \]
 
-This enables us to calculate $N$, by nested iterations over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$ over one single $(RQ+1) \times (Q+1)$ matrix.
+This enables us to calculate $N$ in place. That is we do not have to copy the dynamic programing matrix in each iteration over $a$,
+instead we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$.
+For details see Listing \ref{lst:gensetdp}.
 
@@
 
+We dowloaded network definitions and the BinoX software version FIXME for comparisons from XXX \url{bitbucket/sonnhammerlab}.
+FIXME: add a more detailed description of the test set (i.e.
+organism and number of genes, pathways and network connections).
+
 \section*{Results}
+
+We implemented a Python program that read network and pathway definition files, and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
+We downloaded pathway and network definitions from the BinoX's website.
+
+To illustrate the $p$~value calculation procedure we plotted the score distribution N(s) for a query size of 10 (FIXME) genes in Figure \ref{fig:score_dist}.
+
+\begin{figure}[htb]
+  {\em FIXME: Add this figure.}
+%\includegraphics{img/FIXME.png}
+  \caption{{\bf Score distribution of a query size of 10 genes.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
+how many ways one can reach a certain score when adding up the scores for $Q$ genes.}
+  \label{fig:score_dist}
+\end{figure}
+
+\subsection*{Test of Calibration}
+
+In order to test the statistical calibration of our method, we calculated $p$~values for FIXME random picks of gene sets from the investigated genome.
+As our selection is random, and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed.
+To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration}.
+As a comparison we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.
+
+We note that the calibration of
+
+\begin{figure}[htb]
+  {\em FIXME: Make a loglog plot of the relative rank and pvalues of GeneSetDP and BinoX. Add filled lines for y=x, and dotted lines for y=2x and y=0.5x.}
+%\includegraphics{img/FIXME.png}
+  \caption{{\bf Calibration of GeneSetDP.} We selected FIXME random sets from the investigated genome definition, and calculated their $p$ values with GeneSetDP.
+To test the $p$~values uniformity we plotted them against their quantile.
+We also added the calibration curve of BinoX, evaluated on the same random sets.}
+  \label{fig:calibration}
+\end{figure}
 
@@
 
-\bibliography{refs}
+Here we have implemented a method, GeneSetDP, to calculate unbiased $p$~values for inferences from NEA.
+We demonstrated that the method gave more accurate statistics than the popular BinoX method using random subsets of a genome.
+
+Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
+Previous implementation of NEA in the literature otherwise use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.
+In practice such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics.
+
+\bibliography{./refs}
 

### 2018-04-29 src/experiment.py
Files for generating k[a] and computing p values

@@ -1,0 +1,23 @@
+import pandas as pd
+import numpy as np
+
+import generate_k as genk
+import genesetdp
+
+## example path
+path = "~/git/binox/example/"
+
+## read original patway information file
+pathways_file = 'Pathways.tsv'
+pathways = pd.read_csv(path + pathways_file, sep='\t', header=None)
+pathways.columns = ['gene','pathway']
+
+pathway_name = 'CHEMICAL_CARCINOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
+
+pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.values
+
+k = genk.generate_k(pathway_genes)
+
+N = genesetdp.genesetdp(k,4)
+
+p = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)

### 2018-04-29 src/generate_k.py

@@ -1,0 +1,29 @@
+import pandas as pd
+import numpy as np
+
+## example path
+path = "~/git/binox/example/"
+
+
+## read original network file provided by BinoX example
+network_file = 'Network.tsv'
+network_proto = pd.read_csv(path + network_file, sep='\t')
+
+## manipulate network
+network_treshold = 0.99
+network_cut = network_proto['#0:PFC'] >= network_treshold
+network_one = network_proto.loc[network_cut]
+network = network_one.iloc[:,[0,1]].reset_index(drop=True)
+network.columns = ['gene1','gene2']
+network = network.set_index('gene1')
+
+def generate_k(pathway_genes):
+    links = network.loc[pathway_genes].dropna()
+
+    k_proto = links.gene2.value_counts().value_counts()
+
+    k = k_proto.loc[list(range(max(k_proto.index)+1))]
+    k[k.isna()] = 0
+    k = np.array(k[1:], dtype='u4')
+    
+    return k

### 2018-04-30 src/genesetdp.py
removes N dtype to allow for larger numbers

@@ -21,7 +21,7 @@ def find_maxscore(k,q):
 def genesetdp(k,q):
     max_s = find_maxscore(k,q)
     #print(max_s,q)
-    N = np.zeros((max_s+1,q+1),dtype='u4')
+    N = np.zeros((max_s+1,q+1))
     N[0,0] = 1
 
     for a in range(len(k)):

### 2018-04-30 doc/genesetdp.tex
x

@@
 Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}.
-The technique is used to estimate which functional modules such as complexes or pathways that diverge between different biological conditions, such as disease state or treatment group.
-Particularly in expression analysis, GEA is used to assess alternation in activity of functional models by examining significantly different concentrations of analytes between experimental conditions.
+The technique predict which functional modules, such as complexes or pathways, that are overrepresentated among a set of identified analytes.
+One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.
 
@@
 
-Here, we present a dynamic programing algorithm, that we dubbed GeneSetDP, to calculate the full score distribution of any query of a given number of analytes.
+Here, we present a dynamic programing algorithm, that we dubbed GeneSetDP, that calculate the full score distribution of any query of a given size.
 The algorithm enables us to calculate well defined and exact statistics of inferences from NEA.

### 2018-04-30 src/example-random/generate_random.py
added random example for BinoX

@@ -1,0 +1,29 @@
+import pandas as pd
+import numpy as np
+
+original = pd.read_csv('Signatures.tsv', sep='\t')
+
+genes = np.unique(original.index)
+
+random_signatures = pd.DataFrame()
+
+for i in np.arange(1,10001,1):
+    sample = np.random.choice(len(genes),10, replace=False)
+    sample_genes = genes[sample]
+    new_group = pd.DataFrame(np.transpose([sample_genes, np.repeat('group' + str(i),10)])) 
+    random_signatures = random_signatures.append(new_group)
+
+random_signatures.columns = ['#Node', 'GroupName']
+random_signatures.set_index('#Node').to_csv('Random_Signatures.tsv', sep = '\t')
+
+
+#####
+
+pathways = pd.read_csv('Pathways.tsv', sep='\t', header=None)
+
+pathway_name = 'GLYCOLYSIS_/_GLUCONEOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
+
+one_pathway = pathways.where(pathways.iloc[:,1] == pathway_name).dropna()
+
+one_pathway.to_csv('Pathway_One.tsv', sep='\t', header=None, index=False)
+

### 2018-04-30 doc/genesetdp.tex
x

@@
 This allows us to evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
-This can be expressed as an unbiased empirical $p$~value for obtaining a score $s$,
+This can be expressed as a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score $s$,
 
@@
 \end{equation}
-where $S$ is the maximal score.
+where $S$ is the maximal score $s$ that $Q$ genes can obtain from Equation \ref{eq:sum_ij}.
 

### 2018-04-30 src/experiment.py
first results from random genesets on genesetDP

@@ -20,7 +20,9 @@ k = genk.generate_k(pathway_genes)
 
 N = genesetdp.genesetdp(k,10)
 
-p = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
+p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
+
+####
 
 import matplotlib.pyplot as plt
 plt.step(np.arange(len(N))+1, N, where='mid')
@@ -29,3 +31,22 @@ plt.ylabel('N(s)')
 
 plt.savefig('F1_score_distribuition.png')
 
+
+####
+
+groups = pd.read_csv('../src/example-random/Random_Signatures.tsv', sep='\t')
+
+network = genk.network
+
+linked_genes = network.loc[pathway_genes].dropna().gene2.values.tolist()
+
+results = pd.DataFrame(columns = ['n','p'])
+## start for
+for i in range(1,10001):
+    groupname = 'group' + str(i)
+    genes = groups.loc[groups.GroupName==groupname,'#Node']
+    n = sum(linked_genes.count(x) for x in genes)
+    p = p_vals[n]
+    results.loc[groupname] = [n,p]
+
+results.to_csv('genesetDP_results.tsv', sep = '\t')

### 2018-05-01 src/generate_k.py
fix k[0]

@@ -17,6 +17,8 @@ network = network_one.iloc[:,[0,1]].reset_index(drop=True)
 network.columns = ['gene1','gene2']
 network = network.set_index('gene1')
 
+n_genes = len(np.unique(network.index))
+
 def generate_k(pathway_genes):
     links = network.loc[pathway_genes].dropna()
 
@@ -24,6 +26,9 @@ def generate_k(pathway_genes):
 
     k = k_proto.loc[list(range(max(k_proto.index)+1))]
     k[k.isna()] = 0
-    k = np.array(k[1:], dtype='u4')
     
+    zero_link_genes = n_genes - sum(k)
+    k = np.array(k[1:], dtype='u4')
+    k = np.insert(k,0,zero_link_genes)
+
     return k

### 2018-05-01 src/example-random/generate_random.py
sx

@@ -1,7 +1,9 @@
 import pandas as pd
 import numpy as np
 
-original = pd.read_csv('Signatures.tsv', sep='\t')
+path = "~/git/binox/example/"
+
+original = pd.read_csv(path + 'Signatures.tsv', sep='\t')
 
 genes = np.unique(original.index)
 
@@ -10,7 +12,7 @@ random_signatures = pd.DataFrame()
 for i in np.arange(1,10001,1):
     sample = np.random.choice(len(genes),10, replace=False)
     sample_genes = genes[sample]
-    new_group = pd.DataFrame(np.transpose([sample_genes, np.repeat('group' + str(i),10)])) 
+    new_group = pd.DataFrame(np.transpose([sample_genes, np.repeat('group' + str(i),10)]))
     random_signatures = random_signatures.append(new_group)
 
 random_signatures.columns = ['#Node', 'GroupName']
@@ -19,11 +21,10 @@ random_signatures.set_index('#Node').to_csv('Random_Signatures.tsv', sep = '\t')
 
 #####
 
-pathways = pd.read_csv('Pathways.tsv', sep='\t', header=None)
+pathways = pd.read_csv(path + 'Pathways.tsv', sep='\t', header=None)
 
 pathway_name = 'GLYCOLYSIS_/_GLUCONEOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
 
 one_pathway = pathways.where(pathways.iloc[:,1] == pathway_name).dropna()
 
 one_pathway.to_csv('Pathway_One.tsv', sep='\t', header=None, index=False)
-

### 2018-05-02 exp/example-random/combine_plot.py
reorganization

@@ -1,0 +1,41 @@
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+
+genesetdp = pd.read_csv('genesetDP_results.tsv', sep='\t', index_col=0)
+
+binox = pd.read_csv('Pathway_One_vs_Random_Signatures.tsv', sep='\t')
+binox = binox.set_index('#3:NameGroupB')
+binox.index = [x.lower() for x in binox.index]
+
+
+binox = binox['#4:p.value']
+genesetdp = pd.read_csv('genesetDP_results.tsv', sep='\t', index_col=0)
+
+binox = binox.loc[genesetdp.index]
+binox.loc[np.isnan(binox)] = 1
+
+binox_sorted = binox.sort_values()
+binox_sorted.index = (np.arange(len(binox_sorted.index))+1)/len(binox_sorted.index)
+
+genesetdp_sorted = genesetdp['p'].dropna().sort_values()
+genesetdp_sorted.index = (np.arange(len(genesetdp_sorted.index))+1)/len(genesetdp_sorted)
+
+
+x_y = pd.DataFrame(np.arange(0,1,0.001), index = np.arange(0,1,0.001))
+x_2y = pd.DataFrame(np.arange(0,1,0.001)*2, index = np.arange(0,1,0.001))
+x_halfy = pd.DataFrame(np.arange(0,0.5,0.001), index = np.arange(0,0.5,0.001)*2)
+
+
+plt.loglog(binox_sorted, '.')
+plt.loglog(genesetdp_sorted, '.')
+plt.loglog(x_y, color = 'r', alpha = 0.5)
+plt.loglog(x_2y, '--',color = 'b', alpha = 0.5)
+plt.loglog(x_halfy, '--',color = 'b', alpha = 0.5)
+
+plt.legend(['BinoX', 'genesetDP'])
+
+plt.show()
+
+
+

### 2018-05-02 src/generate_k.py

@@ -10,7 +10,7 @@ network_file = 'Network.tsv'
 network_proto = pd.read_csv(path + network_file, sep='\t')
 
 ## manipulate network
-network_treshold = 0.99
+network_treshold = 0.7
 network_cut = network_proto['#0:PFC'] >= network_treshold
 network_one = network_proto.loc[network_cut]
 network = network_one.iloc[:,[0,1]].reset_index(drop=True)

### 2018-05-02 exp/example-random/combine_plot.py
plot

@@ -33,9 +33,13 @@ plt.loglog(x_y, color = 'r', alpha = 0.5)
 plt.loglog(x_2y, '--',color = 'b', alpha = 0.5)
 plt.loglog(x_halfy, '--',color = 'b', alpha = 0.5)
 
-plt.legend(['BinoX', 'genesetDP'])
+plt.legend(['BinoX', 'genesetDP', 'x=y', 'x=[0.5,2]y'])
+plt.ylabel('p-values')
+plt.xlabel('relative rank')
 
-plt.show()
+plt.savefig('F2_calibration.png')
+
+
 
 
 

### 2018-05-02 exp/experiment.py
plot

@@ -1,6 +1,8 @@
 import pandas as pd
 import numpy as np
 
+import sys
+sys.path.insert(0, '../src')
 import generate_k as genk
 import genesetdp
 
@@ -25,7 +27,7 @@ p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
 ####
 
 import matplotlib.pyplot as plt
-plt.step(np.arange(len(N))+1, N, where='mid')
+plt.step(np.arange(len(N)), N, where='mid')
 plt.xlabel('score')
 plt.ylabel('N(s)')
 

### 2018-05-02 doc/genesetdp.tex
added plot
fix plot

@@
 
-We dowloaded network definitions and the BinoX software version FIXME for comparisons from XXX \url{bitbucket/sonnhammerlab}.
+We dowloaded network definitions and the BinoX software (on 2018/04/28) for comparisons from XXX \url{bitbucket/sonnhammerlab}.
 FIXME: add a more detailed description of the test set (i.e.
@@
 
-To illustrate the $p$~value calculation procedure we plotted the score distribution N(s) for a query size of 10 (FIXME) genes in Figure \ref{fig:score_dist}.
+To illustrate the $p$~value calculation procedure we plotted the score distribution N(s) for a query size of 10 genes in Figure \ref{fig:score_dist}.
 
 \begin{figure}[htb]
-  {\em FIXME: Add this figure.}
+		\begin{center}
+				\includegraphics[scale=0.8]{figures/F1_score_distribuition.png}
+		\end{center}
 %\includegraphics{img/FIXME.png}
@@
 
-In order to test the statistical calibration of our method, we calculated $p$~values for FIXME random picks of gene sets from the investigated genome.
+In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets from the investigated genome.
 As our selection is random, and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed.
@@
 \begin{figure}[htb]
-  {\em FIXME: Make a loglog plot of the relative rank and pvalues of GeneSetDP and BinoX. Add filled lines for y=x, and dotted lines for y=2x and y=0.5x.}
+		\begin{center}
+				\includegraphics[scale=0.8]{figures/F2_calibration.png}
+		\end{center}
+
 %\includegraphics{img/FIXME.png}
-  \caption{{\bf Calibration of GeneSetDP.} We selected FIXME random sets from the investigated genome definition, and calculated their $p$ values with GeneSetDP.
+  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition, and calculated their $p$ values with GeneSetDP.
 To test the $p$~values uniformity we plotted them against their quantile.

### 2018-05-02 exp/example-random/combine_plot.py

@@ -33,7 +33,7 @@ plt.loglog(x_y, color = 'r', alpha = 0.5)
 plt.loglog(x_2y, '--',color = 'b', alpha = 0.5)
 plt.loglog(x_halfy, '--',color = 'b', alpha = 0.5)
 
-plt.legend(['BinoX', 'genesetDP', 'x=y', 'x=[0.5,2]y'])
+plt.legend(['BinoX', 'GeneSetDP', 'x=y', 'x=[0.5,2]y'])
 plt.ylabel('p-values')
 plt.xlabel('relative rank')
 

### 2018-05-02 exp/example-random/combine_plot.py
plot

@@ -29,17 +29,13 @@ x_halfy = pd.DataFrame(np.arange(0,0.5,0.001), index = np.arange(0,0.5,0.001)*2)
 
 plt.loglog(binox_sorted, '.')
 plt.loglog(genesetdp_sorted, '.')
-plt.loglog(x_y, color = 'r', alpha = 0.5)
-plt.loglog(x_2y, '--',color = 'b', alpha = 0.5)
-plt.loglog(x_halfy, '--',color = 'b', alpha = 0.5)
+plt.loglog(x_y, '--', color = 'r', alpha = 0.5)
+plt.loglog(x_2y, ':',color = 'b', alpha = 0.5)
+plt.loglog(x_halfy, ':',color = 'b', alpha = 0.5)
 
 plt.legend(['BinoX', 'GeneSetDP', 'x=y', 'x=[0.5,2]y'])
 plt.ylabel('p-values')
 plt.xlabel('relative rank')
 
 plt.savefig('F2_calibration.png')
-
-
-
-
 

### 2018-05-02 exp/example-random/combine_plot.py
Edited axis name and used seaborn plot style.

@@ -1,6 +1,8 @@
 import pandas as pd
 import numpy as np
 import matplotlib.pyplot as plt
+import seaborn as sns
+
 
 genesetdp = pd.read_csv('genesetDP_results.tsv', sep='\t', index_col=0)
 
@@ -26,6 +28,9 @@ x_y = pd.DataFrame(np.arange(0,1,0.001), index = np.arange(0,1,0.001))
 x_2y = pd.DataFrame(np.arange(0,1,0.001)*2, index = np.arange(0,1,0.001))
 x_halfy = pd.DataFrame(np.arange(0,0.5,0.001), index = np.arange(0,0.5,0.001)*2)
 
+sns.set_style("ticks")
+sns.despine()
+
 
 plt.loglog(binox_sorted, '.')
 plt.loglog(genesetdp_sorted, '.')
@@ -33,9 +38,8 @@ plt.loglog(x_y, '--', color = 'r', alpha = 0.5)
 plt.loglog(x_2y, ':',color = 'b', alpha = 0.5)
 plt.loglog(x_halfy, ':',color = 'b', alpha = 0.5)
 
-plt.legend(['BinoX', 'GeneSetDP', 'x=y', 'x=[0.5,2]y'])
-plt.ylabel('p-values')
-plt.xlabel('relative rank')
+plt.legend(['BinoX', 'GeneSetDP', 'y=x', 'y=0.5x,y=2x'])
+plt.ylabel('$p$ value')
+plt.xlabel('Normalized rank')
 
 plt.savefig('F2_calibration.png')
-

### 2018-05-02 exp/experiment.py

@@ -20,16 +20,20 @@ pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.v
 
 k = genk.generate_k(pathway_genes)
 
-N = genesetdp.genesetdp(k,10)
+N = genesetdp.genesetdp(k,25)
 
 p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
 
 ####
 
 import matplotlib.pyplot as plt
+import seaborn as sns
+
+sns.set_style("ticks")
+sns.despine()
+
 plt.step(np.arange(len(N)), N, where='mid')
-plt.xlabel('score')
-plt.ylabel('N(s)')
+plt.xlabel('Score, $s$')
 
 plt.savefig('F1_score_distribuition.png')
 

### 2018-05-02 doc/genesetdp.tex
parameters
text

@@
 
-We dowloaded network definitions and the BinoX software (on 2018/04/28) for comparisons from XXX \url{bitbucket/sonnhammerlab}.
-FIXME: add a more detailed description of the test set (i.e.
-organism and number of genes, pathways and network connections).
-
+We dowloaded network definitions and the BinoX software (on 2018-04-28) for comparisons from XXX \url{bitbucket/sonnhammerlab}.
+The acompaning example files include a pathway definition file that groups 6819 genes into 289 human pathwats, and a network definition file that, after tresholding, yeilds 1244992 links bewtween those genes.
+% FIXME: add a more detailed description of the test set (i.e. organism and number of genes, pathways and network connections).
+Troughout this study, BinoX was run with the default parameters, and we chose Glycolisi/Glucogenesis as the reference pathway, as it ocontains a number of genes that is median to the set.
 \section*{Results}
@@
 We implemented a Python program that read network and pathway definition files, and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
-We downloaded pathway and network definitions from the BinoX's website.
+We downloaded pathway and network definitions from the BinoX's website and used the same network treshold as BinoX default ($0.7$).
 

### 2018-05-02 doc/genesetdp.tex
Small edits

--- 2018-05-02 doc/genesetdp.tex
@@
 Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}.
-The technique predict which functional modules, such as complexes or pathways, that are overrepresentated among a set of identified analytes.
+The technique estimates which functional modules, such as complexes or pathways, that are overrepresentated among a set of identified analytes.
 One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.
@@
 
-Here, we wanted to determine a full score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$.
+Here, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$.
 This allows us to evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
@@
 
-\subsection*{Computation of the full score distribution}
+\subsection*{Computation of the score distribution}
 
@@
 \end{equation}
-where $N_a(s',c')=0$ for all $s'\le 0$ or $c' \le -1$, with the exception for $N_{-1}(0,0)=1$.
+where $N_a(s,c)=0$ for all $s<0$, $c<0$ or $a<0$, with the exception for $N_{-1}(0,0)=1$.
 
@@
 The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input.
-The function depends on two additional functions {\tt find\_maxscore(k,Q) }, which calculates the maximal score a query of size $Q$ can obtain, and {\tt ncr(a,b)}, which calculates $ a \choose b $.
+The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt ncr(a,b)}, which calculates $ a \choose b $.
 }
@@
     max_s = find_maxscore(k,Q)
-    N = np.zeros((max_s+1,Q+1),dtype='u4')
+    N = np.zeros((max_s+1,Q+1),dtype='float_')
     N[0,0] = 1
@@
 organism and number of genes, pathways and network connections).
-
+Troughout this study, BinoX was run with the default parameters.
 \section*{Results}

### 2018-05-02 src/genesetdp.py

@@ -27,13 +27,9 @@ def genesetdp(k,q):
     for a in range(len(k)):
         for s in range(max_s,-1,-1):
             for c in range(q,-1,-1):
-                for b in range(k[a],0,-1): # Skipping zero (leftover)
+                for b in range(k[a],0,-1):
                     if c-b>=0 and s-a*b>=0:
-                        #print("pre\n",N,"a=",a,"s=",s,"c=",c,"b=",b,"k[a]=",k[a],"N[s-a*b,c-b]=",N[s-a*b,c-b],s-a*b,c-b)
                         N[s,c] += ncr(k[a], b) * N[s-a*b,c-b]
-                        #print("post\n",N,"a=",a,"s=",s,"c=",c,"b=",b,"k[a]=",k[a])
-                    #else:
-                        #print("skip", N,"a=",a,"s=",s,"c=",c,"b=",b,"k[a]=",k[a],s-a*b,c-b)
     return N[:,q]
 
 if __name__ == "__main__":

### 2018-05-02 doc/genesetdp.tex

@@
 
-We dowloaded network definitions and the BinoX software (on 2018/04/28) for comparisons from XXX \url{bitbucket/sonnhammerlab}.
-FIXME: add a more detailed description of the test set (i.e.
-organism and number of genes, pathways and network connections).
-Troughout this study, BinoX was run with the default parameters.
+We dowloaded network definitions and the BinoX software (on 2018-04-28) for comparisons from XXX \url{bitbucket/sonnhammerlab}.
+The acompaning example files include a pathway definition file that groups 6819 genes into 289 human pathwats, and a network definition file that, after tresholding, yeilds 1244992 links bewtween those genes.
+% FIXME: add a more detailed description of the test set (i.e. organism and number of genes, pathways and network connections).
+Troughout this study, BinoX was run with the default parameters, and we chose Glycolisi/Glucogenesis as the reference pathway, as it ocontains a number of genes that is median to the set.
 \section*{Results}

### 2018-05-02 doc/genesetdp.tex
rearangements of text

--- 2018-05-02 doc/genesetdp.tex
@@
         for s in range(max_s,-1,-1):
-            for c in range(q,-1,-1):
-                for b in range(k[a],0,-1):
+            for c in range(Q,-1,-1):
+                for b in range(k[a],0,-1): # Stop at b=1
                     if c-b>=0 and s-a*b>=0:
@@
 
-This enables us to calculate $N$ in place. That is we do not have to copy the dynamic programing matrix in each iteration over $a$,
-instead we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$.
+This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elemets from $N_{a-1}$ still needed to update subsequent elements.
+That is, we do not have to copy the dynamic programing matrix in each iteration over $a$, instead we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$.
 For details see Listing \ref{lst:gensetdp}.
@@
 
-We dowloaded network definitions and the BinoX software (on 2018/04/28) for comparisons from XXX \url{bitbucket/sonnhammerlab}.
-FIXME: add a more detailed description of the test set (i.e.
-organism and number of genes, pathways and network connections).
-Troughout this study, BinoX was run with the default parameters.
+We dowloaded network definitions and the BinoX software (on 2018-04-28) for comparisons from \url{https://bitbucket.org/sonnhammergroup/binox}. BinoX was run with the default parameters.
+% FIXME: add a more detailed description of the test set (i.e. organism and number of genes, pathways and network connections).
+
+We also downloaded the NEA example given at the BinoX web site.
+The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways, and a network definition file that, after tresholding with a score of $0.7$, yeilds $1244992$ links bewtween those genes.
+
+As an representative example for we chose the Glycolysis/Gluconeogenesis pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.
+
 \section*{Results}
@@
 
-To illustrate the $p$~value calculation procedure we plotted the score distribution N(s) for a query size of 10 genes in Figure \ref{fig:score_dist}.
+To illustrate the $p$~value calculation procedure we plotted the score distribution N(s) for a query size of 25 genes in Figure \ref{fig:score_dist}.
 
@@
 		\end{center}
-%\includegraphics{img/FIXME.png}
-  \caption{{\bf Score distribution of a query size of 10 genes.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
+%FIXME: Do not use figure numbers in their file name, this creates havoc if you want to reorganize them later
+  \caption{{\bf Score distribution of a query size of 25 genes against the Glycolysis/Gluconeogenesis pathway.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
 how many ways one can reach a certain score when adding up the scores for $Q$ genes.}
@@
 %\includegraphics{img/FIXME.png}
-  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition, and calculated their $p$ values with GeneSetDP.
-To test the $p$~values uniformity we plotted them against their quantile.
+  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition, and calculated their $p$ values with GeneSetDP against the Glycolysis/Gluconeogenesis pathway.
+To test the $p$~values uniformity we plotted them against their Normalized rank.
 We also added the calibration curve of BinoX, evaluated on the same random sets.}
@@
 
+It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.
+
 Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
@@
 In practice such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics.
+In general, it is easier to understand null models that relates to user actions, than null models that relates to model parameters.
 

### 2018-05-02 doc/genesetdp.tex
Spelling

@@
 \bibliographystyle{plainurl}
-\title{A dynamic programing algorithm to assess the reliability of inferences from network enrichment analysis}
-\titlerunning{Dynamic Programing for Network Enrichment inferences} %optional, in case that the title is too long; the running title should fit into the top page column
+\title{A dynamic programming algorithm to assess the reliability of inferences from network enrichment analysis}
+\titlerunning{Dynamic Programming for Inferences from Network Enrichment Analysis} %optional, in case that the title is too long; the running title should fit into the top page column
 
@@
 Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}.
-The technique estimates which functional modules, such as complexes or pathways, that are overrepresentated among a set of identified analytes.
+The technique estimates which functional modules, such as complexes or pathways, that are overrepresented among a set of identified analytes.
 One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.
@@
 
-A limiting factor of GEA are the pathway definition databases. While \url{http://pathguide.org} list 708 databases of
-pathway definitions\cite{bader2006pathguide},  some chritique has been voiced concerning the completeness and rigor of current pathway databases.
+A limiting factor of GEA is the pathway definition databases.
+While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases.
 Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string}.
@@
 
-Here, we present a dynamic programing algorithm, that we dubbed GeneSetDP, that calculate the full score distribution of any query of a given size.
-The algorithm enables us to calculate well defined and exact statistics of inferences from NEA.
+Here, we present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the score distribution of any query of a given size.
+The algorithm enables us to calculate well defined statistics of inferences from NEA.
 Using simulations we demonstrated that the algorithm produced unbiased statistics.
@@
 We also downloaded the NEA example given at the BinoX web site.
-The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways, and a network definition file that, after tresholding with a score of $0.7$, yeilds $1244992$ links bewtween those genes.
+The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways, and a network definition file that, after thresholding with a score of $0.7$, yeilds $1244992$ links between those genes.
 
-As an representative example for we chose the Glycolysis/Gluconeogenesis pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.
+As a representative example for we chose the Glycolysis/Gluconeogenesis pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.
 
@@
 
-We implemented a Python program that read network and pathway definition files, and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
+We implemented a Python program that read network and pathway definition files and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
 We downloaded pathway and network definitions from the BinoX's website and used the same network treshold as BinoX default ($0.7$).
@@
 In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets from the investigated genome.
-As our selection is random, and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed.
+As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed.
 To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration}.
-As a comparison we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.
+As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.
 
@@
 %\includegraphics{img/FIXME.png}
-  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition, and calculated their $p$ values with GeneSetDP against the Glycolysis/Gluconeogenesis pathway.
+  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition and calculated their $p$ values with GeneSetDP against the Glycolysis/Gluconeogenesis pathway.
 To test the $p$~values uniformity we plotted them against their Normalized rank.
@@
 
-It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.
+It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand then the theory of how to permute networks.
 
 Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
-Previous implementation of NEA in the literature otherwise use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.
-In practice such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics.
-In general, it is easier to understand null models that relates to user actions, than null models that relates to model parameters.
+Previous implementations of NEA in the literature otherwise use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.
+In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics.
+In general, it is easier to understand null models that relate to user actions, than null models that relate to model parameters.
 

### 2018-05-02 doc/genesetdp.tex
More typos

@@
 Here, we present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the score distribution of any query of a given size.
-The algorithm enables us to calculate well defined statistics of inferences from NEA.
+The algorithm enables us to calculate well-defined statistics of inferences from NEA.
 Using simulations we demonstrated that the algorithm produced unbiased statistics.
@@
 In NEA the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.
-The pathways are scored by summing up all conections, i.e.
+The pathways are scored by summing up all connections, i.e.
 the pathway score is
@@
 Here, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$.
-This allows us to evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
+This allows us to evaluate our score against the null hypothesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
 This can be expressed as a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score $s$,
@@
 
-Such number-of-links, $l_i$, can be precomputed ahead of time for all genes $i, \ldots, G$.
+Such number-of-links, $l_i$, can be pre-computed for all genes $i, \ldots, G$.
 Furthermore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{i}{l_i}$.
@@
 
-We dowloaded network definitions and the BinoX software (on 2018-04-28) for comparisons from \url{https://bitbucket.org/sonnhammergroup/binox}. BinoX was run with the default parameters.
+We downloaded network definitions and the BinoX software (on 2018-04-28) for comparisons from \url{https://bitbucket.org/sonnhammergroup/binox}. BinoX was run with the default parameters.
 % FIXME: add a more detailed description of the test set (i.e. organism and number of genes, pathways and network connections).
@@
 We also downloaded the NEA example given at the BinoX web site.
-The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways, and a network definition file that, after thresholding with a score of $0.7$, yeilds $1244992$ links between those genes.
+The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways and a network definition file that, after thresholding with a score of $0.7$, gave $1244992$ links between those genes.
 
-As a representative example for we chose the Glycolysis/Gluconeogenesis pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.
+As a representative example for we chose the ``Glycolysis/Gluconeogenesis'' pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.
 
@@
 We implemented a Python program that read network and pathway definition files and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
-We downloaded pathway and network definitions from the BinoX's website and used the same network treshold as BinoX default ($0.7$).
+We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).
 
@@
 
-It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand then the theory of how to permute networks.
+It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.
 

### 2018-05-03 doc/genesetdp.tex
small edits.

@@
 \maketitle
+
+\begin{abstract}
+FIXME
+\end{abstract}
 
@@
 %FIXME: Do not use figure numbers in their file name, this creates havoc if you want to reorganize them later
-  \caption{{\bf Score distribution of a query size of 25 genes against the Glycolysis/Gluconeogenesis pathway.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
+  \caption{{\bf Score distribution of a query size of 25 genes against the ``Glycolysis/Gluconeogenesis'' pathway.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
 how many ways one can reach a certain score when adding up the scores for $Q$ genes.}
@@
 
-We note that the calibration of
+We note that the calibration of GeneSetDP is slightly conservative, i.e.
+the calculated $p$ values are larger than expected.
+Meanwhile, BinoX appears strongly anti-conservative, that is, the $p$ values are much lower than expected.
 
@@
 %\includegraphics{img/FIXME.png}
-  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition and calculated their $p$ values with GeneSetDP against the Glycolysis/Gluconeogenesis pathway.
+  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition and calculated their $p$ values with GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
 To test the $p$~values uniformity we plotted them against their Normalized rank.

### 2018-05-03 exp/example-random/generate_random.py
new test with query size 25

@@ -1,30 +1,34 @@
 import pandas as pd
 import numpy as np
 
-path = "~/git/binox/example/"
-
-original = pd.read_csv(path + 'Signatures.tsv', sep='\t')
+original = pd.read_csv('Signatures.tsv', sep='\t')
 
 genes = np.unique(original.index)
 
 random_signatures = pd.DataFrame()
 
-for i in np.arange(1,10001,1):
-    sample = np.random.choice(len(genes),10, replace=False)
+n_samples = 10000
+n_genes = 25
+
+for i in np.arange(1,n_samples+1,1):
+    sample = np.random.choice(len(genes),n_genes, replace=False)
     sample_genes = genes[sample]
-    new_group = pd.DataFrame(np.transpose([sample_genes, np.repeat('group' + str(i),10)]))
+    new_group = pd.DataFrame(np.transpose([sample_genes, np.repeat('group' + str(i), n_genes)]))
     random_signatures = random_signatures.append(new_group)
 
 random_signatures.columns = ['#Node', 'GroupName']
-random_signatures.set_index('#Node').to_csv('Random_Signatures.tsv', sep = '\t')
+
+
+random_signatures.set_index('#Node').to_csv('Random_Signatures_25.tsv', sep = '\t')
 
 
 #####
 
-pathways = pd.read_csv(path + 'Pathways.tsv', sep='\t', header=None)
+pathways = pd.read_csv('Pathways.tsv', sep='\t', header=None)
 
 pathway_name = 'GLYCOLYSIS_/_GLUCONEOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
 
 one_pathway = pathways.where(pathways.iloc[:,1] == pathway_name).dropna()
 
 one_pathway.to_csv('Pathway_One.tsv', sep='\t', header=None, index=False)
+

### 2018-05-03 exp/example-random/combine_plot.py
plot

@@ -4,8 +4,6 @@ import matplotlib.pyplot as plt
 import seaborn as sns
 
 
-genesetdp = pd.read_csv('genesetDP_results.tsv', sep='\t', index_col=0)
-
 binox = pd.read_csv('Pathway_One_vs_Random_Signatures.tsv', sep='\t')
 binox = binox.set_index('#3:NameGroupB')
 binox.index = [x.lower() for x in binox.index]
@@ -42,4 +40,4 @@ plt.legend(['BinoX', 'GeneSetDP', 'y=x', 'y=0.5x,y=2x'])
 plt.ylabel('$p$ value')
 plt.xlabel('Normalized rank')
 
-plt.savefig('F2_calibration.png')
+plt.savefig('calibration.png')

### 2018-05-03 exp/example-random/generate_random.py
speed up generation

@@ -1,7 +1,7 @@
 import pandas as pd
 import numpy as np
 
-original = pd.read_csv('Signatures.tsv', sep='\t')
+original = pd.read_csv('../../../binox/example/Signatures.tsv', sep='\t')
 
 genes = np.unique(original.index)
 
@@ -11,8 +11,7 @@ n_samples = 10000
 n_genes = 25
 
 for i in np.arange(1,n_samples+1,1):
-    sample = np.random.choice(len(genes),n_genes, replace=False)
-    sample_genes = genes[sample]
+    sample_genes = np.random.choice(genes, n_genes)
     new_group = pd.DataFrame(np.transpose([sample_genes, np.repeat('group' + str(i), n_genes)]))
     random_signatures = random_signatures.append(new_group)
 

### 2018-05-03 exp/example-random/generate_random.py
fix sampling!

@@ -8,17 +8,17 @@ genes = np.unique(original.index)
 random_signatures = pd.DataFrame()
 
 n_samples = 10000
-n_genes = 25
+n_genes = 10
 
 for i in np.arange(1,n_samples+1,1):
-    sample_genes = np.random.choice(genes, n_genes)
+    sample_genes = np.random.choice(genes, n_genes, replace=False)
     new_group = pd.DataFrame(np.transpose([sample_genes, np.repeat('group' + str(i), n_genes)]))
     random_signatures = random_signatures.append(new_group)
 
 random_signatures.columns = ['#Node', 'GroupName']
 
 
-random_signatures.set_index('#Node').to_csv('Random_Signatures_25.tsv', sep = '\t')
+random_signatures.set_index('#Node').to_csv('Random_Signatures.tsv', sep = '\t')
 
 
 #####

### 2018-05-03 exp/combine_plot.py
results with new sampling

@@ -1,0 +1,43 @@
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+
+binox = pd.read_csv('example-random//Pathway_One_vs_Random_Signatures.tsv', sep='\t')
+binox = binox.set_index('#3:NameGroupB')
+binox.index = [x.lower() for x in binox.index]
+
+binox = binox['#4:p.value']
+
+genesetdp = pd.read_csv('example-random/genesetDP_results.tsv', sep='\t', index_col=0)
+
+binox = binox.loc[genesetdp.index]
+binox.loc[np.isnan(binox)] = 1
+
+binox_sorted = binox.sort_values()
+binox_sorted.index = (np.arange(len(binox_sorted.index))+1)/len(binox_sorted.index)
+
+genesetdp_sorted = genesetdp['p'].sort_values()
+genesetdp_sorted.index = (np.arange(len(genesetdp_sorted.index))+1)/len(genesetdp_sorted)
+
+
+x_y = pd.DataFrame(np.arange(0,1,0.001), index = np.arange(0,1,0.001))
+x_2y = pd.DataFrame(np.arange(0,1,0.001)*2, index = np.arange(0,1,0.001))
+x_halfy = pd.DataFrame(np.arange(0,0.5,0.001), index = np.arange(0,0.5,0.001)*2)
+
+sns.set_style("ticks")
+sns.despine()
+
+
+plt.loglog(binox_sorted, '.')
+plt.loglog(genesetdp_sorted, '.')
+plt.loglog(x_y, '--', color = 'r', alpha = 0.5)
+plt.loglog(x_2y, ':',color = 'b', alpha = 0.5)
+plt.loglog(x_halfy, ':',color = 'b', alpha = 0.5)
+
+plt.legend(['BinoX', 'GeneSetDP', 'y=x', 'y=0.5x,y=2x'])
+plt.ylabel('$p$ value')
+plt.xlabel('Normalized rank')
+
+plt.savefig('calibration.png')

### 2018-05-03 exp/example-random/combine_plot.py

@@ -1,43 +1,0 @@
-import pandas as pd
-import numpy as np
-import matplotlib.pyplot as plt
-import seaborn as sns
-
-
-binox = pd.read_csv('Pathway_One_vs_Random_Signatures.tsv', sep='\t')
-binox = binox.set_index('#3:NameGroupB')
-binox.index = [x.lower() for x in binox.index]
-
-
-binox = binox['#4:p.value']
-genesetdp = pd.read_csv('genesetDP_results.tsv', sep='\t', index_col=0)
-
-binox = binox.loc[genesetdp.index]
-binox.loc[np.isnan(binox)] = 1
-
-binox_sorted = binox.sort_values()
-binox_sorted.index = (np.arange(len(binox_sorted.index))+1)/len(binox_sorted.index)
-
-genesetdp_sorted = genesetdp['p'].sort_values()
-genesetdp_sorted.index = (np.arange(len(genesetdp_sorted.index))+1)/len(genesetdp_sorted)
-
-
-x_y = pd.DataFrame(np.arange(0,1,0.001), index = np.arange(0,1,0.001))
-x_2y = pd.DataFrame(np.arange(0,1,0.001)*2, index = np.arange(0,1,0.001))
-x_halfy = pd.DataFrame(np.arange(0,0.5,0.001), index = np.arange(0,0.5,0.001)*2)
-
-sns.set_style("ticks")
-sns.despine()
-
-
-plt.loglog(binox_sorted, '.')
-plt.loglog(genesetdp_sorted, '.')
-plt.loglog(x_y, '--', color = 'r', alpha = 0.5)
-plt.loglog(x_2y, ':',color = 'b', alpha = 0.5)
-plt.loglog(x_halfy, ':',color = 'b', alpha = 0.5)
-
-plt.legend(['BinoX', 'GeneSetDP', 'y=x', 'y=0.5x,y=2x'])
-plt.ylabel('$p$ value')
-plt.xlabel('Normalized rank')
-
-plt.savefig('calibration.png')

### 2018-05-03 exp/experiment.py

@@ -20,7 +20,7 @@ pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.v
 
 k = genk.generate_k(pathway_genes)
 
-N = genesetdp.genesetdp(k,25)
+N = genesetdp.genesetdp(k,10)
 
 p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
 
@@ -35,12 +35,12 @@ sns.despine()
 plt.step(np.arange(len(N)), N, where='mid')
 plt.xlabel('Score, $s$')
 
-plt.savefig('F1_score_distribuition.png')
+plt.savefig('score_distribuition.png')
 
 
 ####
 
-groups = pd.read_csv('../src/example-random/Random_Signatures.tsv', sep='\t')
+groups = pd.read_csv('../exp/example-random/Random_Signatures.tsv', sep='\t')
 
 network = genk.network
 

### 2018-05-03 exp/combine_plot.py
new plots

@@ -4,7 +4,7 @@ import matplotlib.pyplot as plt
 import seaborn as sns
 
 
-binox = pd.read_csv('example-random//Pathway_One_vs_Random_Signatures.tsv', sep='\t')
+binox = pd.read_csv('example-random/BinoX_results.tsv', sep='\t')
 binox = binox.set_index('#3:NameGroupB')
 binox.index = [x.lower() for x in binox.index]
 

### 2018-05-03 exp/experiment.py

@@ -20,7 +20,7 @@ pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.v
 
 k = genk.generate_k(pathway_genes)
 
-N = genesetdp.genesetdp(k,10)
+N = genesetdp.genesetdp(k,25)
 
 p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
 
@@ -33,6 +33,7 @@ sns.set_style("ticks")
 sns.despine()
 
 plt.step(np.arange(len(N)), N, where='mid')
+plt.ylabel('$N(s)$')
 plt.xlabel('Score, $s$')
 
 plt.savefig('score_distribuition.png')

### 2018-05-03 exp/combine_plot.py
option for only enriched

@@ -5,7 +5,15 @@ import seaborn as sns
 
 
 binox = pd.read_csv('example-random/BinoX_results.tsv', sep='\t')
+
+###### only enriched
+binox.loc[binox['#6:relationType'] == '-', '#4:p.value'] = 1 - binox.loc[binox['#6:relationType'] == '-', '#4:p.value']
+######
+
+
+
 binox = binox.set_index('#3:NameGroupB')
+
 binox.index = [x.lower() for x in binox.index]
 
 binox = binox['#4:p.value']

### 2018-05-03 doc/genesetdp.tex
Not sure why this is needed.
Replacing lost header
readded documment class
new plots
text

@@
 \documentclass[a4paper,american]{lipics-v2016}
-%\usepackage[USenglish]{babel}
-%\usepackage{microtype}
+% \usepackage[USenglish]{babel}
+% \usepackage{microtype}
 \bibliographystyle{plainurl}
@@
 
-The quality of the inferences from NEA has previously been assessed by Monte Carlo-simulations that investigates the scores distribution when evaluating the query set using random networks.
+FIXME: Review the mentioned methods. Particularly which null models do they use?
+
+The quality of the inferences from NEA has previously been assessed by Monte Carlo-simulations that investigates the scores distribution when evaluating the query set using random networks which are generated using methods with varying degrees of topological conservation.
+
+Since these methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desired, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic.
+And the somewhat inaccessible process of so randomising networks produces statistics that are equaly hard to interpret.
+FIXME: Should this be in the dicussion?
 
@@
 		\begin{center}
-				\includegraphics[scale=0.8]{figures/F1_score_distribuition.png}
+				\includegraphics[scale=0.8]{figures/score_distribuition.png}
 		\end{center}
@@
 		\begin{center}
-				\includegraphics[scale=0.8]{figures/F2_calibration.png}
+				\includegraphics[scale=0.8]{figures/calibration.png}
 		\end{center}
@@
 
+FIXME: Is it ok to state the method is more accurate if we are testing it agains a different null hypothesis
+
+FIXME: What about:
+
 It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.

### 2018-05-03 doc/genesetdp.tex
text

@@
 Since these methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desired, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic.
-And the somewhat inaccessible process of so randomising networks produces statistics that are equaly hard to interpret.
+Also, the somewhat inaccessible process used to randomize networks produces statistics that are equaly hard to interpret as the null hypothesis becomes complex.
 FIXME: Should this be in the dicussion?
@@
 
-We implemented a Python program that read network and pathway definition files and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
+We implemented a Python program that reads network and pathway definition files and scored query sets agains a pathway according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
 We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).
@@
 
-FIXME: What about:
-
 It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.

### 2018-05-03 doc/genesetdp.tex
text

@@
 
-Most GEA methods are investigating the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test.
+Most GEA methods are assessing the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test.
 However, variants such as Gene Set Enrichment Analysis (GSEA)\cite{subramanian2005gene} also includes information on expression levels of the analytes of the query set.
@@
 
-Since these methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desired, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic.
+Since these methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desirable, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic.
 Also, the somewhat inaccessible process used to randomize networks produces statistics that are equaly hard to interpret as the null hypothesis becomes complex.
@@
 
-As a representative example for we chose the ``Glycolysis/Gluconeogenesis'' pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.
+As a representative example for we chose the ``Glycolysis/Gluconeogenesis'' pathway, as it contained a number of genes that coincided with the median number of genes of the pathways in the definition file.
 
@@
 
-FIXME: Is it ok to state the method is more accurate if we are testing it agains a different null hypothesis
+FIXME: Is it ok to state the method is more accurate if we are testing it agains a different null hypothesis?
 
-It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.
+It should be noted that our outlined method a could work well with Monte Carlo-sampling, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.
 

### 2018-05-04 exp/combine_plot.py
Rearangements

@@ -24,10 +24,10 @@ binox = binox.loc[genesetdp.index]
 binox.loc[np.isnan(binox)] = 1
 
 binox_sorted = binox.sort_values()
-binox_sorted.index = (np.arange(len(binox_sorted.index))+1)/len(binox_sorted.index)
+binox_ix = (np.arange(len(binox_sorted.index))+0.5)/float(len(binox_sorted.index)+1)
 
 genesetdp_sorted = genesetdp['p'].sort_values()
-genesetdp_sorted.index = (np.arange(len(genesetdp_sorted.index))+1)/len(genesetdp_sorted)
+genesetdp_ix = (np.arange(len(genesetdp_sorted.index))+0.5)/float(len(genesetdp_sorted)+1)
 
 
 x_y = pd.DataFrame(np.arange(0,1,0.001), index = np.arange(0,1,0.001))
@@ -38,8 +38,8 @@ sns.set_style("ticks")
 sns.despine()
 
 
-plt.loglog(binox_sorted, '.')
-plt.loglog(genesetdp_sorted, '.')
+plt.loglog(binox_ix, binox_sorted, '.')
+plt.loglog(genesetdp_ix, genesetdp_sorted,'.')
 plt.loglog(x_y, '--', color = 'r', alpha = 0.5)
 plt.loglog(x_2y, ':',color = 'b', alpha = 0.5)
 plt.loglog(x_halfy, ':',color = 'b', alpha = 0.5)

### 2018-05-04 exp/experiment.py

@@ -24,24 +24,7 @@ N = genesetdp.genesetdp(k,25)
 
 p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
 
-####
-
-import matplotlib.pyplot as plt
-import seaborn as sns
-
-sns.set_style("ticks")
-sns.despine()
-
-plt.step(np.arange(len(N)), N, where='mid')
-plt.ylabel('$N(s)$')
-plt.xlabel('Score, $s$')
-
-plt.savefig('score_distribuition.png')
-
-
-####
-
-groups = pd.read_csv('../exp/example-random/Random_Signatures.tsv', sep='\t')
+groups = pd.read_csv('example-random/Random_Signatures.tsv', sep='\t')
 
 network = genk.network
 
@@ -52,8 +35,9 @@ results = pd.DataFrame(columns = ['n','p'])
 for i in range(1,10001):
     groupname = 'group' + str(i)
     genes = groups.loc[groups.GroupName==groupname,'#Node']
+    print(linked_genes.count(genes.iloc[0]))
     n = sum(linked_genes.count(x) for x in genes)
     p = p_vals[n]
     results.loc[groupname] = [n,p]
 
-results.to_csv('genesetDP_results.tsv', sep = '\t')
+results.to_csv('example-random/genesetDP_results.tsv', sep = '\t')

### 2018-05-04 exp/figure_calibration.py

@@ -1,0 +1,37 @@
+import pandas as pd
+import numpy as np
+
+import sys
+sys.path.insert(0, '../src')
+import generate_k as genk
+import genesetdp
+
+## example path
+path = "~/git/binox/example/"
+
+## read original patway information file
+pathways_file = 'Pathways.tsv'
+pathways = pd.read_csv(path + pathways_file, sep='\t', header=None)
+pathways.columns = ['gene','pathway']
+
+pathway_name = 'GLYCOLYSIS_/_GLUCONEOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
+
+pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.values
+
+k = genk.generate_k(pathway_genes)
+
+N = genesetdp.genesetdp(k,25)
+
+####
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+sns.set_style("ticks")
+sns.despine()
+
+plt.step(np.arange(len(N)), N, where='mid')
+plt.ylabel('$N(s)$')
+plt.xlabel('Score, $s$')
+
+plt.savefig('score_distribuition.png')

### 2018-05-04 exp/generate_random.py

@@ -1,14 +1,14 @@
 import pandas as pd
 import numpy as np
 
-original = pd.read_csv('../../../binox/example/Signatures.tsv', sep='\t')
+original = pd.read_csv('~/git/binox/example/Signatures.tsv', sep='\t')
 
 genes = np.unique(original.index)
 
 random_signatures = pd.DataFrame()
 
 n_samples = 10000
-n_genes = 10
+n_genes = 25
 
 for i in np.arange(1,n_samples+1,1):
     sample_genes = np.random.choice(genes, n_genes, replace=False)
@@ -18,16 +18,15 @@ for i in np.arange(1,n_samples+1,1):
 random_signatures.columns = ['#Node', 'GroupName']
 
 
-random_signatures.set_index('#Node').to_csv('Random_Signatures.tsv', sep = '\t')
+random_signatures.set_index('#Node').to_csv('example-random/Random_Signatures.tsv', sep = '\t')
 
 
 #####
 
-pathways = pd.read_csv('Pathways.tsv', sep='\t', header=None)
+pathways = pd.read_csv('~/git/binox/example/Pathways.tsv', sep='\t', header=None)
 
 pathway_name = 'GLYCOLYSIS_/_GLUCONEOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
 
 one_pathway = pathways.where(pathways.iloc[:,1] == pathway_name).dropna()
 
-one_pathway.to_csv('Pathway_One.tsv', sep='\t', header=None, index=False)
-
+one_pathway.to_csv('example-random/Pathway_One.tsv', sep='\t', header=None, index=False)

### 2018-05-04 exp/generate_random.py
generate based on network

@@ -1,9 +1,9 @@
 import pandas as pd
 import numpy as np
 
-original = pd.read_csv('~/git/binox/example/Signatures.tsv', sep='\t')
+original = pd.read_csv('~/git/binox/example/Network.tsv', sep='\t')
 
-genes = np.unique(original.index)
+genes = np.unique(original['#2:Gene1'])
 
 random_signatures = pd.DataFrame()
 

### 2018-05-04 exp/experiment.py
consistent network direction

@@ -28,14 +28,14 @@ groups = pd.read_csv('example-random/Random_Signatures.tsv', sep='\t')
 
 network = genk.network
 
-linked_genes = network.loc[pathway_genes].dropna().gene2.values.tolist()
+# linked_genes = network.loc[pathway_genes].dropna().gene2.values.tolist()
+linked_genes = network.index[network.gene2.isin(pathway_genes)].tolist()
 
 results = pd.DataFrame(columns = ['n','p'])
 ## start for
 for i in range(1,10001):
     groupname = 'group' + str(i)
     genes = groups.loc[groups.GroupName==groupname,'#Node']
-    print(linked_genes.count(genes.iloc[0]))
     n = sum(linked_genes.count(x) for x in genes)
     p = p_vals[n]
     results.loc[groupname] = [n,p]

### 2018-05-04 src/generate_k.py

@@ -20,9 +20,10 @@ network = network.set_index('gene1')
 n_genes = len(np.unique(network.index))
 
 def generate_k(pathway_genes):
-    links = network.loc[pathway_genes].dropna()
+    # links = network.loc[pathway_genes].dropna()
+    links = network.index[network.gene2.isin(pathway_genes)]
 
-    k_proto = links.gene2.value_counts().value_counts()
+    k_proto = links.value_counts().value_counts()
 
     k = k_proto.loc[list(range(max(k_proto.index)+1))]
     k[k.isna()] = 0

### 2018-05-04 exp/combine_plot.py
readability

@@ -19,16 +19,13 @@ binox.index = [x.lower() for x in binox.index]
 binox = binox['#4:p.value']
 
 genesetdp = pd.read_csv('example-random/genesetDP_results.tsv', sep='\t', index_col=0)
-
-binox = binox.loc[genesetdp.index]
-binox.loc[np.isnan(binox)] = 1
-
-binox_sorted = binox.sort_values()
-binox_ix = (np.arange(len(binox_sorted.index))+0.5)/float(len(binox_sorted.index)+1)
-
 genesetdp_sorted = genesetdp['p'].sort_values()
 genesetdp_ix = (np.arange(len(genesetdp_sorted.index))+0.5)/float(len(genesetdp_sorted)+1)
 
+binox = binox.loc[genesetdp.index]
+binox.loc[np.isnan(binox)] = 1
+binox_sorted = binox.sort_values()
+binox_ix = (np.arange(len(binox_sorted.index))+0.5)/float(len(binox_sorted.index)+1)
 
 x_y = pd.DataFrame(np.arange(0,1,0.001), index = np.arange(0,1,0.001))
 x_2y = pd.DataFrame(np.arange(0,1,0.001)*2, index = np.arange(0,1,0.001))

### 2018-05-04 doc/genesetdp.tex
x
Added abstract

@@
 \documentclass[a4paper,american]{lipics-v2016}
-% \usepackage[USenglish]{babel}
-% \usepackage{microtype}
+%\usepackage[USenglish]{babel}
+%\usepackage{microtype}
 \bibliographystyle{plainurl}
@@
 \begin{abstract}
-FIXME
+A prevailing technique to infer function from findings in high-throughput experiments is gene set enrichemnet analysis, where the sets of genes are compared to predefined sets of related genes, so called pathways.
+As some of such pathways are known to be incomplete, algoritmic efforts have been made to complement them with information from functional association networks, using so called Network Enrichment Analysis (NEA).
+Traditionally significance of inferences from NEA has been asssigned using network perturbations.
+Here we instead have designed a dynamic programming  algorithm that can calculte the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences.
+We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engin, NEA.
+
 \end{abstract}
@@
 Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string}.
-Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{shojaie2010network, alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}.
+Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}.
 This is known as Network Enrichment Analysis (NEA).
 
-FIXME: Review the mentioned methods. Particularly which null models do they use?
+The significance of the inferences from NEA are assessed by null models that investigate the number of expected random links from the query geneset to the pathway.
+Previous efforts have settled for varius parametric null distributions, where the parameters are estimated by Monte Carlo-simulations that permute the network definitions by methods with varying degrees of topological conservation.
 
-The quality of the inferences from NEA has previously been assessed by Monte Carlo-simulations that investigates the scores distribution when evaluating the query set using random networks which are generated using methods with varying degrees of topological conservation.
-
-Since these methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desirable, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic.
+Since such methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desirable, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic.
 Also, the somewhat inaccessible process used to randomize networks produces statistics that are equaly hard to interpret as the null hypothesis becomes complex.

### 2018-05-04 exp/combine_plot.py
Merge branch 'master' of github.com:statisticalbiotechnology/genesetdp

@@ -20,12 +20,12 @@ binox = binox['#4:p.value']
 
 genesetdp = pd.read_csv('example-random/genesetDP_results.tsv', sep='\t', index_col=0)
 genesetdp_sorted = genesetdp['p'].sort_values()
-genesetdp_ix = (np.arange(len(genesetdp_sorted.index))+0.5)/float(len(genesetdp_sorted)+1)
+genesetdp_ix = (np.arange(len(genesetdp_sorted.index))+0.5)/float(len(genesetdp_sorted))
 
 binox = binox.loc[genesetdp.index]
 binox.loc[np.isnan(binox)] = 1
 binox_sorted = binox.sort_values()
-binox_ix = (np.arange(len(binox_sorted.index))+0.5)/float(len(binox_sorted.index)+1)
+binox_ix = (np.arange(len(binox_sorted.index))+0.5)/float(len(binox_sorted.index))
 
 x_y = pd.DataFrame(np.arange(0,1,0.001), index = np.arange(0,1,0.001))
 x_2y = pd.DataFrame(np.arange(0,1,0.001)*2, index = np.arange(0,1,0.001))

### 2018-05-04 exp/experiment.py

@@ -1,10 +1,16 @@
 import pandas as pd
 import numpy as np
+from scipy.special import comb
 
 import sys
 sys.path.insert(0, '../src')
 import generate_k as genk
 import genesetdp
+
+if len(sys.argv)>1:
+    Q=int(sys.argv[1])
+else:
+    Q=25
 
 ## example path
 path = "~/git/binox/example/"
@@ -20,7 +26,9 @@ pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.v
 
 k = genk.generate_k(pathway_genes)
 
-N = genesetdp.genesetdp(k,25)
+N = genesetdp.genesetdp(k,Q)
+#assert(float(sum(N))==float(genesetdp.ncr(sum(k),Q)))
+assert(float(sum(N))==float(comb(sum(k),Q,exact=False)))
 
 p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
 

### 2018-05-04 exp/generate_random.py

@@ -1,5 +1,11 @@
+import sys
 import pandas as pd
 import numpy as np
+
+if len(sys.argv)>1:
+    Q=int(sys.argv[1])
+else:
+    Q=25
 
 original = pd.read_csv('~/git/binox/example/Network.tsv', sep='\t')
 
@@ -8,7 +14,7 @@ genes = np.unique(original['#2:Gene1'])
 random_signatures = pd.DataFrame()
 
 n_samples = 10000
-n_genes = 25
+n_genes = Q
 
 for i in np.arange(1,n_samples+1,1):
     sample_genes = np.random.choice(genes, n_genes, replace=False)

### 2018-05-04 src/genesetdp.py

@@ -1,13 +1,6 @@
 #/usr/bin python
 import numpy as np
-import operator as op
-import functools
-
-def ncr(n, r):
-    r = min(r, n-r)
-    numer = functools.reduce(op.mul, range(n, n-r, -1), 1)
-    denom = functools.reduce(op.mul, range(1, r+1), 1)
-    return numer//denom
+from scipy.special import comb
 
 def find_maxscore(k,q):
     max_s, ix, rem = 0, len(k)-1, q
@@ -29,12 +22,21 @@ def genesetdp(k,q):
             for c in range(q,-1,-1):
                 for b in range(k[a],0,-1):
                     if c-b>=0 and s-a*b>=0:
-                        N[s,c] += ncr(k[a], b) * N[s-a*b,c-b]
+                        N[s,c] += comb(k[a], b, exact=False) * N[s-a*b,c-b]
     return N[:,q]
 
 if __name__ == "__main__":
     # Query size q, number of links k[a]
-    q = 8
-    k = np.array([8,5,3,3,4],dtype='u4')
+    q = 2
+    k = np.array([4,2,1,1],dtype='u4')
     #k = np.array([3,3,3],dtype='u4')
-    print(genesetdp(k,q))
+    N=genesetdp(k,q)
+    p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
+    x,sk,ix=[],float(sum(k)),0
+    for ki in k:
+        for _ in range(ki):
+            x.insert(0,(ix+0.5)/(sk))
+            ix += 1
+    print(N)
+    print(p_vals)
+    print(x)

### 2018-05-04 src/genesetdp_monte_carlo.py
monte carlo

@@ -1,0 +1,17 @@
+import pandas as pd
+import numpy as np
+from genesetdp import find_maxscore
+
+def genesetmc(k,q, n_runs):
+
+    maxscore = find_maxscore(k,q)
+    N_mc = np.zeros(maxscore+10)
+
+    for i in range(n_runs):
+        sample = sum(np.random.choice(np.arange(len(k)), q, p = np.array(k)/sum(k)))
+        N_mc[sample] += 1
+
+
+    p_vals = (np.flipud(np.cumsum(np.flipud(N_mc))) - N_mc/2)/sum(N_mc)
+
+    return p_vals

### 2018-05-04 doc/genesetdp.tex

@@
 \begin{abstract}
-A prevailing technique to infer function from findings in high-throughput experiments is gene set enrichemnet analysis, where the sets of genes are compared to predefined sets of related genes, so called pathways.
-As some of such pathways are known to be incomplete, algoritmic efforts have been made to complement them with information from functional association networks, using so called Network Enrichment Analysis (NEA).
-Traditionally significance of inferences from NEA has been asssigned using network perturbations.
-Here we instead have designed a dynamic programming  algorithm that can calculte the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences.
-We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engin, NEA.
+A prevailing technique to infer function from identifications from molecular biological high-throughput experiments is gene set enrichment analysis, where the identifications are compared to predefined sets of related genes, so-called pathways.
+As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks, with so-called Network Enrichment Analysis (NEA).
+Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations.
+Here we instead have designed a dynamic programming  algorithm that can calculate the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences.
+We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX.
 
@@
 
-The significance of the inferences from NEA are assessed by null models that investigate the number of expected random links from the query geneset to the pathway.
-Previous efforts have settled for varius parametric null distributions, where the parameters are estimated by Monte Carlo-simulations that permute the network definitions by methods with varying degrees of topological conservation.
+The significance of the inferences from NEA is assessed by null models that investigate the number of expected random links from the query gene set to the pathway.
+Previous efforts have settled for various parametric null distributions, where the parameters are estimated by Monte Carlo-simulations that permute the network definitions by methods with varying degrees of topological conservation.
 
 Since such methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desirable, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic.
-Also, the somewhat inaccessible process used to randomize networks produces statistics that are equaly hard to interpret as the null hypothesis becomes complex.
-FIXME: Should this be in the dicussion?
+Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret as the null hypothesis becomes complex.
 
@@
 
-This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elemets from $N_{a-1}$ still needed to update subsequent elements.
-That is, we do not have to copy the dynamic programing matrix in each iteration over $a$, instead we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$.
+This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elements from $N_{a-1}$ still needed to update subsequent elements.
+That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$.
 For details see Listing \ref{lst:gensetdp}.
@@
 
-We implemented a Python program that reads network and pathway definition files and scored query sets agains a pathway according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
+We implemented a Python program that reads network and pathway definition files and scored query sets against a pathway according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
 We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).
@@
 As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed.
-To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration}.
+To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration} for $10000$ randomly assembeled queries of $25$ genes from the human genome.
 As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.
@@
 
-FIXME: Is it ok to state the method is more accurate if we are testing it agains a different null hypothesis?
-
 It should be noted that our outlined method a could work well with Monte Carlo-sampling, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.
@@
 In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics.
-In general, it is easier to understand null models that relate to user actions, than null models that relate to model parameters.
+In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.
 

### 2018-05-04 exp/experiment.py
monte carlo

@@ -7,6 +7,8 @@ sys.path.insert(0, '../src')
 import generate_k as genk
 import genesetdp
 
+import genesetdp_monte_carlo as mc
+
 if len(sys.argv)>1:
     Q=int(sys.argv[1])
 else:
@@ -31,6 +33,8 @@ N = genesetdp.genesetdp(k,Q)
 assert(float(sum(N))==float(comb(sum(k),Q,exact=False)))
 
 p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
+
+p_vals_mc = mc.genesetmc(k, Q, 100000)
 
 groups = pd.read_csv('example-random/Random_Signatures.tsv', sep='\t')
 

### 2018-05-04 exp/experiment.py
monte carlo

@@ -43,13 +43,14 @@ network = genk.network
 # linked_genes = network.loc[pathway_genes].dropna().gene2.values.tolist()
 linked_genes = network.index[network.gene2.isin(pathway_genes)].tolist()
 
-results = pd.DataFrame(columns = ['n','p'])
+results = pd.DataFrame(columns = ['n','p','p_mc'])
 ## start for
 for i in range(1,10001):
     groupname = 'group' + str(i)
     genes = groups.loc[groups.GroupName==groupname,'#Node']
     n = sum(linked_genes.count(x) for x in genes)
     p = p_vals[n]
-    results.loc[groupname] = [n,p]
+    p_mc = p_vals_mc[n]
+    results.loc[groupname] = [n,p, p_mc]
 
 results.to_csv('example-random/genesetDP_results.tsv', sep = '\t')

### 2018-05-04 exp/combine_plot.py
monte carlo results and plot

@@ -20,6 +20,7 @@ binox = binox['#4:p.value']
 
 genesetdp = pd.read_csv('example-random/genesetDP_results.tsv', sep='\t', index_col=0)
 genesetdp_sorted = genesetdp['p'].sort_values()
+genesetdp_sorted_mc = genesetdp['p_mc'].sort_values()
 genesetdp_ix = (np.arange(len(genesetdp_sorted.index))+0.5)/float(len(genesetdp_sorted))
 
 binox = binox.loc[genesetdp.index]
@@ -37,11 +38,12 @@ sns.despine()
 
 plt.loglog(binox_ix, binox_sorted, '.')
 plt.loglog(genesetdp_ix, genesetdp_sorted,'.')
+plt.loglog(genesetdp_ix, genesetdp_sorted_mc,'.')
 plt.loglog(x_y, '--', color = 'r', alpha = 0.5)
 plt.loglog(x_2y, ':',color = 'b', alpha = 0.5)
 plt.loglog(x_halfy, ':',color = 'b', alpha = 0.5)
 
-plt.legend(['BinoX', 'GeneSetDP', 'y=x', 'y=0.5x,y=2x'])
+plt.legend(['BinoX', 'GeneSetDP', 'GeneSetDP, monte carlo', 'y=x', 'y=0.5x,y=2x'])
 plt.ylabel('$p$ value')
 plt.xlabel('Normalized rank')
 

### 2018-05-04 exp/experiment.py
Added tracing printouts of number of genes

@@ -29,8 +29,7 @@ pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.v
 k = genk.generate_k(pathway_genes)
 
 N = genesetdp.genesetdp(k,Q)
-#assert(float(sum(N))==float(genesetdp.ncr(sum(k),Q)))
-assert(float(sum(N))==float(comb(sum(k),Q,exact=False)))
+#assert(float(sum(N))==float(comb(sum(k),Q,exact=False)))
 
 p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
 
@@ -40,7 +39,6 @@ groups = pd.read_csv('example-random/Random_Signatures.tsv', sep='\t')
 
 network = genk.network
 
-# linked_genes = network.loc[pathway_genes].dropna().gene2.values.tolist()
 linked_genes = network.index[network.gene2.isin(pathway_genes)].tolist()
 
 results = pd.DataFrame(columns = ['n','p','p_mc'])

### 2018-05-04 exp/generate_random.py

@@ -10,6 +10,7 @@ else:
 original = pd.read_csv('~/git/binox/example/Network.tsv', sep='\t')
 
 genes = np.unique(original['#2:Gene1'])
+print("Num genes=",len(genes))
 
 random_signatures = pd.DataFrame()
 

### 2018-05-04 src/generate_k.py

@@ -27,9 +27,10 @@ def generate_k(pathway_genes):
 
     k = k_proto.loc[list(range(max(k_proto.index)+1))]
     k[k.isna()] = 0
-    
+
     zero_link_genes = n_genes - sum(k)
     k = np.array(k[1:], dtype='u4')
     k = np.insert(k,0,zero_link_genes)
+    print("Here number of genes are ", n_genes)
 
     return k

### 2018-05-04 src/genesetdp.py

@@ -32,8 +32,9 @@ if __name__ == "__main__":
     #k = np.array([3,3,3],dtype='u4')
     N=genesetdp(k,q)
     p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
-    x,sk,ix=[],float(sum(k)),0
+    x,sk,ix,i2=[],float(sum(k)),0,0
     for ki in k:
+        y.insert
         for _ in range(ki):
             x.insert(0,(ix+0.5)/(sk))
             ix += 1

### 2018-05-04 src/genesetdp_monte_carlo.py

@@ -5,11 +5,12 @@ from genesetdp import find_maxscore
 def genesetmc(k,q, n_runs):
 
     maxscore = find_maxscore(k,q)
-    N_mc = np.zeros(maxscore+10)
+    N_mc = np.zeros(maxscore+1)
+    gene_links = np.array([ix for ix in range(len(k)) for _ in range(k[ix])])
 
     for i in range(n_runs):
-        sample = sum(np.random.choice(np.arange(len(k)), q, p = np.array(k)/sum(k)))
-        N_mc[sample] += 1
+        sample = sum(np.random.choice(gene_links, q, replace=False))
+        N_mc[sample] += 1.0
 
 
     p_vals = (np.flipud(np.cumsum(np.flipud(N_mc))) - N_mc/2)/sum(N_mc)

### 2018-05-04 exp/generate_random.py
Copied structure from one part of the code to the other.

@@ -8,8 +8,15 @@ else:
     Q=25
 
 original = pd.read_csv('~/git/binox/example/Network.tsv', sep='\t')
+## manipulate network
+network_treshold = 0.7
+network_cut = original['#0:PFC'] >= network_treshold
+network_one = original.loc[network_cut]
+network = network_one.iloc[:,[0,1]].reset_index(drop=True)
+network.columns = ['gene1','gene2']
+network = network.set_index('gene1')
 
-genes = np.unique(original['#2:Gene1'])
+genes = np.unique(network.index)
 print("Num genes=",len(genes))
 
 random_signatures = pd.DataFrame()

### 2018-05-05 exp/generate_random.py
fixed k[0], regenerated randrom groups and results

@@ -9,14 +9,14 @@ else:
 
 original = pd.read_csv('~/git/binox/example/Network.tsv', sep='\t')
 ## manipulate network
-network_treshold = 0.7
-network_cut = original['#0:PFC'] >= network_treshold
-network_one = original.loc[network_cut]
-network = network_one.iloc[:,[0,1]].reset_index(drop=True)
-network.columns = ['gene1','gene2']
-network = network.set_index('gene1')
+# network_treshold = 0.7
+# network_cut = original['#0:PFC'] >= network_treshold
+# network_one = original.loc[network_cut]
+# network = network_one.iloc[:,[0,1]].reset_index(drop=True)
+# network.columns = ['gene1','gene2']
+# network = network.set_index('gene1')
 
-genes = np.unique(network.index)
+genes = np.unique(original['#2:Gene1'])
 print("Num genes=",len(genes))
 
 random_signatures = pd.DataFrame()

### 2018-05-05 src/generate_k.py

@@ -17,7 +17,7 @@ network = network_one.iloc[:,[0,1]].reset_index(drop=True)
 network.columns = ['gene1','gene2']
 network = network.set_index('gene1')
 
-n_genes = len(np.unique(network.index))
+n_genes = len(np.unique(network_proto['#2:Gene1']))
 
 def generate_k(pathway_genes):
     # links = network.loc[pathway_genes].dropna()

### 2018-05-06 doc/genesetdp.tex
Reformulated the text to include genesetmc

@@
 \bibliographystyle{plainurl}
-\title{A dynamic programming algorithm to assess the reliability of inferences from network enrichment analysis}
-\titlerunning{Dynamic Programming for Inferences from Network Enrichment Analysis} %optional, in case that the title is too long; the running title should fit into the top page column
+\title{One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis}
+\titlerunning{Null models for Network Enrichment Analysis} %optional, in case that the title is too long; the running title should fit into the top page column
 
@@
 Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations.
-Here we instead have designed a dynamic programming  algorithm that can calculate the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences.
+Here we instead have designed one dynamic programming  algorithm that calculates the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences.
+We also implemented a random sampling method, carrying out the same task.
 We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX.
@@
 
-Since such methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desirable, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic.
-Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret as the null hypothesis becomes complex.
+As it is difficult to design network randomization techniques that are completely random, such methods often introduce an undesirable bias in the statistics.
+For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic.
+Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
 
-Here, we present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the score distribution of any query of a given size.
-The algorithm enables us to calculate well-defined statistics of inferences from NEA.
-Using simulations we demonstrated that the algorithm produced unbiased statistics.
+Users of NEA typically want to test if there is a fuctional association between a query set of genes and a pathway\cite{alexeyenko2012network}.
+We argue that such test are much easier implemented as tests of the query set, i.e.
+is the query set a randomly selected set of genes, or is there an overrepresentation of well-connected genes in the query set?
 
-\section*{Algorithm}
+Hence, here we employ a null model of the form ``The query genes are randomly selected the genome``.
+We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
+We also implemented one random sampling algoritm for the same task, that we called GeneSetMC.
+The algorithms enabled us to calculate well-defined statistics of inferences from NEA.  Using simulations we demonstrated that both the algorithms produced unbiased statistics.
 
-In network-based gene set analysis one score a query set of genes $ \{g_1 \ldots g_Q\} $ from a genome with genes $\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$.
+\section*{Algorithms}
+
+In network-based gene set analysis one score a query set of genes $ \mathcal{Q}=\{g_1 \ldots g_Q\} $ from a genome with genes $\mathcal{G}=\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$.
 In NEA the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.
@@
 Here, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$.
-This allows us to evaluate our score against the null hypothesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
+This allows us to evaluate our score against the null hypothesis, $H_0$ : ``The Q query genes are randomly selected from the genome''.
 This can be expressed as a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score $s$,
@@
 
-\subsection*{Computation of the score distribution}
-
 We can reformulate Equation \ref{eq:sum_ij} by defining the number-of-links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
@@
 \end{equation}
+Such number-of-links, $l_i$, can be pre-computed for all genes $i, \ldots, G$.
 
-Such number-of-links, $l_i$, can be pre-computed for all genes $i, \ldots, G$.
+\subsection{Random sampling algorithm}
+
+We first implemented a random sampling algorithm to assess $N(s)$.
+We randomly selected sets of $Q$ genes from $\mathcal{G}$.
+These sets scores were calculated with Equation \ref{eq:sum_i}.
+We counted the number of times $F_B(s)$ a score of $s$ was obtained when sampling $B$ gene sets.
+We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Equation~\ref{eq:pval}.
+We refer to this procedure as GeneSetMC.
+
+\subsection*{Computation of the score distribution}
+
 Furthermore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{i}{l_i}$.
@@
 For details see Listing \ref{lst:gensetdp}.
+We refer to this method as GeneSetDP.
 
@@
 
+\subsection{Availability}
+
+Source code of an implementation of GeneSetDP and GeneSetMC  as well as to the experiments conducte here in, is available from \url{https://github.com/statisticalbiotechnology/genesetdp}.
+
 \section*{Results}
 
-We implemented a Python program that reads network and pathway definition files and scored query sets against a pathway according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
+We implemented a Python program that reads network and pathway definition files and scored query sets against a pathway according to Equation \ref{eq:sum_i}, using GeneSetDP and GeneSetMC described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
 We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).
 
-To illustrate the $p$~value calculation procedure we plotted the score distribution N(s) for a query size of 25 genes in Figure \ref{fig:score_dist}.
+To illustrate the $p$~value calculation procedure we plotted the score distributions N(s) for a query size of 10, 20 and 30 genes in Figure \ref{fig:score_dist}, using GeneSetDP.
 
@@
 		\begin{center}
-				\includegraphics[scale=0.8]{figures/score_distribuition.png}
+    \begin{array}[ccc]
+      \includegraphics[with=4.5cm]{figures/score_distribuition.png &
+      \includegraphics[with=4.5cm]{figures/score_distribuition.png} &
+      \includegraphics[with=4.5cm]{figures/score_distribuition.png} \\
+      $Q=10$ & $Q=20$ & $Q=30$
+    \end{array}
 		\end{center}
-%FIXME: Do not use figure numbers in their file name, this creates havoc if you want to reorganize them later
-  \caption{{\bf Score distribution of a query size of 25 genes against the ``Glycolysis/Gluconeogenesis'' pathway.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
+\caption{{\bf Score distribution of a query size of 10, 20 and 30 genes against the ``Glycolysis/Gluconeogenesis'' pathway.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
 how many ways one can reach a certain score when adding up the scores for $Q$ genes.}
   \label{fig:score_dist}
+FIXME: update these figures.
 \end{figure}
@@
 As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed.
-To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration} for $10000$ randomly assembeled queries of $25$ genes from the human genome.
+To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration} for $10000$ randomly assembeled queries from the human genome.
 As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.
@@
   \begin{center}
-				\includegraphics[scale=0.8]{figures/calibration.png}
-		\end{center}
-
-%\includegraphics{img/FIXME.png}
-  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets from the investigated genome definition and calculated their $p$ values with GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
-To test the $p$~values uniformity we plotted them against their Normalized rank.
+  \begin{array}[ccc]
+    \includegraphics[with=4.5cm]{figures/calibration.png &
+    \includegraphics[with=4.5cm]{figures/calibration.png} &
+    \includegraphics[with=4.5cm]{figures/calibration.png} \\
+    $Q=10$ & $Q=20$ & $Q=30$
+  \end{array}
+  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
+To test the $p$~values uniformity we plotted them against their normalized rank.
 We also added the calibration curve of BinoX, evaluated on the same random sets.}
@@
 
-Here we have implemented a method, GeneSetDP, to calculate unbiased $p$~values for inferences from NEA.
+Here we have implemented two methods, GeneSetDP and GeneSetMC, to calculate unbiased $p$~values for inferences from NEA.
 We demonstrated that the method gave more accurate statistics than the popular BinoX method using random subsets of a genome.
 
-It should be noted that our outlined method a could work well with Monte Carlo-sampling, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.
-
-Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''.
+Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genome''.
 Previous implementations of NEA in the literature otherwise use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.

### 2018-05-06 doc/genesetdp.tex
Now compiling.

@@
 
-To illustrate the $p$~value calculation procedure we plotted the score distributions N(s) for a query size of 10, 20 and 30 genes in Figure \ref{fig:score_dist}, using GeneSetDP.
+To illustrate the $p$~value calculation procedure, we plotted the score distributions $N(s)$ for a query size of 10, 20 and 30 genes in Figure \ref{fig:score_dist}, using GeneSetDP.
 
@@
 	\begin{center}
-    \begin{array}[ccc]
-      \includegraphics[with=4.5cm]{figures/score_distribuition.png &
-      \includegraphics[with=4.5cm]{figures/score_distribuition.png} &
-      \includegraphics[with=4.5cm]{figures/score_distribuition.png} \\
-      $Q=10$ & $Q=20$ & $Q=30$
-    \end{array}
+    \begin{tabular}{c@{}c@{}c}
+      \includegraphics[width=5cm]{figures/score_distribuition.png} &
+      \includegraphics[width=5cm]{figures/score_distribuition.png} &
+      \includegraphics[width=5cm]{figures/score_distribuition.png} \\
+      $Q=10$ & $Q=20$ & $Q=30$ \\
+    \end{tabular}
 	\end{center}
+  {\em FIXME: update these figures.}
   \caption{{\bf Score distribution of a query size of 10, 20 and 30 genes against the ``Glycolysis/Gluconeogenesis'' pathway.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
@@
   \label{fig:score_dist}
-FIXME: update these figures.
 \end{figure}
@@
   \begin{center}
-  \begin{array}[ccc]
-    \includegraphics[with=4.5cm]{figures/calibration.png &
-    \includegraphics[with=4.5cm]{figures/calibration.png} &
-    \includegraphics[with=4.5cm]{figures/calibration.png} \\
+    \begin{tabular}{c@{}c@{}c}
+      \includegraphics[width=5cm]{figures/calibration.png} &
+      \includegraphics[width=5cm]{figures/calibration.png} &
+      \includegraphics[width=5cm]{figures/calibration.png} \\
       $Q=10$ & $Q=20$ & $Q=30$
-  \end{array}
+    \end{tabular}
+  \end{center}
   \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.

### 2018-05-06 doc/genesetdp.tex
copyright

@@
 \authorrunning{G. S. Jeuken and L. K\"{a}ll}
+
+\Copyright{Gustavo S. Jeuken and Lukas K\"{a}ll}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+\subjclass{G.3.
+Probabilistic algorithms (including Monte Carlo)}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html .
+E.g., cite as "F.1.1 Models of Computation".
+\keywords{Gene Set Enrichment Analysis; Null model; p value; Sampling}% mandatory: Please provide 1-5 keywords
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\EventEditors{John Q. Open and Joan R. Acces}
+%\EventNoEds{2}
+\EventLongTitle{18th Workshop on Algorithms in Bioinformatics (WABI 2018)}
+\EventShortTitle{WABI 2018}
+\EventAcronym{WABI}
+\EventYear{2018}
+\EventDate{20–22 August 2018}
+\EventLocation{Helsinki, Finland}
+%\EventLogo{}
+%\SeriesVolume{42}
+%\ArticleNo{23}
+% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
@@
 
-Users of NEA typically want to test if there is a fuctional association between a query set of genes and a pathway\cite{alexeyenko2012network}.
+Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}.
 We argue that such test are much easier implemented as tests of the query set, i.e.
@@
 
-\subsection{Random sampling algorithm}
+\subsection*{Random sampling algorithm}
 
@@
 The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input.
-The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt ncr(a,b)}, which calculates $ a \choose b $.
+The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt comb(a,b)}, which calculates $ a \choose b $.
 }
@@
     max_s = find_maxscore(k,Q)
-    N = np.zeros((max_s+1,Q+1),dtype='float_')
+    N = np.zeros((max_s+1,Q+1))
     N[0,0] = 1
@@
                     if c-b>=0 and s-a*b>=0:
-                        N[s,c] += ncr(k[a], b) * N[s-a*b,c-b]
+                        N[s,c] += comb(k[a], b) * N[s-a*b,c-b]
     return N[:,q]
@@
 
-As a representative example for we chose the ``Glycolysis/Gluconeogenesis'' pathway, as it contained a number of genes that coincided with the median number of genes of the pathways in the definition file.
+For the purpose of demonstrating our argument, we selected a representative example in the ``Glycolysis/Gluconeogenesis'' pathway, as it contained a number of genes that coincided with the median number of genes of the pathways in the definition file.
 
-\subsection{Availability}
+\subsection*{Availability}
 
-Source code of an implementation of GeneSetDP and GeneSetMC  as well as to the experiments conducte here in, is available from \url{https://github.com/statisticalbiotechnology/genesetdp}.
+Source code of an implementation of GeneSetDP and GeneSetMC, as well as code for generating the plots of this paper, is available from \url{https://github.com/statisticalbiotechnology/genesetdp}.
 
@@
 As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed.
-To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration} for $10000$ randomly assembeled queries from the human genome.
+To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration} for $10000$ randomly assembled queries from the human genome.
 As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.
@@
   \end{center}
-  \caption{{\bf Calibration of GeneSetDP.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
+  \caption{{\bf Calibration of GeneSetDP and GeneSetMC.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
 To test the $p$~values uniformity we plotted them against their normalized rank.

### 2018-05-06 doc/genesetdp.tex
Edited discussion, and added some references.

@@
 As it is difficult to design network randomization techniques that are completely random, such methods often introduce an undesirable bias in the statistics.
-For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic.
+For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
 Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
@@
 We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
-We also implemented one random sampling algoritm for the same task, that we called GeneSetMC.
+We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
 The algorithms enabled us to calculate well-defined statistics of inferences from NEA.  Using simulations we demonstrated that both the algorithms produced unbiased statistics.
@@
 the calculated $p$ values are larger than expected.
-Meanwhile, BinoX appears strongly anti-conservative, that is, the $p$ values are much lower than expected.
+Meanwhile, BinoX appears strongly anti-conservative, that is, the $p$ values are lower than expected.
 
@@
 Here we have implemented two methods, GeneSetDP and GeneSetMC, to calculate unbiased $p$~values for inferences from NEA.
-We demonstrated that the method gave more accurate statistics than the popular BinoX method using random subsets of a genome.
+Instead of testing our methods performance in terms of sensitivity, we here instead chose to test the method in terms of the methods statistical accuracy.
+We argue that more studies should make a point at demonstrate that their methods are well calibrated, as it is a prerequisite for measuring performance at least when ground truth of biological effect is not known.
+With this in mind, we demonstrated that our methods gave a better calibration than the popular BinoX method, by random sampling of subsets of a genome.
 

### 2018-05-07 doc/genesetdp.tex
edited key words.
specification of null.

@@
 E.g., cite as "F.1.1 Models of Computation".
-\keywords{Gene Set Enrichment Analysis; Null model; p value; Sampling}% mandatory: Please provide 1-5 keywords
+\keywords{Gene Set Analysis; Calibration; Null model; p value; Sampling}% mandatory: Please provide 1-5 keywords
 
@@
 
-Hence, here we employ a null model of the form ``The query genes are randomly selected the genome``.
+Hence, here we employ a different null model, more directly aimed at model lack of association between query and pathway gene sets, which we formulate as ``The query genes are randomly selected the genome''.
 We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.

### 2018-05-07 exp/figure_calibration.py
some results for Q10,20,30

@@ -1,37 +1,0 @@
-import pandas as pd
-import numpy as np
-
-import sys
-sys.path.insert(0, '../src')
-import generate_k as genk
-import genesetdp
-
-## example path
-path = "~/git/binox/example/"
-
-## read original patway information file
-pathways_file = 'Pathways.tsv'
-pathways = pd.read_csv(path + pathways_file, sep='\t', header=None)
-pathways.columns = ['gene','pathway']
-
-pathway_name = 'GLYCOLYSIS_/_GLUCONEOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
-
-pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.values
-
-k = genk.generate_k(pathway_genes)
-
-N = genesetdp.genesetdp(k,25)
-
-####
-
-import matplotlib.pyplot as plt
-import seaborn as sns
-
-sns.set_style("ticks")
-sns.despine()
-
-plt.step(np.arange(len(N)), N, where='mid')
-plt.ylabel('$N(s)$')
-plt.xlabel('Score, $s$')
-
-plt.savefig('score_distribuition.png')

### 2018-05-07 exp/figure_distribuition.py

@@ -1,0 +1,39 @@
+import pandas as pd
+import numpy as np
+
+import sys
+sys.path.insert(0, '../src')
+import generate_k as genk
+import genesetdp
+
+## example path
+path = "~/git/binox/example/"
+
+## read original patway information file
+pathways_file = 'Pathways.tsv'
+pathways = pd.read_csv(path + pathways_file, sep='\t', header=None)
+pathways.columns = ['gene','pathway']
+
+pathway_name = 'GLYCOLYSIS_/_GLUCONEOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
+
+pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.values
+
+k = genk.generate_k(pathway_genes)
+
+N = genesetdp.genesetdp(k,10)
+
+####
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+sns.set_style("ticks")
+sns.despine()
+
+plt.step(np.arange(len(N)), N, where='mid')
+plt.ylabel('$N(s)$')
+plt.xlabel('Score, $s$')
+plt.title('Q = 10')
+
+plt.savefig('score_distribuition_Q10.png')
+plt.show()

### 2018-05-07 doc/genesetdp.tex
some results for Q10,20,30

@@
     \begin{tabular}{c@{}c@{}c}
-      \includegraphics[width=5cm]{figures/score_distribuition.png} &
-      \includegraphics[width=5cm]{figures/score_distribuition.png} &
-      \includegraphics[width=5cm]{figures/score_distribuition.png} \\
+      \includegraphics[width=5cm]{figures/score_distribuition_Q10.png} &
+      \includegraphics[width=5cm]{figures/score_distribuition_Q20.png} &
+      \includegraphics[width=5cm]{figures/score_distribuition_Q30.png} \\
       $Q=10$ & $Q=20$ & $Q=30$ \\

### 2018-05-07 exp/figure_distribuition.py

@@ -20,7 +20,7 @@ pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.v
 
 k = genk.generate_k(pathway_genes)
 
-N = genesetdp.genesetdp(k,10)
+N = genesetdp.genesetdp(k,30)
 
 ####
 
@@ -33,7 +33,6 @@ sns.despine()
 plt.step(np.arange(len(N)), N, where='mid')
 plt.ylabel('$N(s)$')
 plt.xlabel('Score, $s$')
-plt.title('Q = 10')
 
-plt.savefig('score_distribuition_Q10.png')
+plt.savefig('score_distribuition_Q30.png')
 plt.show()

### 2018-05-07 exp/combine_plot.py
new calibration plots

@@ -3,30 +3,36 @@ import numpy as np
 import matplotlib.pyplot as plt
 import seaborn as sns
 
+Q = 30
 
-binox = pd.read_csv('example-random/BinoX_results.tsv', sep='\t')
+binox = pd.read_csv('example-random/BinoX_results_Q' + str(Q) + '.tsv', sep='\t')
 
 ###### only enriched
-binox.loc[binox['#6:relationType'] == '-', '#4:p.value'] = 1 - binox.loc[binox['#6:relationType'] == '-', '#4:p.value']
+binox.loc[:, 'enrichment'] = binox.loc[:, '#4:p.value']
+binox.loc[binox['#6:relationType'] == '-', 'enrichment'] = 1 - binox.loc[binox['#6:relationType'] == '-', '#4:p.value']
 ######
-
-
 
 binox = binox.set_index('#3:NameGroupB')
 
 binox.index = [x.lower() for x in binox.index]
 
-binox = binox['#4:p.value']
+binox_both = binox['#4:p.value']
+binox_enrich = binox['enrichment']
 
-genesetdp = pd.read_csv('example-random/genesetDP_results.tsv', sep='\t', index_col=0)
+genesetdp = pd.read_csv('example-random/genesetDP_results_Q' + str(Q) + '.tsv', sep='\t', index_col=0)
 genesetdp_sorted = genesetdp['p'].sort_values()
 genesetdp_sorted_mc = genesetdp['p_mc'].sort_values()
 genesetdp_ix = (np.arange(len(genesetdp_sorted.index))+0.5)/float(len(genesetdp_sorted))
 
-binox = binox.loc[genesetdp.index]
-binox.loc[np.isnan(binox)] = 1
-binox_sorted = binox.sort_values()
-binox_ix = (np.arange(len(binox_sorted.index))+0.5)/float(len(binox_sorted.index))
+binox_both = binox_both.loc[genesetdp.index]
+binox_enrich = binox_enrich.loc[genesetdp.index]
+binox_both.loc[np.isnan(binox_both)] = 1
+binox_enrich.loc[np.isnan(binox_enrich)] = 1
+
+binox_both_sorted = binox_both.sort_values()
+binox_enrich_sorted = binox_enrich.sort_values()
+
[…]

### 2018-05-07 exp/combine_plot_multiple.py

@@ -1,0 +1,77 @@
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+def one_plot(ax, Q):
+
+    binox = pd.read_csv('example-random/BinoX_results_Q' + str(Q) + '.tsv', sep='\t')
+
+    ###### only enriched
+    binox.loc[:, 'enrichment'] = binox.loc[:, '#4:p.value']
+    binox.loc[binox['#6:relationType'] == '-', 'enrichment'] = 1 - binox.loc[binox['#6:relationType'] == '-', '#4:p.value']
+    ######
+
+    binox = binox.set_index('#3:NameGroupB')
+
+    binox.index = [x.lower() for x in binox.index]
+
+    binox_both = binox['#4:p.value']
+    binox_enrich = binox['enrichment']
+
+    genesetdp = pd.read_csv('example-random/genesetDP_results_Q' + str(Q) + '.tsv', sep='\t', index_col=0)
+    genesetdp_sorted = genesetdp['p'].sort_values()
+    genesetdp_sorted_mc = genesetdp['p_mc'].sort_values()
+    genesetdp_ix = (np.arange(len(genesetdp_sorted.index))+0.5)/float(len(genesetdp_sorted))
+
+    binox_both = binox_both.loc[genesetdp.index]
+    binox_enrich = binox_enrich.loc[genesetdp.index]
+    binox_both.loc[np.isnan(binox_both)] = 1
+    binox_enrich.loc[np.isnan(binox_enrich)] = 1
+
+    binox_both_sorted = binox_both.sort_values()
+    binox_enrich_sorted = binox_enrich.sort_values()
+
+    binox_ix = (np.arange(len(binox_both_sorted.index))+0.5)/float(len(binox_both_sorted.index))
+
+    x_y = pd.DataFrame(np.arange(0,1,0.001), index = np.arange(0,1,0.001))
+    x_2y = pd.DataFrame(np.arange(0,1,0.001)*2, index = np.arange(0,1,0.001))
+    x_halfy = pd.DataFrame(np.arange(0,0.5,0.001), index = np.arange(0,0.5,0.001)*2)
+
+    sns.set_style("ticks")
+    sns.despine()
+
+    ax.loglog(binox_ix, binox_both_sorted, '.')
[…]

### 2018-05-07 doc/genesetdp.tex
more BinoX results and plots
new calibration plots
edit, and adition of citation

--- 2018-05-07 doc/genesetdp.tex
@@
 While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases.
-Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string}.
+Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}.
 Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}.

### 2018-05-07 doc/genesetdp.tex

--- 2018-05-07 doc/genesetdp.tex
@@
 	\end{center}
-  {\em FIXME: update these figures.}
   \caption{{\bf Score distribution of a query size of 10, 20 and 30 genes against the ``Glycolysis/Gluconeogenesis'' pathway.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
@@
   \begin{center}
-    \begin{tabular}{c@{}c@{}c}
-      \includegraphics[width=5cm]{figures/calibration.png} &
-      \includegraphics[width=5cm]{figures/calibration.png} &
-      \includegraphics[width=5cm]{figures/calibration.png} \\
-      $Q=10$ & $Q=20$ & $Q=30$
-    \end{tabular}
+      \includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png}
   \end{center}

### 2018-05-07 exp/figure_distribuition.py
text

@@ -31,8 +31,8 @@ sns.set_style("ticks")
 sns.despine()
 
 plt.step(np.arange(len(N)), N, where='mid')
-plt.ylabel('$N(s)$')
-plt.xlabel('Score, $s$')
+plt.ylabel('$N(s)$', fontsize=14)
+plt.xlabel('Score, $s$', fontsize=14)
 
 plt.savefig('score_distribuition_Q30.png')
 plt.show()

### 2018-05-07 doc/genesetdp.tex
text

@@
 
-We implemented a Python program that reads network and pathway definition files and scored query sets against a pathway according to Equation \ref{eq:sum_i}, using GeneSetDP and GeneSetMC described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
+We implemented a Python program that reads network and pathway definition files and scores a query sets against a pathway according to Equation \ref{eq:sum_i}, using GeneSetDP and GeneSetMC described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
 We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).
@@
       \includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png}
+	  % \includegraphics{figures/calibration_legend.png}
   \end{center}
@@
 To test the $p$~values uniformity we plotted them against their normalized rank.
-We also added the calibration curve of BinoX, evaluated on the same random sets.}
+We also added the calibration curve of BinoX, evaluated on the same random sets.
+On all plots the dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$, for comparison}
   \label{fig:calibration}

### 2018-05-07 doc/genesetdp.tex
description of normalized rank.

--- 2018-05-07 doc/genesetdp.tex
@@
       \includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png}
+	  % \includegraphics{figures/calibration_legend.png}
   \end{center}
   \caption{{\bf Calibration of GeneSetDP and GeneSetMC.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
-To test the $p$~values uniformity we plotted them against their normalized rank.
-We also added the calibration curve of BinoX, evaluated on the same random sets.}
+To test the $p$~values uniformity we plotted them against their normalized rank, i.e.
+each $p$~value's (<rank>-0.5)/<total number of $p$~values>).
+We also added the calibration curve of BinoX, evaluated on the same random sets.
+The dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$}
   \label{fig:calibration}

### 2018-05-07 doc/genesetdp.tex
Typo
removed  word

--- 2018-05-07 doc/genesetdp.tex
@@
   \caption{{\bf Calibration of GeneSetDP and GeneSetMC.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
-To test the $p$~values uniformity we plotted them against their normalized rank.
+To test the $p$~values uniformity we plotted them against their {\em normalized rank}, i.e.
+each $p$~value's (<rank>-0.5)/<total number of $p$~values>.
 We also added the calibration curve of BinoX, evaluated on the same random sets.
@@
 Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genome''.
-Previous implementations of NEA in the literature otherwise use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.
+Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.
 In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics.

### 2018-05-08 doc/genesetdp.tex
Added discussion on depletion.

@@
 
+Both our implementations, GeneSetDP and GeneSetMC, are intended for measuring enrichment of links between the query and the pathway set of genes.
+Most other NEA methods also report depletion of such links.
+One could easily modify our code to instead of only measuring the higher scoring tail measuring both tails of $N(s)$.
+However, this might result in a drop in sensitivity, so we have so far no implemented such a mechanism.
+
 Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genome''.

### 2018-05-08 exp/combine_plot.py
text

@@ -50,8 +50,11 @@ plt.loglog(x_y, '--', color = 'r', alpha = 0.5)
 plt.loglog(x_2y, ':',color = 'b', alpha = 0.5)
 plt.loglog(x_halfy, ':',color = 'b', alpha = 0.5)
 
+plt.legend(['BinoX', 'BinoX (enrichment)' 'GeneSetDP', 'GeneSetMC', 'y=x', 'y=0.5x,y=2x'])
 plt.ylabel('$p$ value')
 plt.xlabel('Normalized rank')
+
+
 
 plt.savefig('calibration_Q' + str(Q) + '.png')
 plt.show()

### 2018-05-08 exp/combine_plot_multiple.py

@@ -41,10 +41,11 @@ def one_plot(ax, Q):
     sns.set_style("ticks")
     sns.despine()
 
-    ax.loglog(binox_ix, binox_both_sorted, '.')
-    ax.loglog(binox_ix, binox_enrich_sorted, '.')
     ax.loglog(genesetdp_ix, genesetdp_sorted,'.')
     ax.loglog(genesetdp_ix, genesetdp_sorted_mc,'.')
+    ax.loglog(binox_ix, binox_both_sorted, '.')
+    ax.loglog(binox_ix, binox_enrich_sorted, '.')
+
     ax.loglog(x_y, '--', color = 'r', alpha = 0.5)
     ax.loglog(x_2y, ':',color = 'b', alpha = 0.5)
     ax.loglog(x_halfy, ':',color = 'b', alpha = 0.5)
@@ -56,22 +57,27 @@ one_plot(ax1, 10)
 one_plot(ax2, 20)
 one_plot(ax3, 30)
 
-ax1.set_title('Q = 10')
-ax2.set_title('Q = 30')
-ax3.set_title('Q = 30')
+ax1.set_title('Q = 10', fontsize=18)
+ax2.set_title('Q = 30', fontsize=18)
+ax3.set_title('Q = 30', fontsize=18)
 
-ax1.set(ylabel='$p$ value', xlabel = 'Normalized rank')
-ax2.set(xlabel = 'Normalized rank')
-ax3.set(xlabel = 'Normalized rank')
+ax1.set_ylabel('$p$ value', fontsize = 16)
+ax1.set_xlabel('Normalized rank', fontsize = 16)
+ax2.set_xlabel('Normalized rank', fontsize = 16)
+ax3.set_xlabel('Normalized rank', fontsize = 16)
 
-# for ax in [ax1,ax2,ax3]:
-#     ax.set_aspect('equal')
+for ax in [ax1,ax2,ax3]:
+    ax.xaxis.set_tick_params(labelsize=12)
+    ax.yaxis.set_tick_params(labelsize=12)
 
-lgd = fig.legend(['BinoX', 'BinoX (enriched)     ', 'GeneSetDP', 'GeneSetMC', 'y=x', 'y=0.5x,y=2x'], loc='center left', bbox_to_anchor=(0.85,0.5))
+# lgd = fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX', 'BinoX (enriched)'], loc='center left', bbox_to_anchor=(0.85,0.5))
[…]

### 2018-05-08 doc/genesetdp.tex
text
new plot
legend
Edited null

@@
 
-Hence, here we employ a different null model, more directly aimed at model lack of association between query and pathway gene sets, which we formulate as ``The query genes are randomly selected the genome''.
+Hence, here we employ a different null model, more directly aimed at model lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''.
 We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
@@
 Here, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$.
-This allows us to evaluate our score against the null hypothesis, $H_0$ : ``The Q query genes are randomly selected from the genome''.
+This allows us to evaluate our score against the null hypothesis, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''.
+This translates to evaluate yje distribution of scores that would occure if the $Q$ query genes are randomly selected from the genome''.
 This can be expressed as a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score $s$,
@@
 
-Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genome''.
+Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''.
 Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.

### 2018-05-08 doc/genesetdp.tex
plots

--- 2018-05-08 doc/genesetdp.tex
@@
 	\begin{center}
-    \begin{tabular}{c@{}c@{}c}
-      \includegraphics[width=5cm]{figures/score_distribuition_Q10.png} &
-      \includegraphics[width=5cm]{figures/score_distribuition_Q20.png} &
-      \includegraphics[width=5cm]{figures/score_distribuition_Q30.png} \\
-      $Q=10$ & $Q=20$ & $Q=30$ \\
-    \end{tabular}
+		\includegraphics[width=1.0\textwidth]{figures/score_distribuition_multiple.png} 
     \end{center}
@@
   \begin{center}
-      \includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png}
-	  % \includegraphics{figures/calibration_legend.png}
+		\begin{tabular}[t]{c@{}c}
+				\includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png} &
+				\includegraphics[width=0.2\textwidth]{figures/calibration_legend.png}
+		\end{tabular}
   \end{center}

### 2018-05-08 exp/figure_distribuition.py

@@ -20,19 +20,37 @@ pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.v
 
 k = genk.generate_k(pathway_genes)
 
-N = genesetdp.genesetdp(k,30)
+N10 = genesetdp.genesetdp(k,10)
+N20 = genesetdp.genesetdp(k,20)
+N30 = genesetdp.genesetdp(k,30)
 
 ####
 
 import matplotlib.pyplot as plt
 import seaborn as sns
 
+fig, (ax1, ax2, ax3) = plt.subplots(3, 1, sharex=True)
+
 sns.set_style("ticks")
 sns.despine()
 
-plt.step(np.arange(len(N)), N, where='mid')
-plt.ylabel('$N(s)$', fontsize=14)
-plt.xlabel('Score, $s$', fontsize=14)
+ax1.set_ylabel('Q = 10\n$N(s)$', fontsize=16)
+ax2.set_ylabel('Q = 20\n$N(s)$', fontsize=16)
+ax3.set_ylabel('Q = 30\n$N(s)$', fontsize=16)
+ax3.set_xlabel('Score, $s$', fontsize=16)
 
-plt.savefig('score_distribuition_Q30.png')
+for ax in [ax1,ax2,ax3]:
+    ax.xaxis.set_tick_params(labelsize=12)
+    ax.yaxis.set_tick_params(labelsize=12)
+
+
+ax1.step(np.arange(len(N10)), N10, where='mid')
+ax2.step(np.arange(len(N20)), N20, where='mid')
+ax3.step(np.arange(len(N30)), N30, where='mid')
+
+plt.tight_layout()
+plt.savefig('score_distribuition_multiple.png')
 plt.show()
+
+
+

### 2018-05-08 exp/experiment.py
reorg

@@ -7,7 +7,7 @@ sys.path.insert(0, '../src')
 import generate_k as genk
 import genesetdp
 
-import genesetdp_monte_carlo as mc
+import genesetmc as mc
 
 if len(sys.argv)>1:
     Q=int(sys.argv[1])

### 2018-05-08 src/generate_k.py
reorg

@@ -10,16 +10,17 @@ network_file = 'Network.tsv'
 network_proto = pd.read_csv(path + network_file, sep='\t')
 
 ## manipulate network
-network_treshold = 0.7
-network_cut = network_proto['#0:PFC'] >= network_treshold
-network_one = network_proto.loc[network_cut]
-network = network_one.iloc[:,[0,1]].reset_index(drop=True)
-network.columns = ['gene1','gene2']
-network = network.set_index('gene1')
 
 n_genes = len(np.unique(network_proto['#2:Gene1']))
 
-def generate_k(pathway_genes):
+def generate_k(pathway_genes, network_treshold = 0.7):
+    network_cut = network_proto['#0:PFC'] >= network_treshold
+    network_one = network_proto.loc[network_cut]
+    network = network_one.iloc[:,[0,1]].reset_index(drop=True)
+    network.columns = ['gene1','gene2']
+    network = network.set_index('gene1')
+
+
     # links = network.loc[pathway_genes].dropna()
     links = network.index[network.gene2.isin(pathway_genes)]
 

### 2018-05-08 src/genesetdp.py

@@ -25,6 +25,10 @@ def genesetdp(k,q):
                         N[s,c] += comb(k[a], b, exact=False) * N[s-a*b,c-b]
     return N[:,q]
 
+def calculate_p(N):
+    p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
+    return p_vals
+
 if __name__ == "__main__":
     # Query size q, number of links k[a]
     q = 2

### 2018-05-08 exp/combine_plot_multiple.py
plot

@@ -43,7 +43,7 @@ def one_plot(ax, Q):
 
     ax.loglog(genesetdp_ix, genesetdp_sorted,'.')
     ax.loglog(genesetdp_ix, genesetdp_sorted_mc,'.')
-    ax.loglog(binox_ix, binox_both_sorted, '.')
+    # ax.loglog(binox_ix, binox_both_sorted, '.')
     ax.loglog(binox_ix, binox_enrich_sorted, '.')
 
     ax.loglog(x_y, '--', color = 'r', alpha = 0.5)
@@ -70,14 +70,14 @@ for ax in [ax1,ax2,ax3]:
     ax.xaxis.set_tick_params(labelsize=12)
     ax.yaxis.set_tick_params(labelsize=12)
 
-# lgd = fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX', 'BinoX (enriched)'], loc='center left', bbox_to_anchor=(0.85,0.5))
-# fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX', 'BinoX (enriched)'], fontsize=15)
+lgd = fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX'], loc='center', ncol =3, bbox_to_anchor=(0.5,0), fontsize=16)
+# fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX'], fontsize=15, ncol=3)
 
 
 fig.set_size_inches(15,4)
-# fig.savefig('calibration_multiple.png', bbox_extra_artists=(lgd,), bbox_inches='tight')
-fig.savefig('calibration_multiple.png', bbox_inches='tight')
-plt.show()
+fig.savefig('calibration_multiple.png', bbox_extra_artists=(lgd,), bbox_inches='tight')
+# fig.savefig('calibration_multiple.png', bbox_inches='tight')
+# plt.show()
 
 # plt.savefig('calibration_multiple.png')
 # plt.show()

### 2018-05-08 exp/figure_distribuition.py

@@ -34,10 +34,14 @@ fig, (ax1, ax2, ax3) = plt.subplots(3, 1, sharex=True)
 sns.set_style("ticks")
 sns.despine()
 
-ax1.set_ylabel('Q = 10\n$N(s)$', fontsize=16)
-ax2.set_ylabel('Q = 20\n$N(s)$', fontsize=16)
-ax3.set_ylabel('Q = 30\n$N(s)$', fontsize=16)
+ax1.set_ylabel('$N(s)$', fontsize=16)
+ax2.set_ylabel('$N(s)$', fontsize=16)
+ax3.set_ylabel('$N(s)$', fontsize=16)
 ax3.set_xlabel('Score, $s$', fontsize=16)
+
+ax1.set_title('                       $Q=10$', fontsize=14, horizontalalignment='left')
+ax2.set_title('                       $Q=20$', fontsize=14, horizontalalignment='left')
+ax3.set_title('                       $Q=30$', fontsize=14, horizontalalignment='left')
 
 for ax in [ax1,ax2,ax3]:
     ax.xaxis.set_tick_params(labelsize=12)
@@ -45,11 +49,14 @@ for ax in [ax1,ax2,ax3]:
 
 
 ax1.step(np.arange(len(N10)), N10, where='mid')
+ax1.plot(len(N10),1, '.', color='r', markersize=12)
 ax2.step(np.arange(len(N20)), N20, where='mid')
+ax2.plot(len(N20),1, '.', color='r', markersize=12)
 ax3.step(np.arange(len(N30)), N30, where='mid')
+ax3.plot(len(N30),1, '.', color='r', markersize=12)
 
 plt.tight_layout()
-plt.savefig('score_distribuition_multiple.png')
+# plt.savefig('score_distribuition_multiple.png')
 plt.show()
 
 

### 2018-05-08 exp/figure_distribuition.py
plot

@@ -39,9 +39,9 @@ ax2.set_ylabel('$N(s)$', fontsize=16)
 ax3.set_ylabel('$N(s)$', fontsize=16)
 ax3.set_xlabel('Score, $s$', fontsize=16)
 
-ax1.set_title('                       $Q=10$', fontsize=14, horizontalalignment='left')
-ax2.set_title('                       $Q=20$', fontsize=14, horizontalalignment='left')
-ax3.set_title('                       $Q=30$', fontsize=14, horizontalalignment='left')
+ax1.set_title('                                 $Q=10$', fontsize=14, horizontalalignment='left')
+ax2.set_title('                                 $Q=20$', fontsize=14, horizontalalignment='left')
+ax3.set_title('                                 $Q=30$', fontsize=14, horizontalalignment='left')
 
 for ax in [ax1,ax2,ax3]:
     ax.xaxis.set_tick_params(labelsize=12)

### 2018-05-08 exp/combine_plot_multiple.py
plot fix

@@ -58,7 +58,7 @@ one_plot(ax2, 20)
 one_plot(ax3, 30)
 
 ax1.set_title('Q = 10', fontsize=18)
-ax2.set_title('Q = 30', fontsize=18)
+ax2.set_title('Q = 20', fontsize=18)
 ax3.set_title('Q = 30', fontsize=18)
 
 ax1.set_ylabel('$p$ value', fontsize = 16)
@@ -70,13 +70,13 @@ for ax in [ax1,ax2,ax3]:
     ax.xaxis.set_tick_params(labelsize=12)
     ax.yaxis.set_tick_params(labelsize=12)
 
-lgd = fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX'], loc='center', ncol =3, bbox_to_anchor=(0.5,0), fontsize=16)
+# lgd = fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX'], loc='center', ncol =3, bbox_to_anchor=(0.5,0), fontsize=16)
 # fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX'], fontsize=15, ncol=3)
 
 
 fig.set_size_inches(15,4)
-fig.savefig('calibration_multiple.png', bbox_extra_artists=(lgd,), bbox_inches='tight')
-# fig.savefig('calibration_multiple.png', bbox_inches='tight')
+# fig.savefig('calibration_multiple.png', bbox_extra_artists=(lgd,), bbox_inches='tight')
+fig.savefig('calibration_multiple.png', bbox_inches='tight')
 # plt.show()
 
 # plt.savefig('calibration_multiple.png')

### 2018-05-08 doc/genesetdp.tex
legend on bottom
plot
edits

@@
 We argue that such test are much easier implemented as tests of the query set, i.e.
-is the query set a randomly selected set of genes, or is there an overrepresentation of well-connected genes in the query set?
+is the query set a randomly selected set of genes or is there an overrepresentation of well-connected genes in the query set?
 
@@
 
-In network-based gene set analysis one score a query set of genes $ \mathcal{Q}=\{g_1 \ldots g_Q\} $ from a genome with genes $\mathcal{G}=\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$.
+In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$.
 In NEA the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.
@@
 
-Here, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$.
-This allows us to evaluate our score against the null hypothesis, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''.
-This translates to evaluate yje distribution of scores that would occure if the $Q$ query genes are randomly selected from the genome''.
-This can be expressed as a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score $s$,
+We wanted to evaluate our score under the null hypothesis, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''.
+This  formulation translates to calculating the distribution of scores that would occurs if the $Q$ query genes are randomly selected from the genome, and see how frequently a score as extreme or more extreme than the current score would appear in that distribution.
+Hence, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$, which we will give two methods for below.
+Given such a score distribution, we can express a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score, $s$, as,
 
@@
 This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elements from $N_{a-1}$ still needed to update subsequent elements.
-That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$.
+That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ S, S-1, \ldots, 0 \}$ and $c \in \{ Q, Q-1, \ldots, 0 \}$.
 For details see Listing \ref{lst:gensetdp}.
@@
   \caption{{\bf Score distribution of a query size of 10, 20 and 30 genes against the ``Glycolysis/Gluconeogenesis'' pathway.} GeneSetDP enables us to calculate the full score distribution for a query of a given size, $Q$, i.e.
-how many ways one can reach a certain score when adding up the scores for $Q$ genes.}
+how many ways one can reach a certain score when adding up the scores for $Q$ genes.
+A red dot marks the maximum score, $S$, for each case.}
   \label{fig:score_dist}
@@
   \begin{center}
-		\begin{tabular}[t]{c@{}c}
-				\includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png} &
-				\includegraphics[width=0.2\textwidth]{figures/calibration_legend.png}
+		  \begin{tabular}[t]{c}
+				\includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png} \\
+				\includegraphics[width=0.5\textwidth]{figures/calibration_legend.png}
 		\end{tabular}
@@
 We also added the calibration curve of BinoX, evaluated on the same random sets.
-On all plots the dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$, for comparison}
+The dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$, for comparison§§.}
   \label{fig:calibration}
@@
 
-\bibliography{./refs}
+\bibliography{./refs.bib}
 

### 2018-05-08 doc/genesetdp.tex
text

@@
 \begin{abstract}
-A prevailing technique to infer function from identifications from molecular biological high-throughput experiments is gene set enrichment analysis, where the identifications are compared to predefined sets of related genes, so-called pathways.
+A prevailing technique to infer function from identifications, from molecular biological high-throughput experiments, is gene set enrichment analysis, where the identifications are compared to predefined sets of related genes, so-called pathways.
 As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks, with so-called Network Enrichment Analysis (NEA).
@@
 
-Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}.
-The technique estimates which functional modules, such as complexes or pathways, that are overrepresented among a set of identified analytes.
+Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts, proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}.
+The technique estimates which functional modules, such as complexes or pathways, are overrepresented among a set of identified analytes.
 One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.
@@
 
-As it is difficult to design network randomization techniques that are completely random, such methods often introduce an undesirable bias in the statistics.
+As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics.
 For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
@@
 
-Hence, here we employ a different null model, more directly aimed at model lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''.
+Hence, here we employ a different null model, more directly aimed at modeling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''.
 We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
@@
 We also downloaded the NEA example given at the BinoX web site.
-The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways and a network definition file that, after thresholding with a score of $0.7$, gave $1244992$ links between those genes.
+The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways and a network definition file that, after thresholding the links with a score of $0.7$, gave $1244992$ links between those genes.
 
@@
 
-In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets from the investigated genome.
+In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets of size $Q$ from the investigated genome using both GeneSetDP and GeneSetMC with $B = 100000$.
 As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed.
@@
 As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.
+Here, for all three methods, we are only testing the queries for enrichment, that is, we are only concerned with a higher number of links than expected by random, and this differs from the default BinoX setting which tests for both enrichment and depletion.
 

### 2018-05-08 doc/genesetdp.tex
text

@@
 We also added the calibration curve of BinoX, evaluated on the same random sets.
-The dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$, for comparison§§.}
+The dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$, for comparison.}
   \label{fig:calibration}
@@
 Most other NEA methods also report depletion of such links.
-One could easily modify our code to instead of only measuring the higher scoring tail measuring both tails of $N(s)$.
-However, this might result in a drop in sensitivity, so we have so far no implemented such a mechanism.
+One could easily modify our code to, instead of only measuring the higher scoring tail, measuring both tails of $N(s)$.
+However, this might result in a drop in sensitivity, so we have so far not implemented such a mechanism.
 
-Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''.
+We made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''.
 Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.

### 2018-05-08 doc/genesetdp.tex
x

--- 2018-05-08 doc/genesetdp.tex
@@
 In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$.
-In NEA the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.
+In NEA the pathway maps to the genome through a network, ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.
 The pathways are scored by summing up all connections, i.e.
-the pathway score is
+the pathway score can be expressed as,
 
@@
 \end{equation}
-where $S$ is the maximal score $s$ that $Q$ genes can obtain from Equation \ref{eq:sum_ij}.
+where $S$ is the maximal score, $s$, that $Q$ genes can obtain from Equation \ref{eq:sum_ij}.
 
@@
 These sets scores were calculated with Equation \ref{eq:sum_i}.
-We counted the number of times $F_B(s)$ a score of $s$ was obtained when sampling $B$ gene sets.
+We counted the number of times, $F_B(s)$, a score of $s$ was obtained when sampling $B$ gene sets.
 We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Equation~\ref{eq:pval}.
@@
 
-Source code of an implementation of GeneSetDP and GeneSetMC, as well as code for generating the plots of this paper, is available from \url{https://github.com/statisticalbiotechnology/genesetdp}.
+The source code of an implementation of GeneSetDP and GeneSetMC, as well as code for generating the plots of this paper, is available from \url{https://github.com/statisticalbiotechnology/genesetdp}.
 

### 2018-05-08 doc/genesetdp.tex
Typos

@@
 
-In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$.
+In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway, $\mathcal{P}=\{p_1 \ldots p_P\}$.
 In NEA the pathway maps to the genome through a network, ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.
@@
 We wanted to evaluate our score under the null hypothesis, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''.
-This  formulation translates to calculating the distribution of scores that would occurs if the $Q$ query genes are randomly selected from the genome, and see how frequently a score as extreme or more extreme than the current score would appear in that distribution.
+This  formulation translates to calculating the distribution of scores that would occur if the $Q$ query genes are randomly selected from the genome, and see how frequently a score as extreme or more extreme than the current score would appear in that distribution.
 Hence, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$, which we will give two methods for below.
@@
 \end{equation}
-Such number-of-links, $l_i$, can be pre-computed for all genes $i, \ldots, G$.
+Such number-of-links, $l_i$, can be pre-computed for all genes in $\mathcal{G}$.
 
@@
 
-Furthermore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{i}{l_i}$.
-
-In our strive to calculate the score distribution, $N(s)$, we first want to investigate the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway.
+We also realized that one could calculate the full score distribution, $N(s)$, using dynamic programming.
+First, we define a function, $k_a=\sum_{\{i:l_i=a\}}1$, giving the number of genes having $a$ links to the investigated pathway, and we define $R=\max_{i}{l_i}$.
+Furthermore, we define the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway.
 This helps us formulate the recursion, using the fact that there are $k_a \choose b$ ways to select $b$ from $k_a$ elements, as
@@
 \end{equation}
-where $N_a(s,c)=0$ for all $s<0$, $c<0$ or $a<0$, with the exception for $N_{-1}(0,0)=1$.
+where $N_a(s,c)=0$ for all $s<0$, $c<0$ or $a<0$, with the exception of $N_{-1}(0,0)=1$.
 
@@
 This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elements from $N_{a-1}$ still needed to update subsequent elements.
-That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ S, S-1, \ldots, 0 \}$ and $c \in \{ Q, Q-1, \ldots, 0 \}$.
+That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add a $D_a$ to previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ S, S-1, \ldots, 0 \}$ and $c \in \{ Q, Q-1, \ldots, 0 \}$.
 For details see Listing \ref{lst:gensetdp}.
@@
 We downloaded network definitions and the BinoX software (on 2018-04-28) for comparisons from \url{https://bitbucket.org/sonnhammergroup/binox}. BinoX was run with the default parameters.
-% FIXME: add a more detailed description of the test set (i.e. organism and number of genes, pathways and network connections).
 
@@
 Here we have implemented two methods, GeneSetDP and GeneSetMC, to calculate unbiased $p$~values for inferences from NEA.
-Instead of testing our methods performance in terms of sensitivity, we here instead chose to test the method in terms of the methods statistical accuracy.
-We argue that more studies should make a point at demonstrate that their methods are well calibrated, as it is a prerequisite for measuring performance at least when ground truth of biological effect is not known.
-With this in mind, we demonstrated that our methods gave a better calibration than the popular BinoX method, by random sampling of subsets of a genome.
+Instead of testing our methods' performance in terms of sensitivity, we here instead chose to test the method in terms of the methods statistical accuracy.
+We argue that more studies should make a point at demonstrating that their methods are well-calibrated, as it is a prerequisite for measuring performance at least when the ground truth of biological effect is not known.
+With this in mind, we demonstrated that both our methods gave a better calibration than the popular BinoX method, by random sampling of subsets of a genome.
 
-Both our implementations, GeneSetDP and GeneSetMC, are intended for measuring enrichment of links between the query and the pathway set of genes.
+Both our implementations, GeneSetDP and GeneSetMC, are intended for measuring the enrichment of links between the query and the pathway set of genes.
 Most other NEA methods also report depletion of such links.

### 2018-05-09 exp/combine_plot.py
organization of exp

@@ -1,60 +1,0 @@
-import pandas as pd
-import numpy as np
-import matplotlib.pyplot as plt
-import seaborn as sns
-
-Q = 30
-
-binox = pd.read_csv('example-random/BinoX_results_Q' + str(Q) + '.tsv', sep='\t')
-
-###### only enriched
-binox.loc[:, 'enrichment'] = binox.loc[:, '#4:p.value']
-binox.loc[binox['#6:relationType'] == '-', 'enrichment'] = 1 - binox.loc[binox['#6:relationType'] == '-', '#4:p.value']
-######
-
-binox = binox.set_index('#3:NameGroupB')
-
-binox.index = [x.lower() for x in binox.index]
-
-binox_both = binox['#4:p.value']
-binox_enrich = binox['enrichment']
-
-genesetdp = pd.read_csv('example-random/genesetDP_results_Q' + str(Q) + '.tsv', sep='\t', index_col=0)
-genesetdp_sorted = genesetdp['p'].sort_values()
-genesetdp_sorted_mc = genesetdp['p_mc'].sort_values()
-genesetdp_ix = (np.arange(len(genesetdp_sorted.index))+0.5)/float(len(genesetdp_sorted))
-
-binox_both = binox_both.loc[genesetdp.index]
-binox_enrich = binox_enrich.loc[genesetdp.index]
-binox_both.loc[np.isnan(binox_both)] = 1
-binox_enrich.loc[np.isnan(binox_enrich)] = 1
-
-binox_both_sorted = binox_both.sort_values()
-binox_enrich_sorted = binox_enrich.sort_values()
-
-binox_ix = (np.arange(len(binox_both_sorted.index))+0.5)/float(len(binox_both_sorted.index))
-
-x_y = pd.DataFrame(np.arange(0,1,0.001), index = np.arange(0,1,0.001))
-x_2y = pd.DataFrame(np.arange(0,1,0.001)*2, index = np.arange(0,1,0.001))
-x_halfy = pd.DataFrame(np.arange(0,0.5,0.001), index = np.arange(0,0.5,0.001)*2)
-
-sns.set_style("ticks")
-sns.despine()
-
-plt.rc('font', size=12)
-plt.loglog(binox_ix, binox_both_sorted, '.')
-plt.loglog(binox_ix, binox_enrich_sorted, '.')
[…]

### 2018-05-09 exp/experiment.py

@@ -14,6 +14,9 @@ if len(sys.argv)>1:
 else:
     Q=25
 
+## Especify pathway here
+pathway_name = 'GLYCOLYSIS_/_GLUCONEOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
+
 ## example path
 path = "~/git/binox/example/"
 
@@ -21,8 +24,6 @@ path = "~/git/binox/example/"
 pathways_file = 'Pathways.tsv'
 pathways = pd.read_csv(path + pathways_file, sep='\t', header=None)
 pathways.columns = ['gene','pathway']
-
-pathway_name = 'GLYCOLYSIS_/_GLUCONEOGENESIS_-_HOMO_SAPIENS_(HUMAN)'
 
 pathway_genes = pathways.where(pathways.pathway == pathway_name).dropna().gene.values
 
@@ -33,7 +34,8 @@ N = genesetdp.genesetdp(k,Q)
 
 p_vals = (np.flipud(np.cumsum(np.flipud(N))) - N/2)/sum(N)
 
-p_vals_mc = mc.genesetmc(k, Q, 100000)
+n_runs = 100000 ## number of monte carlo samples
+p_vals_mc = mc.genesetmc(k, Q, n_runs)
 
 groups = pd.read_csv('example-random/Random_Signatures.tsv', sep='\t')
 
@@ -51,4 +53,4 @@ for i in range(1,10001):
     p_mc = p_vals_mc[n]
     results.loc[groupname] = [n,p, p_mc]
 
-results.to_csv('example-random/genesetDP_results.tsv', sep = '\t')
+results.to_csv('example-random/genesetDP_results_Q' + str(Q) + '.tsv', sep = '\t')

### 2018-05-09 exp/figure_calibration.py

@@ -70,14 +70,11 @@ for ax in [ax1,ax2,ax3]:
     ax.xaxis.set_tick_params(labelsize=12)
     ax.yaxis.set_tick_params(labelsize=12)
 
-# lgd = fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX'], loc='center', ncol =3, bbox_to_anchor=(0.5,0), fontsize=16)
+lgd = fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX'], loc='center', ncol =3, bbox_to_anchor=(0.5,0), fontsize=16)
 # fig.legend(['GeneSetDP', 'GeneSetMC', 'BinoX'], fontsize=15, ncol=3)
 
 
-fig.set_size_inches(15,4)
-# fig.savefig('calibration_multiple.png', bbox_extra_artists=(lgd,), bbox_inches='tight')
+# fig.set_size_inches(15,4)
 fig.savefig('calibration_multiple.png', bbox_inches='tight')
 # plt.show()
 
-# plt.savefig('calibration_multiple.png')
-# plt.show()

### 2018-05-09 exp/figure_distribuition.py

@@ -56,8 +56,8 @@ ax3.step(np.arange(len(N30)), N30, where='mid')
 ax3.plot(len(N30),1, '.', color='r', markersize=12)
 
 plt.tight_layout()
-# plt.savefig('score_distribuition_multiple.png')
-plt.show()
+plt.savefig('score_distribuition_multiple.png')
+# plt.show()
 
 
 

### 2018-05-09 doc/genesetdp.tex
subset

@@
 
-In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway, $\mathcal{P}=\{p_1 \ldots p_P\}$.
+In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ (i.e.
+$\mathcal{Q} \subseteq \mathcal{G}$), and how they relate to a pathway, $\mathcal{P}=\{p_1 \ldots p_P\}$.
 In NEA the pathway maps to the genome through a network, ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise.

### 2018-06-30 doc/genesetdp.tex
Updates after comments from reviewer and ES.

@@
 %\usepackage{microtype}
+
 \bibliographystyle{plainurl}
+
 \title{One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis}
@@
 %\EventNoEds{2}
-\EventLongTitle{18th Workshop on Algorithms in Bioinformatics (WABI 2018)}
-\EventShortTitle{WABI 2018}
-\EventAcronym{WABI}
-\EventYear{2018}
-\EventDate{20–22 August 2018}
-\EventLocation{Helsinki, Finland}
+%\EventShortTitle{WABI 2018}
+%\EventAcronym{WABI}
+%\EventYear{2018}
+%\EventDate{20–22 August 2018}
+%\EventLocation{Helsinki, Finland}
 %\EventLogo{}
@@
 \end{equation}
-where $S$ is the maximal score, $s$, that $Q$ genes can obtain from Equation \ref{eq:sum_ij}.
+where $S$ is the maximal score, $s$, that $Q$ genes can obtain from Equation~(\ref{eq:sum_ij}).
 
-We can reformulate Equation \ref{eq:sum_ij} by defining the number-of-links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
+We can reformulate Equation~(\ref{eq:sum_ij}) by defining the number-of-links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
 \begin{equation}
@@
 We randomly selected sets of $Q$ genes from $\mathcal{G}$.
-These sets scores were calculated with Equation \ref{eq:sum_i}.
+These sets scores were calculated with Equation~(\ref{eq:sum_i}).
 We counted the number of times, $F_B(s)$, a score of $s$ was obtained when sampling $B$ gene sets.
-We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Equation~\ref{eq:pval}.
+We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Equation~(\ref{eq:pval}).
 We refer to this procedure as GeneSetMC.
@@
 
-We also realized that one could calculate the full score distribution, $N(s)$, using dynamic programming.
-First, we define a function, $k_a=\sum_{\{i:l_i=a\}}1$, giving the number of genes having $a$ links to the investigated pathway, and we define $R=\max_{i}{l_i}$.
-Furthermore, we define the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway.
-This helps us formulate the recursion, using the fact that there are $k_a \choose b$ ways to select $b$ from $k_a$ elements, as
 \begin{equation}
@@
 
-\begin{lstlisting}[caption={
-The central part of the dynamic programing algorithm for finding $N(s)$.
+\begin{lstlisting}[language=Python, caption={ The central part of the dynamic programing algorithm for finding $N(s)$.
 The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input.
-The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt comb(a,b)}, which calculates $ a \choose b $.
-}
-, label=lst:gensetdp,captionpos=t,float,abovecaptionskip=-\medskipamount]
+The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt comb(a,b)}, which calculates $ a \choose b $.}, label=lst:gensetdp, captionpos=t, float, abovecaptionskip=-\medskipamount]
 def genesetdp(k,Q):
@@
 \[
-N_a(s,c)=\sum_{b=1}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b) + N_{a-1}(s,c)=D_a+N_{a-1}(s,c).
+N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b)=\sum_{b=1}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b) + N_{a-1}(s,c).
 \]
@@
 This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elements from $N_{a-1}$ still needed to update subsequent elements.
-That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add a $D_a$ to previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ S, S-1, \ldots, 0 \}$ and $c \in \{ Q, Q-1, \ldots, 0 \}$.
+That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add $\sum_{b=1}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b)$ to previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ S, S-1, \ldots, 0 \}$ and $c \in \{ Q, Q-1, \ldots, 0 \}$.
 For details see Listing \ref{lst:gensetdp}.
@@
 
-We implemented a Python program that reads network and pathway definition files and scores a query sets against a pathway according to Equation \ref{eq:sum_i}, using GeneSetDP and GeneSetMC described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}.
+We implemented a Python program that reads network and pathway definition files and scores a query sets against a pathway according to Equation~(\ref{eq:sum_i}), using GeneSetDP and GeneSetMC described in the Algorithm section, that enabled us to assign $p$~values according to Equation~(\ref{eq:pval}).
 We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).
@@
 
+To test the consistency of the methods, we also investigated the correspondance between the $p$~values reported by GeneSetDP and GeneSetMC.
+As can be seen in Figure~\ref{fig:pscatter}, the relative difference is larger for lower $p$~values, likely due to the sampling errors of GeneSetMC.
+
+\begin{figure}[htb]
+  \begin{center}
+		  \begin{tabular}[t]{c}
+        FIXME: Make this scatterplot in log-log, GeneSetDP on x-axis, GeneSetMC on y-axis.
+        (Idea from reviewer comment in email 2018-06-08.)
+%				\includegraphics[width=1.0\textwidth]{FIXME}
+		\end{tabular}
+  \end{center}
+  \caption{{\bf Correspondance between GeneSetDP's and GeneSetMC's $p$~values.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
+We subsequently plotted the obtained $p$~values against each other .}
+  \label{fig:pscatter}
+\end{figure}
+
 \section*{Discussion}

### 2018-07-02 doc/genesetdp.tex
added figure comparing p values between methods

@@
 		  \begin{tabular}[t]{c}
-        FIXME: Make this scatterplot in log-log, GeneSetDP on x-axis, GeneSetMC on y-axis.
-        (Idea from reviewer comment in email 2018-06-08.)
-%				\includegraphics[width=1.0\textwidth]{FIXME}
+        % FIXME: Make this scatterplot in log-log, GeneSetDP on x-axis, GeneSetMC on y-axis.
+        % (Idea from reviewer comment in email 2018-06-08.)
+				\includegraphics[width=1.0\textwidth]{figures/scatter_dp_mc.png}
 		\end{tabular}
@@
   \caption{{\bf Correspondance between GeneSetDP's and GeneSetMC's $p$~values.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
-We subsequently plotted the obtained $p$~values against each other .}
+We subsequently plotted the obtained $p$~values against each other.
+Again, the dashed line shows $y = x$ and the dotted lines $y = 2x$ and $y = 0.5x$.}
   \label{fig:pscatter}

### 2018-07-31 doc/genesetdp.tex
remove mention to low pvalues
Describing the recursion

--- 2018-06-30 doc/genesetdp.tex
@@
-\documentclass[a4paper,american]{lipics-v2016}
-%\usepackage[USenglish]{babel}
-%\usepackage{microtype}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 
-\bibliographystyle{plainurl}
+\usepackage{listings}
+\usepackage[T1]{fontenc}
 
-\title{One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis}
-\titlerunning{Null models for Network Enrichment Analysis} %optional, in case that the title is too long; the running title should fit into the top page column
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
 
-%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
-\author[1]{Gustavo S. Jeuken}
-\author[2]{Lukas K\"{a}ll}
-\affil[1]{Science for Life Laboratory, School of
-Engineering Sciences in Chemistry, Biotechnology and Health,
-Royal Institute of Technology -- KTH, Box 1031, 17121 Solna, Sweden\\ \texttt{gustavo.jeuken@scilifelab.se}}
-\affil[2]{Science for Life Laboratory, School of
-Engineering Sciences in Chemistry, Biotechnology and Health,
-Royal Institute of Technology -- KTH, Box 1031, 17121 Solna, Sweden\\ \texttt{lukas.kall@scilifelab.se}}
-\authorrunning{G. S. Jeuken and L. K\"{a}ll}
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
 
-\Copyright{Gustavo S. Jeuken and Lukas K\"{a}ll}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
 
-\subjclass{G.3.
-Probabilistic algorithms (including Monte Carlo)}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html .
-E.g., cite as "F.1.1 Models of Computation".
-\keywords{Gene Set Analysis; Calibration; Null model; p value; Sampling}% mandatory: Please provide 1-5 keywords
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
 
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\EventEditors{John Q. Open and Joan R. Acces}
-%\EventNoEds{2}
-%\EventShortTitle{WABI 2018}
-%\EventAcronym{WABI}
-%\EventYear{2018}
-%\EventDate{20–22 August 2018}
-%\EventLocation{Helsinki, Finland}
-%\EventLogo{}
-%\SeriesVolume{42}
-%\ArticleNo{23}
-% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace}
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
+
+%\titlerunning{Null models for Network Enrichment Analysis}
+%\authorrunning{G. S. Jeuken and L. K\"{a}ll}
+
+%\Copyright{Gustavo S. Jeuken and Lukas K\"{a}ll}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
 \begin{document}
+\vspace*{0.2in}
 
-\maketitle
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Gustavo S. Jeuken\textsuperscript{1} and
+Lukas K\"{a}ll\textsuperscript{1,*}
+\\
+\bigskip
+\textbf{1}Science for Life Laboratory, School of Engineering Sciences in Chemistry, Biotechnology and Health, KTH -- Royal Institute of Technology, Box 1031, 17121 Solna, Sweden
+\\
+\bigskip
 
-\begin{abstract}
-A prevailing technique to infer function from identifications, from molecular biological high-throughput experiments, is gene set enrichment analysis, where the identifications are compared to predefined sets of related genes, so-called pathways.
-As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks, with so-called Network Enrichment Analysis (NEA).
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* lukas.kall@scilifelab.se
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
+A prevailing technique to infer function from identifications, from molecular biological high-throughput experiments, is gene set enrichment analysis, where the identifications are compared to predefined sets of related genes often refered to as pathways.
+As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks.
+While the termonolgy vary in the litterature, we will here refer to such methods as Network Enrichment Analysis (NEA).
 Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations.
@@
 We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX.
-
-\end{abstract}
 
@@
 Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}.
-This is known as Network Enrichment Analysis (NEA).
+While the termonolgy vary in the litterature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).
 
@@
 
-We can reformulate Equation~(\ref{eq:sum_ij}) by defining the number-of-links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
+We can reformulate Equation~(\ref{eq:sum_ij}) by defining the number of links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
 \begin{equation}
@@
 \end{equation}
-Such number-of-links, $l_i$, can be pre-computed for all genes in $\mathcal{G}$.
+Such number of links, $l_i$, can be pre-computed for all genes in $\mathcal{G}$.
+For later purposes we also define a mapping, $k_a=\sum_{\{i:l_i=a\}}1$, giving the number of genes having $a$ links to the investigated pathway.
 
@@
 
+An alternative approach is to calculate the exact score distribution using dynamic programming. This can be done if we formulate $N(s)$ as a recrsion, by observing a partial sum function,  $N_a(s,c)$,
+ which expresses the number of ways to select a set of $c$ query genes, each with $\le a$  links to the pathway set and obtain a score of $s$.
+If we know the distribution of the number of ways to select genes with $\le a-1$ genes,  $N_{a-1}(s,c)$, for all $s$ and $c$,   $N_a(s,c)$ can be determined by investigating the ways to select genes with exactly $a$ links to the investigated pathway.
+As there are $k_a \choose b$ ways to select $b$ from $k_a$ elements, there contribution to $N_a(s,c)$ from the selection of  $b$ genes with exactly $a$ links are ${k_a \choose b} N_{a-1}(s-ab,c-b)$.
+When evaluating all posibile values of $b$ we see that
  \begin{equation}
-N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b),
+N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b).
 \end{equation}
-where $N_a(s,c)=0$ for all $s<0$, $c<0$ or $a<0$, with the exception of $N_{-1}(0,0)=1$.
+We also see that $N_a(s,c)=0$ for all $s<0$, $c<0$ or $a<0$, with the exception of $N_{-1}(0,0)=1$.
 
-The final score distribution is given by $N(s)=N_R(s,Q)$.
+The final score distribution is given by $N(s)=N_R(s,Q)$, where $R=\max_{i}{l_i}$, i.e the larges number of links to the pathway from any gene in the genome.
 
@@
 
-To illustrate the $p$~value calculation procedure, we plotted the score distributions $N(s)$ for a query size of 10, 20 and 30 genes in Figure \ref{fig:score_dist}, using GeneSetDP.
+To illustrate the $p$~value calculation procedure, we plotted the score distributions $N(s)$ for a query size of 10, 20 and 30 genes in Fig \ref{fig:score_dist}, using GeneSetDP.
 
@@
 As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed.
-To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration} for $10000$ randomly assembled queries from the human genome.
+To test this we plotted the $p$~values against their quantile in Fig \ref{fig:calibration} for $10000$ randomly assembled queries from the human genome.
 As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.
@@
 To test the consistency of the methods, we also investigated the correspondance between the $p$~values reported by GeneSetDP and GeneSetMC.
-As can be seen in Figure~\ref{fig:pscatter}, the relative difference is larger for lower $p$~values, likely due to the sampling errors of GeneSetMC.
+As can be seen in Fig~\ref{fig:pscatter}, the relative difference is larger for lower $p$~values, likely due to the sampling errors of GeneSetMC.
 

### 2018-07-31 doc/genesetdp.tex

@@
 To test the consistency of the methods, we also investigated the correspondance between the $p$~values reported by GeneSetDP and GeneSetMC.
-As can be seen in Fig~\ref{fig:pscatter}, the relative difference is larger for lower $p$~values, likely due to the sampling errors of GeneSetMC.
+As can be seen in Figure~\ref{fig:pscatter}, the results are consistent as expected.
 

### 2018-07-31 doc/genesetdp.tex
Prettifying affiliation

@@
 \bigskip
-\textbf{1}Science for Life Laboratory, School of Engineering Sciences in Chemistry, Biotechnology and Health, KTH -- Royal Institute of Technology, Box 1031, 17121 Solna, Sweden
+\textbf{1} Science for Life Laboratory, School of Engineering Sciences in Chemistry, Biotechnology and Health, KTH -- Royal Institute of Technology, Box 1031, 17121 Solna, Sweden
 \\
@@
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* lukas.kall@scilifelab.se
+*lukas.kall@scilifelab.se
 \end{flushleft}

### 2018-07-31 doc/genesetdp.tex
Equation -> Eq

@@
 \end{equation}
-where $S$ is the maximal score, $s$, that $Q$ genes can obtain from Equation~(\ref{eq:sum_ij}).
+where $S$ is the maximal score, $s$, that $Q$ genes can obtain from Eq~(\ref{eq:sum_ij}).
 
-We can reformulate Equation~(\ref{eq:sum_ij}) by defining the number of links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
+We can reformulate Eq~(\ref{eq:sum_ij}) by defining the number of links, $l_i=\sum_{j=1}^P x_{ij}$, which gives us a score,
 \begin{equation}
@@
 We randomly selected sets of $Q$ genes from $\mathcal{G}$.
-These sets scores were calculated with Equation~(\ref{eq:sum_i}).
+These sets scores were calculated with Eq~(\ref{eq:sum_i}).
 We counted the number of times, $F_B(s)$, a score of $s$ was obtained when sampling $B$ gene sets.
-We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Equation~(\ref{eq:pval}).
+We see that $F_B(s)$ will approximately have the same shape as $N(s)$ when selecting a large $B$, and we can hence calculate $p$~values by replacing $N$ with $F_B$ in Eq~(\ref{eq:pval}).
 We refer to this procedure as GeneSetMC.
@@
 
-We implemented a Python program that reads network and pathway definition files and scores a query sets against a pathway according to Equation~(\ref{eq:sum_i}), using GeneSetDP and GeneSetMC described in the Algorithm section, that enabled us to assign $p$~values according to Equation~(\ref{eq:pval}).
+We implemented a Python program that reads network and pathway definition files and scores a query sets against a pathway according to Eq~(\ref{eq:sum_i}), using GeneSetDP and GeneSetMC described in the Algorithm section, that enabled us to assign $p$~values according to Eq~(\ref{eq:pval}).
 We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).

### 2018-07-31 doc/genesetdp.tex
Added ref (from discussion with Bill Noble during asms).

@@
 
-Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts, proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}.
+Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts, proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten,goeman2007analyzing}.
 The technique estimates which functional modules, such as complexes or pathways, are overrepresented among a set of identified analytes.
@@
 
-The final score distribution is given by $N(s)=N_R(s,Q)$, where $R=\max_{i}{l_i}$, i.e the larges number of links to the pathway from any gene in the genome.
+The final score distribution is given by $N(s)=N_R(s,Q)$, where $R=\max_{i}{l_i}$, i.e the largest number of links to the pathway from any gene in the genome.
 

### 2018-08-01 doc/genesetdp.tex
Changed title

@@
 {\Large
-\textbf\newline{One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{A simple null model for inferences from network enrichment analysis} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 
+\section*{Acknowledgments}
+
+LK was supported by a grant from Swedish Foundation for Strategic Research (BD15-0043).
+
 \bibliography{./refs.bib}

### 2018-08-01 doc/genesetdp.tex
GEA -> ORA after talking to Andrey

@@
 \section*{Abstract}
-A prevailing technique to infer function from identifications, from molecular biological high-throughput experiments, is gene set enrichment analysis, where the identifications are compared to predefined sets of related genes often refered to as pathways.
+A prevailing technique to infer function from lists of identifications, from molecular biological high-throughput experiments, is over-representation analysis, where the identifications are compared to predefined sets of related genes often refered to as pathways.
 As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks.
@@
 
-Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts, proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten,goeman2007analyzing}.
+Over-Representation Analysis (ORA) is commonly used to infer function from sets of analytes such as genes, transcripts, proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten,goeman2007analyzing}.
 The technique estimates which functional modules, such as complexes or pathways, are overrepresented among a set of identified analytes.
-One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.
+One prominent application of the technique is expression analysis, where ORA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.
 
-Most GEA methods are assessing the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test.
+Most ORA methods are assessing the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test.
 However, variants such as Gene Set Enrichment Analysis (GSEA)\cite{subramanian2005gene} also includes information on expression levels of the analytes of the query set.
 
-A limiting factor of GEA is the pathway definition databases.
+A limiting factor of ORA is the pathway definition databases.
 While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases.

### 2018-08-01 doc/genesetdp.tex
Rewording

@@
 
-A limiting factor of ORA is the pathway definition databases.
+A limiting factor of ORA is the quality of the databases defining pathways.
 While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases.
@@
 	\begin{center}
-		\includegraphics[width=1.0\textwidth]{figures/score_distribuition_multiple.png}
+		\includegraphics[width=.9\textwidth]{figures/score_distribuition_multiple.png}
     \end{center}
@@
 		  \begin{tabular}[t]{c}
-				\includegraphics[width=1.0\textwidth]{figures/calibration_multiple.png} \\
-				\includegraphics[width=0.5\textwidth]{figures/calibration_legend.png}
+				\includegraphics[width=.9\textwidth]{figures/calibration_multiple.png} \\
+				\includegraphics[width=0.45\textwidth]{figures/calibration_legend.png}
 		\end{tabular}
@@
 		  \begin{tabular}[t]{c}
-        % FIXME: Make this scatterplot in log-log, GeneSetDP on x-axis, GeneSetMC on y-axis.
         % (Idea from reviewer comment in email 2018-06-08.)
-				\includegraphics[width=1.0\textwidth]{figures/scatter_dp_mc.png}
+				\includegraphics[width=.9\textwidth]{figures/scatter_dp_mc.png}
 		\end{tabular}
@@
   \label{fig:pscatter}
+{\em FIXME: Rename x/y-labels to GeneSetXX $p$~value}
 \end{figure}

### 2018-08-01 doc/coverletter.tex
Added coverletter

\documentclass[a4paper]{letter}
\usepackage{url}

\begin{document}

\begin{letter}{}
\vspace*{-5.0cm}

\opening{Dear editor:}

Enclosed you will find a manuscript entitled: ``A simple null model for inferences from network enrichment analysis''. The manuscript describes an in our minds natural null model for network enrichment analysis, as well as two implementations of the null model as an exact test and a sampling method.

Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations.  In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters. We also demonstrate our implementations to render far better calibration than the state-of the-art method, BinoX.

\vspace{0.15cm}

The manuscript is acompanied by an open github acount with all software and all tests that we conducted, available for reviewers, or the general public. The manuscript is an original contribution and none of the work it describes has been published elsewhere except for in abstract form. The manuscript will remain with you and will not be submitted elsewhere, until you made a decision as to its suitability for publication in PLOS ONE.

We suggest the following reviewers:

\begin{itemize}

\item Dr. Lieven Clement,
Gent University, \\
\url{Lieven.Clement@ugent.be}

\item Dr. Andrey Alexeyenko, Karolinska Institutet, \\
\url{Andrey.Alexeyenko@ki.se}

\item Professor Ernst C. Wit, University of Groningen\\
\url{e.c.wit@rug.nl}

\item Professor Alfonso Valencia, Barcelona Supercomputing Center,\\
\url{alfonso.valencia@bsc.es}

\item Professor Reinhard Schneider,  Luxembourg Centre for Systems Biomedicine \\
\url{Reinhard.Schneider@uni.lu}

\end{itemize}

\vspace*{1.5em}

Sincerely,\\[2em]
Mr. Gustavo Jeuken and\\
Prof. Lukas K\"all, \\
Science~for~Life~Laboratory, School of Biotechnology,\\
Royal Institute of Technology - KTH, 17165 Solna, Sweden\\
\url{lukas.kall@scilifelab.se}\\
Tel: +46 8 52481196, Fax: +46~8~52481425

\end{letter}
\end{document}

### 2018-08-01 exp/figure_scatter.py
change figure label

@@ -36,10 +36,10 @@ ax1.set_title('Q = 10', fontsize=18)
 ax2.set_title('Q = 20', fontsize=18)
 ax3.set_title('Q = 30', fontsize=18)
 
-ax1.set_ylabel('GeneSetDP', fontsize = 16)
-ax1.set_xlabel('GeneSetMC', fontsize = 16)
-ax2.set_xlabel('GeneSetMC', fontsize = 16)
-ax3.set_xlabel('GeneSetMC', fontsize = 16)
+ax1.set_ylabel('GeneSetDP $p$-values', fontsize = 16)
+ax1.set_xlabel('GeneSetMC $p$-values', fontsize = 16)
+ax2.set_xlabel('GeneSetMC $p$-values', fontsize = 16)
+ax3.set_xlabel('GeneSetMC $p$-values', fontsize = 16)
 
 for ax in [ax1,ax2,ax3]:
     ax.xaxis.set_tick_params(labelsize=12)

### 2018-08-02 exp/figure_scatter.py
fix label

@@ -36,10 +36,10 @@ ax1.set_title('Q = 10', fontsize=18)
 ax2.set_title('Q = 20', fontsize=18)
 ax3.set_title('Q = 30', fontsize=18)
 
-ax1.set_ylabel('GeneSetDP $p$-values', fontsize = 16)
-ax1.set_xlabel('GeneSetMC $p$-values', fontsize = 16)
-ax2.set_xlabel('GeneSetMC $p$-values', fontsize = 16)
-ax3.set_xlabel('GeneSetMC $p$-values', fontsize = 16)
+ax1.set_ylabel('GeneSetDP $p$-value', fontsize = 16)
+ax1.set_xlabel('GeneSetMC $p$-value', fontsize = 16)
+ax2.set_xlabel('GeneSetMC $p$-value', fontsize = 16)
+ax3.set_xlabel('GeneSetMC $p$-value', fontsize = 16)
 
 for ax in [ax1,ax2,ax3]:
     ax.xaxis.set_tick_params(labelsize=12)

### 2018-08-02 doc/coverletter.tex
spell check

@@
 Enclosed you will find a manuscript entitled: ``A simple null model for inferences from network enrichment analysis''.
-The manuscript describes an in our minds natural null model for network enrichment analysis, as well as two implementations of the null model as an exact test and a sampling method.
+The manuscript describes an in our minds natural null model for network enrichment analysis (NEA), as well as two implementations of the null model as an exact test and a sampling method.
 
@@
 In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.
-We also demonstrate our implementations to render far better calibration than the state-of the-art method, BinoX.
+We also demonstrate our implementations to render far better calibration than the state-of-the-art method, BinoX.
 
@@
 
-The manuscript is acompanied by an open github acount with all software and all tests that we conducted, available for reviewers, or the general public.
+While the manuscript does not present a user-friendly software, it instead focuses on how to accurately assess error probabilities for NEA.
+We note that several methods in the field do not report calibrated statistics and hence are hard to compare.
+In our minds, accurate statistics is the first step in the process of improving the field of NEA.
+
+\vspace{0.15cm}
+
+The manuscript is accompanied by an open Github account with all software and all tests that we conducted, available for reviewers, or the general public.
 The manuscript is an original contribution and none of the work it describes has been published elsewhere except for in abstract form.
-The manuscript will remain with you and will not be submitted elsewhere, until you made a decision as to its suitability for publication in PLOS ONE.
+The manuscript will remain with you and will not be submitted elsewhere until you made a decision as to its suitability for publication in PLOS ONE.
 

### 2018-08-02 doc/genesetdp.tex
change figure label
fix label
spell check

@@
 \section*{Abstract}
-A prevailing technique to infer function from lists of identifications, from molecular biological high-throughput experiments, is over-representation analysis, where the identifications are compared to predefined sets of related genes often refered to as pathways.
+A prevailing technique to infer function from lists of identifications, from molecular biological high-throughput experiments, is over-representation analysis, where the identifications are compared to predefined sets of related genes often referred to as pathways.
 As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks.
-While the termonolgy vary in the litterature, we will here refer to such methods as Network Enrichment Analysis (NEA).
+While the terminology varies in the literature, we will here refer to such methods as Network Enrichment Analysis (NEA).
 Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations.
@@
 Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}.
-While the termonolgy vary in the litterature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).
+While the terminology varies in the literature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).
 
@@
 Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}.
-We argue that such test are much easier implemented as tests of the query set, i.e.
+We argue that such test is much easier implemented as tests of the query set, i.e.
 is the query set a randomly selected set of genes or is there an overrepresentation of well-connected genes in the query set?
@@
 
-In network-based gene set analysis one score a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ (i.e.
+In network-based gene set analysis one scores a query set of genes, $ \mathcal{Q}=\{g_1 \ldots g_Q\}$, from a genome with genes, $\mathcal{G}=\{g_1 \ldots g_G\}$ (i.e.
 $\mathcal{Q} \subseteq \mathcal{G}$), and how they relate to a pathway, $\mathcal{P}=\{p_1 \ldots p_P\}$.
@@
 This formulation translates to calculating the distribution of scores that would occur if the $Q$ query genes are randomly selected from the genome, and see how frequently a score as extreme or more extreme than the current score would appear in that distribution.
-Hence, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$, which we will give two methods for below.
+Hence, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtain a score, $s$, which we will give two methods for below.
 Given such a score distribution, we can express a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score, $s$, as,
@@
 \end{lstlisting}
+
 There is a memory efficient implementation. We first note that
@@
 
-To test the consistency of the methods, we also investigated the correspondance between the $p$~values reported by GeneSetDP and GeneSetMC.
+To test the consistency of the methods, we also investigated the correspondence between the $p$~values reported by GeneSetDP and GeneSetMC.
 As can be seen in Figure~\ref{fig:pscatter}, the results are consistent as expected.
@@
   \label{fig:pscatter}
-{\em FIXME: Rename x/y-labels to GeneSetXX $p$~value}
+% {\em FIXME: Rename x/y-labels to GeneSetXX $p$~value}
 \end{figure}
@@
 
-LK was supported by a grant from Swedish Foundation for Strategic Research (BD15-0043).
+LK was supported by a grant from the Swedish Foundation for Strategic Research (BD15-0043).
 

### 2018-09-21 doc/response.tex
Adding response.

\documentclass[a4paper]{article}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\newcommand{\breview}{\begin{quotation}\begin{bf}\noindent}
\newcommand{\ereview}{\end{bf}\end{quotation}}
\newcommand{\reviewbullet}[1]{\breview \begin{itemize}\item #1 \end{itemize}\ereview}
\begin{document}

Dear editors:

We have attached a revised version of our manuscript ``A simple null model for inferences from network enrichment analysis'', and a diff
against our original version. We have also answered the reviewers comments
below.

Yours Sincerely,\\[1.5cm]
Gustavo Jeuken and
Lukas K\"{a}ll
\\

\section*{Reviewer \#1}
\breview
Jeunken and Käll describe a method for generating a null distribution for so called ``network enrichment analyses'', which are sometimes framed as ``crosstalk'' analyses. The only data presented is showing that the method generates accurate pvalues under the null.
\medskip
As far as I can tell the work is technically correct. Their statement of the null hypothesis they are testing is sufficiently precise. I agree with the authors that their conceptualization of the problem is simpler than methods like BinoX. But I disagree that the interpretation can be assumed to be meaningful, and the authors provide no evidence.
\ereview
The reviewer points out that the manuscript only discusses the calibration of $p$ values, and that actually never uses any positive tests.
\ldots
It is also worth noting that according to the Criteria For Publication PLOS ONE, (\url{https://journals.plos.org/plosone/s/criteria-for-publication}) conceved meaningfulness is not a publication criteria.

\medskip

\breview
They claim that BinoX yields very inaccurate p-values. If the results yielded by the proposed method are basically the same as BinoX (in terms of ranks), then there is no major problem. But the authors state that they don’t want to get into evaluating “performance in terms of sensitivity”. I find this disingenuous; they have to at least show how their ranked results compare to BinoX on real data and do some kind of benchmarking. This is important because I believe there is a difference in the null hypothesis stated in the current work and that used by BinoX that makes the comparison presented here unfair – they are not testing the same thing.

\medskip
The crux is that the method proposed here focuses on the properties of the query genes; randomly sampling query genes makes the proposed method a test of whether the query genes are unusual, not necessarily specifically with respect to a particular pathway. For example, if the query contains a hub gene (assuming hubs are rare) then it might be found “significantly” connected to many pathways. The “feature unlikely to occur by chance” is that the query contains a hub gene, not that it connects to many pathways. To state it in an extreme way, in this situation biologist should skip the pathway analysis and just take note of the fact that the query includes a hub. Using the authors’ method without this insight could lead to very misleading conclusions.

\medskip
As I understand it, BinoX tries to account for these difficulties by modeling realizations of the network that retain key topological properties like node degree distribution while holding the query constant (in effect). The current authors criticize the BinoX approach, but ignore the problem entirely. A closer (but not very close) approximation of the BinoX approach might be, for the Monte Carlo method proposed here, to bias the sampling of Q’ from G such that the distribution of the node degrees of the genes in Q’ resembles the distribution of the actual query set of interest Q. In any case, I believe that the BinoX approach is not directly comparable. The relationship needs to be discussed and some demonstration of the method yielding useful and specific results is needed. If the authors cannot do this then they must remove the discussion of BinoX. Without that, and without any benchmarking, the paper is essentially about an algorithm to compute N(s), with no biological content or relevance.
\ereview

\breview
Minor: There are some typos I noted such as 'litterature' and 'posibile'.
\ereview

\end{document}

### 2018-10-01 doc/genesetdp.tex
text

@@
 We also implemented a random sampling method, carrying out the same task.
-We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX.
+We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX, while also providing statistics that are easier to interpret.
 
@@
 
-As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics.
+% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics.
 For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
-Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
+% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
+The methods that are often used for network randomization, while computationaly efficient, are often very mentally untractable. FIXME: follow the reasoning
+We will then come to a null hypothesis for which rejection is uninformative, defeating the purpose of doing a statistical test in the first place.
 
@@
 We argue that such test is much easier implemented as tests of the query set, i.e.
-is the query set a randomly selected set of genes or is there an overrepresentation of well-connected genes in the query set?
+is the query set a randomly selected set of genes or is there an over representation of well-connected genes in the query set?
 
@@
 
-An alternative approach is to calculate the exact score distribution using dynamic programming. This can be done if we formulate $N(s)$ as a recrsion, by observing a partial sum function,  $N_a(s,c)$,
+An alternative approach is to calculate the exact score distribution using dynamic programming.
+This can be done if we formulate $N(s)$ as a recursion, by observing a partial sum function,  $N_a(s,c)$,
  which expresses the number of ways to select a set of $c$ query genes, each with $\le a$  links to the pathway set and obtain a score of $s$.
@@
 As there are $k_a \choose b$ ways to select $b$ from $k_a$ elements, there contribution to $N_a(s,c)$ from the selection of  $b$ genes with exactly $a$ links are ${k_a \choose b} N_{a-1}(s-ab,c-b)$.
-When evaluating all posibile values of $b$ we see that
+When evaluating all possible values of $b$ we see that
  \begin{equation}
@@
 Here, for all three methods, we are only testing the queries for enrichment, that is, we are only concerned with a higher number of links than expected by random, and this differs from the default BinoX setting which tests for both enrichment and depletion.
+Also of note is the fact that, while we display all methods in the same figure, each represents its own calibration, and that the Spearman's correlation between BinoX and the other methods is fairly low ($0.49$, $0.55$ and $0.55$ for $Q=10$, $20$ and $30$ respectively)
 
@@
   \end{center}
-  \caption{{\bf Correspondance between GeneSetDP's and GeneSetMC's $p$~values.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
+  \caption{{\bf Correspondence between GeneSetDP's and GeneSetMC's $p$~values.} We selected 10000 random sets of 10, 20 and 30 genes from the investigated genome definition and calculated their $p$ values with GeneSetMC and GeneSetDP against the ``Glycolysis/Gluconeogenesis'' pathway.
 We subsequently plotted the obtained $p$~values against each other.
@@
   \label{fig:pscatter}
-% {\em FIXME: Rename x/y-labels to GeneSetXX $p$~value}
 \end{figure}

### 2018-10-02 doc/genesetdp.tex
intro

@@
 % Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
-The methods that are often used for network randomization, while computationaly efficient, are often very mentally untractable. FIXME: follow the reasoning
-We will then come to a null hypothesis for which rejection is uninformative, defeating the purpose of doing a statistical test in the first place.
+The methods that are used for network randomization, while computationaly efficient, are often mentally untractable as they include a big number of steps that are subject to complex constrains.
+As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow contrains that essentially depends on all previous steps.
+And so, while in the end we do come to a network that is randomized while preserving some desired topological features, the nested constrains make it very hard to describe the probability space to which this network belongs.
+We will then come to a null hypothesis for which rejection is uninformative, defeating the purpose the statistical test.
 
@@
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
+Of note is the fact that our dinamic programming algorithm does not make any assumptions on the nature of the probability distribuition and instead calculates the exact distribuition over a given network.
 The algorithms enabled us to calculate well-defined statistics of inferences from NEA.  Using simulations we demonstrated that both the algorithms produced unbiased statistics.
@@
 Here, for all three methods, we are only testing the queries for enrichment, that is, we are only concerned with a higher number of links than expected by random, and this differs from the default BinoX setting which tests for both enrichment and depletion.
-Also of note is the fact that, while we display all methods in the same figure, each represents its own calibration, and that the Spearman's correlation between BinoX and the other methods is fairly low ($0.49$, $0.55$ and $0.55$ for $Q=10$, $20$ and $30$ respectively)
+Also of note is the fact that, while we display all methods in the same figure, each represents its own calibration, and that the Spearman's correlation between BinoX and the other methods is fairly low ($\rho = 0.49$, $0.55$ and $0.55$ for $Q=10$, $20$ and $30$ respectively)
 

### 2018-10-04 doc/genesetdp.tex
.

@@
 Hence, here we employ a different null model, more directly aimed at modeling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''.
-We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
+We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size, noting here that we do not need to make any assumptions on the nature of the score distribuition.
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
-Of note is the fact that our dinamic programming algorithm does not make any assumptions on the nature of the probability distribuition and instead calculates the exact distribuition over a given network.
-The algorithms enabled us to calculate well-defined statistics of inferences from NEA.  Using simulations we demonstrated that both the algorithms produced unbiased statistics.
+The algorithms enabled us to calculate well-defined statistics of inferences from NEA that we argue is more intuitive from an end-user perspective and thus more likely to be interpreted in a correct manner.
+Using simulations we demonstrated that both the algorithms produced unbiased statistics.
 

### 2018-10-11 doc/genesetdp.tex
pair edited text

@@
 The significance of the inferences from NEA is assessed by null models that investigate the number of expected random links from the query gene set to the pathway.
-Previous efforts have settled for various parametric null distributions, where the parameters are estimated by Monte Carlo-simulations that permute the network definitions by methods with varying degrees of topological conservation.
+Previous efforts have settled for network randomization, where the network definitions are randomly constructed by methods with varying degrees of topological conservation in respect to the original network.
 
@@
 The methods that are used for network randomization, while computationaly efficient, are often mentally untractable as they include a big number of steps that are subject to complex constrains.
-As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow contrains that essentially depends on all previous steps.
-And so, while in the end we do come to a network that is randomized while preserving some desired topological features, the nested constrains make it very hard to describe the probability space to which this network belongs.
-We will then come to a null hypothesis for which rejection is uninformative, defeating the purpose the statistical test.
+As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow contrains that depends on previous link assignments.
+While this results in a network that is both randomized and contain some preserved desired topological features, these nested constrains make it very hard to verify that the resulting network is representative.
+In theory there is a space of all possible networks that could be made up from a certain node degree distribution, however, it is not clear that the  second-order conservation method samples evenly from that distribution.
+And if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose the statistical test.
 
 Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}.
-We argue that such test is much easier implemented as tests of the query set, i.e.
+However, we argue that the network pertubation theory used in most flavors of NEA is a red herring.
+The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in repect to a pathway.
+In this question, the network is not relevant.
+A network may help to increase the sensitivity of ORA methods, but it is better left out of the null formation of the null hypothesis.
+We argue that such test is naturally implemented as tests of the query set, i.e.
 is the query set a randomly selected set of genes or is there an over representation of well-connected genes in the query set?
@@
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
-The algorithms enabled us to calculate well-defined statistics of inferences from NEA that we argue is more intuitive from an end-user perspective and thus more likely to be interpreted in a correct manner.
-Using simulations we demonstrated that both the algorithms produced unbiased statistics.
+Combining the GeneSetDP and  GeneSetMC with the scoring system of the popular method BinoX, we using simulations to demonstrate that both our algorithms produced unbiased statistics.
 

### 2018-10-12 doc/genesetdp.tex
Edits

@@
 % Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
-The methods that are used for network randomization, while computationaly efficient, are often mentally untractable as they include a big number of steps that are subject to complex constrains.
+The methods that are used for network randomization, are often mentally untractable as they include a big number of steps that are subject to complex constrains.
 As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow contrains that depends on previous link assignments.
@@
 In theory there is a space of all possible networks that could be made up from a certain node degree distribution, however, it is not clear that the  second-order conservation method samples evenly from that distribution.
-And if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose the statistical test.
+And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose the statistical test.
 
-Users of NEA typically want to test if there is a functional association between a query set of genes and a pathway\cite{alexeyenko2012network}.
-However, we argue that the network pertubation theory used in most flavors of NEA is a red herring.
+We argue that the network pertubation theory used in most flavors of NEA is a red herring.
+Users of NEA do not differ from users of ORA they want to test an association between a query set of genes and a pathways.
 The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in repect to a pathway.
-In this question, the network is not relevant.
-A network may help to increase the sensitivity of ORA methods, but it is better left out of the null formation of the null hypothesis.
+In this question, the introduction of a  network is not relevant.
+A network may help to increase the sensitivity of ORA methods, but it is better left out of the null hypothesis.
 We argue that such test is naturally implemented as tests of the query set, i.e.
@@
 
+In effect, NEA is an attempt to increase the sensitivity of ORA.
+The standard user of NEA would typically have identified a set of query genes which she wants to evaluate for over representation in respect to a pathway database.
+This goal is the same as for users of ORA.
+Hence the null model should be of the same nature for NEA and ORA.
+GeneSetDP and GeneSetMC are two algorithms that proves that one in NEA can test the uniqness of the querry set in respect to a pathway, just as done in ORA.
+
 \section*{Acknowledgments}

### 2018-10-12 doc/response.tex
Edited the response letter and made a diff.pdf

@@
 We have attached a revised version of our manuscript ``A simple null model for inferences from network enrichment analysis'', and a diff
-against our original version. We have also answered the reviewers comments
+against our original manuscript. We have also answered the reviewers comments
 below.
@@
 \ereview
-The reviewer points out that the manuscript only discusses the calibration of $p$ values, and that actually never uses any positive tests.
-\ldots
-It is also worth noting that according to the Criteria For Publication PLOS ONE, (\url{https://journals.plos.org/plosone/s/criteria-for-publication}) conceved meaningfulness is not a publication criteria.
-
-\medskip
+The reviewer points out that the manuscript only discusses the calibration of $p$ values, and never actually demonstartes the utility of such $p$ values.
+The reviewer even does not find the findings of the manuscript meaningful.
+While it is worth noting that  according to the Criteria For Publication PLOS ONE, (\url{https://journals.plos.org/plosone/s/criteria-for-publication}) conceved meaningfulness is not a publication criteria for a manuscript, lack of meaning is obviously an undersirable property of any scientific report.
+We have carefully rewritten the introduction to make the readers to get a sence of the underlying conflict driving our manuscript.
+We feal ourselves limited from writing that excisting methods for network enrichment analysis (NEA) produce grosely inflated statistics.
+We instead have tried to, in a constructive manner, described a method for deriving statistics that is natural from a users perspective.
+This model could be integrated into any of the excisting models for NEA.
+It is an open question if one could demonstrate improved sensitivity with NEA over over-representation analysis in such implementations.
 
@@
 I find this disingenuous; they have to at least show how their ranked results compare to BinoX on real data and do some kind of benchmarking.
-This is important because I believe there is a difference in the null hypothesis stated in the current work and that used by BinoX that makes the comparison presented here unfair – they are not testing the same thing.
+This is important because I believe there is a difference in the null hypothesis stated in the current work and that used by BinoX that makes the comparison presented here unfair -- they are not testing the same thing.
+\ereview
 
-\medskip
+We fully agree with the reviewer that the null of BinoX is different that GeneSetDP.
+In fact that is the purpose of our manuscript.
+The reason why no sensitivity test is made is ofcorse that there is no way to compete in sensitivity against a method with inflated statistics.
+If the yardstick is flawed no comparison can be made.
+Here we establish a fair yardstick to compare against.
+
+\breview
 The crux is that the method proposed here focuses on the properties of the query genes; randomly sampling query genes makes the proposed method a test of whether the query genes are unusual, not necessarily specifically with respect to a particular pathway.
 For example, if the query contains a hub gene (assuming hubs are rare) then it might be found “significantly” connected to many pathways.
-The “feature unlikely to occur by chance” is that the query contains a hub gene, not that it connects to many pathways.
+The ``feature unlikely to occur by chance'' is that the query contains a hub gene, not that it connects to many pathways.
 To state it in an extreme way, in this situation biologist should skip the pathway analysis and just take note of the fact that the query includes a hub.
@@
 \medskip
+
 As I understand it, BinoX tries to account for these difficulties by modeling realizations of the network that retain key topological properties like node degree distribution while holding the query constant (in effect).
@@
 Without that, and without any benchmarking, the paper is essentially about an algorithm to compute N(s), with no biological content or relevance.
+
 \ereview
+
+Here, we respectfully disagree with the reviewer.
+The reviewer's perspective do not reflect the actual use case.
+The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in repect to a pathway database.
+In this question, the introduction of a  network is not relevant more than in that it might shift the score distribution against pathways.
+A network may or may not help to increase the sensitivity of ORA methods, but it is better left out of the null hypothesis.
+A user should not have to worry about the conectivity of individual genes.
+Just as the reviewer points out, there are certain hub genes that are connected to more pathways than others through any network.
+This will be reflected by the score distribution produced by our described methods.
+If the query set contains such a gene, the query is more likely to be significant against a pathway, however that is the desired property of the methods.
+Conversly a querry containing one or more genes not connected to any pathways is less likely to be significantly enriched.
+This is the nature of this type of analysis.
+
+Perhaps more problematic, the methods in this manuscript, just in all NEA methods, test every pathway independently for significance, and do not evaluate correlation between different pathways.
+\breview
 
@@
 \ereview
+We updated these typos, and we as well tried to improve the language of the manuscript.
 

### 2018-10-14 doc/genesetdp.tex
Spell check

@@
 % Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
-The methods that are used for network randomization, are often mentally untractable as they include a big number of steps that are subject to complex constrains.
-As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow contrains that depends on previous link assignments.
-While this results in a network that is both randomized and contain some preserved desired topological features, these nested constrains make it very hard to verify that the resulting network is representative.
+The methods that are used for network randomization, are often mentally untractable as they include a big number of steps that are subject to complex constraints.
+As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step, they must follow constraints that depend on previous link assignments.
+While this results in a network that is both randomized and contains some desired topological features preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative.
 In theory there is a space of all possible networks that could be made up from a certain node degree distribution, however, it is not clear that the second-order conservation method samples evenly from that distribution.
-And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose the statistical test.
+And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose of the statistical test.
 
-We argue that the network pertubation theory used in most flavors of NEA is a red herring.
-Users of NEA do not differ from users of ORA they want to test an association between a query set of genes and a pathways.
-The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in repect to a pathway.
+We argue that the network perturbation theory used in most flavors of NEA is a red herring.
+Users of NEA do not differ from users of ORA they want to test an association between a query set of genes and pathways.
+The typical user has used an experimental or computational method to derive a set of query genes that the user wants to know if it is random, or if it is enriched with respect to a pathway.
 In this question, the introduction of a  network is not relevant.
-A network may help to increase the sensitivity of ORA methods, but it is better left out of the null hypothesis.
-We argue that such test is naturally implemented as tests of the query set, i.e.
-is the query set a randomly selected set of genes or is there an over representation of well-connected genes in the query set?
+A network may help to increase the sensitivity of ORA methods, but for practical reasons it is better left out of the null hypothesis.
+We argue that tests of significance in this type of analysis is naturally implemented as tests of the query set.
+That is, is the query set a randomly selected set of genes or is there an over-representation of well-connected genes in the query set?
 
 Hence, here we employ a different null model, more directly aimed at modeling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''.
-We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size, noting here that we do not need to make any assumptions on the nature of the score distribuition.
+We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
-Combining the GeneSetDP and  GeneSetMC with the scoring system of the popular method BinoX, we using simulations to demonstrate that both our algorithms produced unbiased statistics.
+Both algorithms circumvent the need of parametric score distributions for the calculatios of significance.
+Combining the GeneSetDP and  GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.
 
@@
 \end{equation}
-Such number of links, $l_i$, can be pre-computed for all genes in $\mathcal{G}$.
-For later purposes we also define a mapping, $k_a=\sum_{\{i:l_i=a\}}1$, giving the number of genes having $a$ links to the investigated pathway.
+Such a number of links, $l_i$, can be pre-computed for all genes in $\mathcal{G}$.
+For later purposes we also define a mapping, $k_a=\sum_{\{i:l_i=a\}}1$, giving the number of genes having, $a$, links to the investigated pathway.
 
@@
 
-We made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``There are not more links between the query and pathway gene sets than expected by chance''.
+We made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$: ``There are not more links between the query and pathway gene sets than expected by chance''.
 Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.
@@
 In effect, NEA is an attempt to increase the sensitivity of ORA.
-The standard user of NEA would typically have identified a set of query genes which she wants to evaluate for over representation in respect to a pathway database.
+The standard user of NEA would typically have identified a set of query genes which she wants to evaluate for over-representation in respect to a pathway database.
 This goal is the same as for users of ORA.
 Hence the null model should be of the same nature for NEA and ORA.
-GeneSetDP and GeneSetMC are two algorithms that proves that one in NEA can test the uniqness of the querry set in respect to a pathway, just as done in ORA.
+GeneSetDP and GeneSetMC are two algorithms that prove that one in NEA can test the uniqueness of the query set in respect to a pathway, just as done in ORA.
 

### 2018-10-14 doc/response.tex
More spell check

@@
 We have attached a revised version of our manuscript ``A simple null model for inferences from network enrichment analysis'', and a diff
-against our original manuscript. We have also answered the reviewers comments
+against our original manuscript. We have also answered the reviewer's comments
 below.
+
+In general, by reading the reviewers comments, we feel that we failed to deliver the purpose of the manuscript to the reviewer, who indeed appears to have misunderstood what we are aiming at with our manuscript.
+We have hence added clarifications to the introduction and the discussion.
+We hope that this will make the story even more interesting.
 
@@
 \ereview
-The reviewer points out that the manuscript only discusses the calibration of $p$ values, and never actually demonstartes the utility of such $p$ values.
+The reviewer points out that the manuscript only discusses the calibration of $p$ values, and never actually demonstrates the utility of such $p$ values.
 The reviewer even does not find the findings of the manuscript meaningful.
-While it is worth noting that  according to the Criteria For Publication PLOS ONE, (\url{https://journals.plos.org/plosone/s/criteria-for-publication}) conceved meaningfulness is not a publication criteria for a manuscript, lack of meaning is obviously an undersirable property of any scientific report.
-We have carefully rewritten the introduction to make the readers to get a sence of the underlying conflict driving our manuscript.
-We feal ourselves limited from writing that excisting methods for network enrichment analysis (NEA) produce grosely inflated statistics.
+While it is worth noting that  according to the Criteria For Publication PLOS ONE, (\url{https://journals.plos.org/plosone/s/criteria-for-publication}) perceived meaningfulness is not a publication criterion for a manuscript, lack of meaning is obviously an undesirable property of any scientific report.
+We have carefully rewritten the introduction to make the readers get a sense of the underlying conflict driving our manuscript.
+We feel ourselves limited from writing that existing methods for network enrichment analysis (NEA) produce grossly inflated statistics.
 We instead have tried to, in a constructive manner, described a method for deriving statistics that is natural from a users perspective.
-This model could be integrated into any of the excisting models for NEA.
+This model could be integrated into any of the currently used  models for NEA.
 It is an open question if one could demonstrate improved sensitivity with NEA over over-representation analysis in such implementations.
@@
 If the results yielded by the proposed method are basically the same as BinoX (in terms of ranks), then there is no major problem.
-But the authors state that they don’t want to get into evaluating “performance in terms of sensitivity”.
+But the authors state that they don’t want to get into evaluating ``performance in terms of sensitivity''.
 I find this disingenuous; they have to at least show how their ranked results compare to BinoX on real data and do some kind of benchmarking.
@@
 In fact that is the purpose of our manuscript.
-The reason why no sensitivity test is made is ofcorse that there is no way to compete in sensitivity against a method with inflated statistics.
+The reason why no sensitivity test is made is of course that there is no way to compete in sensitivity against a method with inflated statistics.
 If the yardstick is flawed no comparison can be made.
 Here we establish a fair yardstick to compare against.
+In our manuscript we are not trying to sell a new method for NEA, we are merely trying to convince the field that they should report statistics that makes sense to a user.
 
@@
 The crux is that the method proposed here focuses on the properties of the query genes; randomly sampling query genes makes the proposed method a test of whether the query genes are unusual, not necessarily specifically with respect to a particular pathway.
-For example, if the query contains a hub gene (assuming hubs are rare) then it might be found “significantly” connected to many pathways.
+For example, if the query contains a hub gene (assuming hubs are rare) then it might be found ``significantly'' connected to many pathways.
 The ``feature unlikely to occur by chance'' is that the query contains a hub gene, not that it connects to many pathways.
@@
 The reviewer's perspective do not reflect the actual use case.
-The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in repect to a pathway database.
+The typical user have used an experimental or computational method to derive a set of query genes that the user want to know if it is random, or if it is enriched in respect to a pathway database.
 In this question, the introduction of a  network is not relevant more than in that it might shift the score distribution against pathways.
 A network may or may not help to increase the sensitivity of ORA methods, but it is better left out of the null hypothesis.
-A user should not have to worry about the conectivity of individual genes.
+A user should not have to worry about the connectivity of individual genes.
 Just as the reviewer points out, there are certain hub genes that are connected to more pathways than others through any network.
@@
 If the query set contains such a gene, the query is more likely to be significant against a pathway, however that is the desired property of the methods.
-Conversly a querry containing one or more genes not connected to any pathways is less likely to be significantly enriched.
+Conversely, a query containing one or more genes not connected to any pathways is less likely to be significantly enriched.
 This is the nature of this type of analysis.
 
-Perhaps more problematic, the methods in this manuscript, just in all NEA methods, test every pathway independently for significance, and do not evaluate correlation between different pathways.
+However, a perhaps more problematic fact is that the methods in this manuscript, just in all NEA methods, test every pathway independently for significance, and do not evaluate correlation between different pathways.
 \breview
@@
 \ereview
-We updated these typos, and we as well tried to improve the language of the manuscript.
+We updated these typos, and we also overall improved the language of the manuscript.
 

### 2018-10-15 doc/genesetdp.tex
More spell check
Simplified preamble
grammar check

@@
 
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
 \usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
 \usepackage{changepage}
-
-% Use Unicode characters when possible
 \usepackage[utf8x]{inputenc}
-
-% textcomp package and marvosym package for additional characters
 \usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
+\usepackage{cite,nameref,hyperref}
 \usepackage{microtype}
@@
 
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
+\usepackage[right]{lineno}
 \usepackage{array}
-
-% create "+" rule type for thick vertical lines
 \newcolumntype{+}{!{\vrule width 2pt}}
@@
 While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases.
-Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}.
+Hence, efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}.
 Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}.
@@
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
-Both algorithms circumvent the need of parametric score distributions for the calculatios of significance.
+Both algorithms circumvent the need of parametric score distributions for the calculations of significance.
 Combining the GeneSetDP and  GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.
@@
 Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.
-In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics.
-In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.
+In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use for calculating accurate statistics.
+Generally, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.
 
@@
 This goal is the same as for users of ORA.
-Hence the null model should be of the same nature for NEA and ORA.
+Hence, the null model should be of the same nature for NEA and ORA.
 GeneSetDP and GeneSetMC are two algorithms that prove that one in NEA can test the uniqueness of the query set in respect to a pathway, just as done in ORA.

### 2018-10-15 doc/genesetdp.tex
edits

@@
 % Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
-The methods that are used for network randomization, are often mentally untractable as they include a big number of steps that are subject to complex constraints.
+The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
 As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step, they must follow constraints that depend on previous link assignments.
 While this results in a network that is both randomized and contains some desired topological features preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative.
-In theory there is a space of all possible networks that could be made up from a certain node degree distribution, however, it is not clear that the second-order conservation method samples evenly from that distribution.
+In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.
 And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose of the statistical test.
 
-We argue that the network perturbation theory used in most flavors of NEA is a red herring.
-Users of NEA do not differ from users of ORA they want to test an association between a query set of genes and pathways.
-The typical user has used an experimental or computational method to derive a set of query genes that the user wants to know if it is random, or if it is enriched with respect to a pathway.
+We argue that the network perturbation methods used in most flavors of NEA is a red herring.
+Users of NEA do not differ from users of ORA and want to test an association between a query set of genes and pathways.
+The typical user has used an experimental or computational method to derive a set of query genes and wants to know if it is a random set, or if it is enriched with respect to a pathway.
 In this question, the introduction of a network is not relevant.
@@
 Hence, here we employ a different null model, more directly aimed at modeling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''.
-We present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
+We present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
@@
 To test this we plotted the $p$~values against their quantile in Fig \ref{fig:calibration} for $10000$ randomly assembled queries from the human genome.
-As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.
+As a comparison, we also added a calibration curve, under our null hypothesis, for the popular NEA method BinoX\cite{ogris2016novel}.
 Here, for all three methods, we are only testing the queries for enrichment, that is, we are only concerned with a higher number of links than expected by random, and this differs from the default BinoX setting which tests for both enrichment and depletion.
-Also of note is the fact that, while we display all methods in the same figure, each represents its own calibration, and that the Spearman's correlation between BinoX and the other methods is fairly low ($\rho = 0.49$, $0.55$ and $0.55$ for $Q=10$, $20$ and $30$ respectively)
+Also of note is the fact that, while we display all methods in the same figure, each represents its own calibration, and that the Spearman's correlation between BinoX and our methods is fairly low ($\rho = 0.49$, $0.55$ and $0.55$ for $Q=10$, $20$ and $30$ respectively)
 
@@
 the calculated $p$ values are larger than expected.
-Meanwhile, BinoX appears strongly anti-conservative, that is, the $p$ values are lower than expected.
+Meanwhile, BinoX appears strongly anti-conservative, that is, the $p$ values are lower than expected, even when not taking into account the significance by depletion.
 
@@
 We argue that more studies should make a point at demonstrating that their methods are well-calibrated, as it is a prerequisite for measuring performance at least when the ground truth of biological effect is not known.
-With this in mind, we demonstrated that both our methods gave a better calibration than the popular BinoX method, by random sampling of subsets of a genome.
+% With this in mind, we demonstrated that both our methods gave a better calibration than the popular BinoX method, by random sampling of subsets of a genome.
 
@@
 We made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$: ``There are not more links between the query and pathway gene sets than expected by chance''.
-Previous implementations of NEA in the literature use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network.
+Previous implementations of NEA in the literature use network randomization methods to determine parameters for various types parametric distributions, by perturbations of the investigated network.
 In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use for calculating accurate statistics.
 Generally, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.
+In this line, network randomization methods would be better suited for studying the relevance of the links in the network than the query set itself.
 

### 2018-10-15 doc/response.tex
edits
Edited together.

@@
 \breview
-Jeunken and Käll describe a method for generating a null distribution for so called ``network enrichment analyses'', which are sometimes framed as ``crosstalk'' analyses.
+Jeunken and K\"{a}ll describe a method for generating a null distribution for so called ``network enrichment analyses'', which are sometimes framed as ``crosstalk'' analyses.
 The only data presented is showing that the method generates accurate pvalues under the null.
+
 \medskip
+
 As far as I can tell the work is technically correct.
@@
 We have carefully rewritten the introduction to make the readers get a sense of the underlying conflict driving our manuscript.
-We feel ourselves limited from writing that existing methods for network enrichment analysis (NEA) produce grossly inflated statistics.
+However, we tried to not to bee too direct when writing that existing methods for network enrichment analysis (NEA) produce grossly inflated statistics.
 We instead have tried to, in a constructive manner, described a method for deriving statistics that is natural from a users perspective.
@@
 In fact that is the purpose of our manuscript.
-The reason why no sensitivity test is made is of course that there is no way to compete in sensitivity against a method with inflated statistics.
+One reason why no sensitivity test is made is that there is no way to compete in sensitivity against a method with inflated statistics.
 If the yardstick is flawed no comparison can be made.
@@
 In our manuscript we are not trying to sell a new method for NEA, we are merely trying to convince the field that they should report statistics that makes sense to a user.
+Furthermore, there is nothing to benchmark but the statistics itself. We use the same scoring system as BinoX, i.e. number of links.
 
@@
 The current authors criticize the BinoX approach, but ignore the problem entirely.
-A closer (but not very close) approximation of the BinoX approach might be, for the Monte Carlo method proposed here, to bias the sampling of Q’ from G such that the distribution of the node degrees of the genes in Q’ resembles the distribution of the actual query set of interest Q.
+A closer (but not very close) approximation of the BinoX approach might be, for the Monte Carlo method proposed here, to bias the sampling of $Q’$ from $G$ such that the distribution of the node degrees of the genes in $Q’$ resembles the distribution of the actual query set of interest $Q$.
 In any case, I believe that the BinoX approach is not directly comparable.
@@
 If the authors cannot do this then they must remove the discussion of BinoX.
-Without that, and without any benchmarking, the paper is essentially about an algorithm to compute N(s), with no biological content or relevance.
+Without that, and without any benchmarking, the paper is essentially about an algorithm to compute $N(s)$, with no biological content or relevance.
 
@@
 This will be reflected by the score distribution produced by our described methods.
-If the query set contains such a gene, the query is more likely to be significant against a pathway, however that is the desired property of the methods.
-Conversely, a query containing one or more genes not connected to any pathways is less likely to be significantly enriched.
+If the query set contains such a gene, the query is more likely to be significant against a pathway, however, that is the desired property of the methods.
+Conversely, any query that contain one or more genes that are not connected to any pathways is less likely to be significantly enriched.
 This is the nature of this type of analysis.
@@
 However, a perhaps more problematic fact is that the methods in this manuscript, just in all NEA methods, test every pathway independently for significance, and do not evaluate correlation between different pathways.
-\breview
 

### 2018-10-16 doc/genesetdp.tex
Edited together.
small text changes

@@
 While the terminology varies in the literature, we will here refer to such methods as Network Enrichment Analysis (NEA).
-Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations.
-Here we instead have designed one dynamic programming algorithm that calculates the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences.
+Traditionally, the significance of inferences from NEA has been assigned using a null model that derives from a randomization of the network.
+Here we instead argue for a null model that more direcly relates to the set of genes being studied, and have designed one dynamic programming algorithm that calculates the score distribution of NEA scores that makes it possible to assign unbiased mid $p$~values to inferences.
 We also implemented a random sampling method, carrying out the same task.
@@
 The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
-As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step, they must follow constraints that depend on previous link assignments.
-While this results in a network that is both randomized and contains some desired topological features preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative.
+As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
+While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative.
 In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.
@@
 the calculated $p$ values are larger than expected.
-Meanwhile, BinoX appears strongly anti-conservative, that is, the $p$ values are lower than expected, even when not taking into account the significance by depletion.
+Meanwhile, BinoX appears strongly anti-conservative, that is, the $p$ values are lower than expected, even when not taking into account significance by depletion.
 
@@
 Instead of testing our methods' performance in terms of sensitivity, we here instead chose to test the method in terms of the methods statistical accuracy.
-We argue that more studies should make a point at demonstrating that their methods are well-calibrated, as it is a prerequisite for measuring performance at least when the ground truth of biological effect is not known.
+We argue that more studies should make a point at demonstrating that their methods are well-calibrated, as it is a prerequisite for measuring performance, most importantly when the ground truth of biological effect is not known.
 % With this in mind, we demonstrated that both our methods gave a better calibration than the popular BinoX method, by random sampling of subsets of a genome.
@@
 Previous implementations of NEA in the literature use network randomization methods to determine parameters for various types parametric distributions, by perturbations of the investigated network.
-In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use for calculating accurate statistics.
+In practice, such perturbations are very heuristic in their nature, and hence are difficult to use for calculating accurate statistics.
 Generally, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.
-In this line, network randomization methods would be better suited for studying the relevance of the links in the network than the query set itself.
 
+Furthermore, network randomization methods focus on significance of the links in the network instead of the query set itself.
 In effect, NEA is an attempt to increase the sensitivity of ORA.

### 2018-10-16 doc/genesetdp.tex
editing together.

@@
 While the terminology varies in the literature, we will here refer to such methods as Network Enrichment Analysis (NEA).
-Traditionally, the significance of inferences from NEA has been assigned using a null model that derives from a randomization of the network.
+Traditionally, the significance of inferences from NEA has been assigned using a null model constructed from randomizations of the network.
 Here we instead argue for a null model that more direcly relates to the set of genes being studied, and have designed one dynamic programming algorithm that calculates the score distribution of NEA scores that makes it possible to assign unbiased mid $p$~values to inferences.
@@
 Here, for all three methods, we are only testing the queries for enrichment, that is, we are only concerned with a higher number of links than expected by random, and this differs from the default BinoX setting which tests for both enrichment and depletion.
-Also of note is the fact that, while we display all methods in the same figure, each represents its own calibration, and that the Spearman's correlation between BinoX and our methods is fairly low ($\rho = 0.49$, $0.55$ and $0.55$ for $Q=10$, $20$ and $30$ respectively)
+% Also of note is the fact that, while we display all methods in the same figure, each represents its own calibration, and that the Spearman's correlation between BinoX and our methods is fairly low ($\rho = 0.49$, $0.55$ and $0.55$ for $Q=10$, $20$ and $30$ respectively)
 
@@
 In practice, such perturbations are very heuristic in their nature, and hence are difficult to use for calculating accurate statistics.
-Generally, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.
+Generally, it is easier to understand null models that relate to user actions.
 

### 2018-10-16 doc/genesetdp.tex
spellcheck

@@
 Traditionally, the significance of inferences from NEA has been assigned using a null model constructed from randomizations of the network.
-Here we instead argue for a null model that more direcly relates to the set of genes being studied, and have designed one dynamic programming algorithm that calculates the score distribution of NEA scores that makes it possible to assign unbiased mid $p$~values to inferences.
+Here we instead argue for a null model that more directly relates to the set of genes being studied, and have designed one dynamic programming algorithm that calculates the score distribution of NEA scores that makes it possible to assign unbiased mid $p$~values to inferences.
 We also implemented a random sampling method, carrying out the same task.
@@
 A limiting factor of ORA is the quality of the databases defining pathways.
-While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide}, some critique has been voiced concerning the completeness and rigor of current pathway databases.
+While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide}, some critique has been voiced concerning the completeness and rigour of current pathway databases.
 Hence, efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}.
@@
 
-We argue that the network perturbation methods used in most flavors of NEA is a red herring.
+We argue that the network perturbation methods used in most flavours of NEA is a red herring.
 Users of NEA do not differ from users of ORA and want to test an association between a query set of genes and pathways.
@@
 
-Hence, here we employ a different null model, more directly aimed at modeling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''.
+Hence, here we employ a different null model, more directly aimed at modelling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''.
 We present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.

### 2018-10-25 doc/genesetdp.tex
items 5 and 6 of checklist
commented out figures

@@
 \\
-Gustavo S. Jeuken\textsuperscript{1} and
+Gustavo Jeuken\textsuperscript{1} and
 Lukas K\"{a}ll\textsuperscript{1,*}
@@
 	\begin{center}
-		\includegraphics[width=.9\textwidth]{figures/score_distribuition_multiple.png}
+		% \includegraphics[width=.9\textwidth]{figures/score_distribuition_multiple.png}
     \end{center}
@@
 		  \begin{tabular}[t]{c}
-				\includegraphics[width=.9\textwidth]{figures/calibration_multiple.png} \\
-				\includegraphics[width=0.45\textwidth]{figures/calibration_legend.png}
+				% \includegraphics[width=.9\textwidth]{figures/calibration.png}
 		\end{tabular}
@@
 To test the consistency of the methods, we also investigated the correspondence between the $p$~values reported by GeneSetDP and GeneSetMC.
-As can be seen in Figure~\ref{fig:pscatter}, the results are consistent as expected.
+As can be seen in Fig~\ref{fig:pscatter}, the results are consistent as expected.
 
@@
         % (Idea from reviewer comment in email 2018-06-08.)
-				\includegraphics[width=.9\textwidth]{figures/scatter_dp_mc.png}
+				% \includegraphics[width=.9\textwidth]{figures/scatter_dp_mc.png}
 		\end{tabular}

### 2018-10-25 doc/genesetdp.tex
Style changes.

@@
 Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}.
-While the terminology varies in the literature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).
+We will refer to such methods as {\em Network Enrichment Analysis} (NEA), although the concept also is known as crosstalk analysis.
 
@@
 Previous efforts have settled for network randomization, where the network definitions are randomly constructed by methods with varying degrees of topological conservation in respect to the original network.
+The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
 
-% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics.
-For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
-% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
-The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
 As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
 While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative.
+
 In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.
@@
 
-We argue that the network perturbation methods used in most flavours of NEA is a red herring.
+We argue that the network perturbation methods used in most flavours of NEA are missleading.
 Users of NEA do not differ from users of ORA and want to test an association between a query set of genes and pathways.
@@
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
-Both algorithms circumvent the need of parametric score distributions for the calculations of significance.
+Both algorithms circumvent the need of network pertubations for the calculations of significance.
 Combining the GeneSetDP and GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.

### 2018-10-26 doc/genesetdp.tex
minor grammatic corrections, bibliography in the tex file

--- 2018-10-25 doc/genesetdp.tex
@@
 % Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
-The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
+Yet, the methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
 As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
@@
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
-Both algorithms circumvent the need of parametric score distributions for the calculations of significance.
+Both algorithms circumvent the need for parametric score distributions for the calculations of significance.
 Combining the GeneSetDP and GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.
@@
 
-\bibliography{./refs.bib}
+\begin{thebibliography}{10}
+
+\bibitem{tavazoie1999systematic}
+Tavazoie S, Hughes JD, Campbell MJ, Cho RJ, Church GM.
+\newblock Systematic determination of genetic network architecture.
+\newblock Nature genetics. 1999;22(3):281.
+
+\bibitem{khatri2012ten}
+Khatri P, Sirota M, Butte AJ.
+\newblock Ten years of pathway analysis: current approaches and outstanding
+  challenges.
+\newblock PLoS computational biology. 2012;8(2):e1002375.
+
+\bibitem{goeman2007analyzing}
+Goeman JJ, B\"{u}hlmann P.
+\newblock Analyzing gene expression data in terms of gene sets: methodological
+  issues.
+\newblock Bioinformatics. 2007;23(8):980--987.
+
+\bibitem{subramanian2005gene}
+Subramanian A, Tamayo P, Mootha VK, Mukherjee S, Ebert BL, Gillette MA, et~al.
+\newblock Gene set enrichment analysis: a knowledge-based approach for
+  interpreting genome-wide expression profiles.
+\newblock Proceedings of the National Academy of Sciences.
+  2005;102(43):15545--15550.
+
+\bibitem{bader2006pathguide}
+Bader GD, Cary MP, Sander C.
+\newblock Pathguide: a pathway resource list.
+\newblock Nucleic acids research. 2006;34(suppl\_1):D504--D506.
+
+\bibitem{szklarczyk2014string}
+Szklarczyk D, Franceschini A, Wyder S, Forslund K, Heller D, Huerta-Cepas J,
+  et~al.
+\newblock STRING v10: protein--protein interaction networks, integrated over
+  the tree of life.
+\newblock Nucleic acids research. 2014;43(D1):D447--D452.
+
+\bibitem{ogris2017funcoup}
+Ogris C, Guala D, Kaduk M, Sonnhammer EL.
+\newblock FunCoup 4: new species, data, and visualization.
+\newblock Nucleic acids research. 2017;46(D1):D601--D607.
+
+\bibitem{alexeyenko2012network}
+Alexeyenko A, Lee W, Pernemalm M, Guegan J, Dessen P, Lazar V, et~al.
+\newblock Network enrichment analysis: extension of gene-set enrichment
+  analysis to gene networks.
+\newblock {BMC} bioinformatics. 2012;13(1):226.
+
+\bibitem{glaab2012enrichnet}
+Glaab E, Baudot A, Krasnogor N, Schneider R, Valencia A.
+\newblock EnrichNet: network-based gene set enrichment analysis.
+\newblock Bioinformatics. 2012;28(18):i451--i457.
+
+\bibitem{mccormack2013statistical}
+McCormack T, Frings O, Alexeyenko A, Sonnhammer EL.
+\newblock Statistical assessment of crosstalk enrichment between gene groups in
+  biological networks.
+\newblock {PloS} one. 2013;8(1):e54945.
+
+\bibitem{ogris2016novel}
+Ogris C, Guala D, Helleday T, Sonnhammer EL.
+\newblock A novel method for crosstalk analysis of biological networks:
+  improving accuracy of pathway annotation.
+\newblock Nucleic acids research. 2016;45(2):e8--e8.
+
+\bibitem{signorelli2016neat}
+Signorelli M, Vinciotti V, Wit EC.
+\newblock NEAT: an efficient network enrichment analysis test.
+\newblock BMC bioinformatics. 2016;17(1):352.
+
+\bibitem{lancaster1961significance}
+Lancaster HO.
+\newblock Significance tests in discrete distributions.
+\newblock Journal of the American Statistical Association.
+  1961;56(294):223--234.
+
+\bibitem{hwang2001optimality}
+Hwang JG, Yang MC.
+\newblock An optimality theory for mid p--values In 2$\times$ 2 contingency
+  tables.
+\newblock Statistica Sinica. 2001; p. 807--826.
+
+\end{thebibliography}
 

### 2018-10-26 doc/genesetdp.tex

@@
 
-% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics.
-For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
-% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
 Yet, the methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
+
 As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
 While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative.
+
 In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.
@@
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
-Both algorithms circumvent the need for parametric score distributions for the calculations of significance.
+Both algorithms circumvent the need of network pertubations for the calculations of significance.
 Combining the GeneSetDP and GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.

### 2018-10-26 doc/genesetdp.tex
Back to old name.

@@
 \\
-Gustavo Jeuken\textsuperscript{1} and
+Gustavo S. Jeuken\textsuperscript{1} and
 Lukas K\"{a}ll\textsuperscript{1,*}
@@
 One prominent application of the technique is expression analysis, where ORA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.
-
 Most ORA methods are assessing the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test.
@@
 Previous efforts have settled for network randomization, where the network definitions are randomly constructed by methods with varying degrees of topological conservation in respect to the original network.
-
 Yet, the methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
-
 As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
@@
 We argue that the network perturbation methods used in most flavours of NEA are missleading.
-Users of NEA do not differ from users of ORA and want to test an association between a query set of genes and pathways.
+Users of NEA do not differ from users of ORA, they want to test an association between a query set of genes and pathways.
 The typical user has used an experimental or computational method to derive a set of query genes and wants to know if it is a random set, or if it is enriched with respect to a pathway.
@@
 A network may help to increase the sensitivity of ORA methods, but for practical reasons it is better left out of the null hypothesis.
-We argue that tests of significance in this type of analysis is naturally implemented as tests of the query set.
+Tests of significance in this type of analysis is naturally implemented as tests of the query set.
 That is, is the query set a randomly selected set of genes or is there an over-representation of well-connected genes in the query set?
@@
 Hence, here we employ a different null model, more directly aimed at modelling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''.
-We present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
+Here, we present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size.
 We also implemented one random sampling algorithm for the same task, that we called GeneSetMC.
